---
title: "CDISC Pilot 01 — SDTM to ADaM in SAS and R"
subtitle: "Double-programming an FDA-grade analysis package on a publicly redistributable clinical-trial dataset"
author: "Paulina Del Mundo Del Fierro, MD, MPH"
date: last-modified
format:
  html:
    theme: [cosmo, ../_theme.scss]
    highlight-style: github-dark
    toc: true
    toc-depth: 3
    toc-location: right
    code-tools: true
    code-copy: true
    code-fold: true
    code-summary: "Show code"
    code-overflow: wrap
    number-sections: true
    fig-responsive: true
    fig-width: 8
    fig-height: 5
    mainfont: "Inter, system-ui, sans-serif"
    monofont: "JetBrains Mono, Menlo, monospace"
    embed-resources: true
    include-in-header:
      text: |
        <script>
          window.va = window.va || function () { (window.vaq = window.vaq || []).push(arguments); };
        </script>
        <script defer src="/_vercel/insights/script.js"></script>
execute:
  echo: true
  eval: false
  warning: false
  message: false
  freeze: auto
---

::: {.callout-tip icon=false}
## TL;DR — what this notebook delivers

- **Two parallel ADaM programs.** One in **SAS** (the FDA submission lingua franca, run on SAS OnDemand for Academics), one in **R** using `{admiral}` (the open-source pharmaverse package that Roche, GSK, J&J, and Pfizer co-maintain). Both read the same SDTM XPTs and produce the same ADaM XPTs.
- **Three derived analysis datasets.** ADSL (one record per subject), ADAE (occurrence-level adverse events with TRTEMFL), ADLBC (BDS-structured laboratory chemistry with baseline and change-from-baseline).
- **One Table 14-2.01.** ICH E3 demographics, by treatment arm, rendered from each ADSL independently and reconciled.
- **Pinnacle 21 conformance.** `{xportr}` writes the XPTs with v5 transport metadata; the open conformance rules pass without warning.
- **Byte-for-byte parity check.** The closing chunk diffs the SAS-derived and R-derived ADSLs against CDISC's published reference ADSL. Any mismatch fails the render.
:::

::: {.callout-warning icon=false}
## Status — scaffold only (in progress)

This page is the published scaffold. Code chunks are shown for review but **do not execute on render** until the SDTM XPTs are downloaded (`make data`) and the `{admiral}` / `{xportr}` / `{metacore}` packages are added to `renv.lock`. The SAS programs in `sas/` are complete and runnable on SAS OnDemand for Academics; rendered logs and listings will be linked in once the run completes.
:::

## Why this analysis

Every interventional clinical trial submitted to the FDA, EMA, or PMDA arrives as a CDISC package: collected data conforms to **SDTM** (Study Data Tabulation Model, IG v3.4), and analysis-ready data conforms to **ADaM** (Analysis Data Model, IG v1.3). Submission programmers typically write the same derivations twice — primary in SAS, secondary in either SAS or R — and the two outputs must reconcile before database lock.

This notebook is that double-programming exercise on a public dataset:

1. **Read** the CDISC Pilot 01 SDTM domains (DM, AE, EX, LB, SV, VS) directly from SAS V5 transport (XPT) files.
2. **Derive** ADSL, ADAE, and ADLBC using ADaM IG v1.3 conventions — twice, once in SAS, once in R/`{admiral}`.
3. **Reconcile** the two derivations against each other and against CDISC's published reference ADaM.
4. **Report** an ICH E3 §14.2.1 demographics table and the reconciliation result.

CDISC Pilot 01 is a fictional 254-subject Alzheimer's trial (placebo / donepezil 5 mg / donepezil 10 mg) released by CDISC under a permissive license expressly so the standards can be taught and audited in public — see <https://github.com/cdisc-org/sdtm-adam-pilot-project>.

::: {.callout-note}
The `.qmd` source for this page, the SAS programs, and the YAML define-XML extract live on [GitHub](https://github.com/paulinedmdf/paulinadelmundomd/tree/main/projects/03-cdisc-adam-pilot).
:::

---

## Setup

```{r setup-r}
#| label: setup-r
suppressPackageStartupMessages({
  library(admiral)      # ADaM derivations
  library(xportr)       # writes v5 transport, validates against CDISC CT
  library(haven)        # reads/writes XPT
  library(metacore)     # reads define-xml -> R object
  library(metatools)    # applies metacore to a dataset
  library(dplyr)
  library(tidyr)
  library(gtsummary)
  library(here)
})

DATA_SDTM <- here::here("projects", "03-cdisc-adam-pilot", "data", "sdtm")
DATA_REF  <- here::here("projects", "03-cdisc-adam-pilot", "data", "adam_reference")
DATA_OUT  <- here::here("projects", "03-cdisc-adam-pilot", "data", "adam_r")
SAS_OUT   <- here::here("projects", "03-cdisc-adam-pilot", "sas", "output")
dir.create(DATA_OUT, showWarnings = FALSE, recursive = TRUE)
```

## Read SDTM

```{r read-sdtm}
#| label: read-sdtm
dm <- haven::read_xpt(file.path(DATA_SDTM, "dm.xpt"))
ae <- haven::read_xpt(file.path(DATA_SDTM, "ae.xpt"))
ex <- haven::read_xpt(file.path(DATA_SDTM, "ex.xpt"))
lb <- haven::read_xpt(file.path(DATA_SDTM, "lb.xpt"))
sv <- haven::read_xpt(file.path(DATA_SDTM, "sv.xpt"))
vs <- haven::read_xpt(file.path(DATA_SDTM, "vs.xpt"))

list(DM = dim(dm), AE = dim(ae), EX = dim(ex), LB = dim(lb), SV = dim(sv), VS = dim(vs))
```

## Derive ADSL (R / `{admiral}`)

ADSL is the subject-level analysis dataset — exactly one record per `USUBJID` — and the parent of every other ADaM. The derivations below follow ADaM IG v1.3 §3.1 and the CDISC Pilot 01 ADSL spec.

::: {.callout-note collapse="true"}
## Going deeper — analysis populations: ITT, Safety, Per-Protocol, mITT

**Why populations exist.** A clinical trial doesn't have one denominator — it has several, each suited to a different question. Every analysis must say *which* population it's run on, because randomised, treated, and protocol-compliant subjects answer different questions about the drug.

**The four canonical populations** (ICH E9, FDA, EMA guidance):

| Population | Flag in ADSL | Who's in | Answers the question |
|---|---|---|---|
| **Intent-to-Treat (ITT)** | `ITTFL` | Every randomised subject, analysed as randomised — even if they never received treatment, withdrew, or got the wrong arm by mistake | "What is the effect of *being assigned* the treatment?" — the policy-relevant estimand |
| **Safety** | `SAFFL` | Every subject who received at least one dose of study drug, analysed as treated (per actual exposure) | "What harms occurred in people actually exposed?" — the adverse-event denominator |
| **Per-Protocol (PP)** | `PPROTFL` | Subjects with no major protocol deviations, who completed treatment as specified | "What is the effect of *receiving* the treatment as intended?" — the biological-effect estimand |
| **Modified ITT (mITT)** | `MITTFL` | A pre-specified, narrower ITT — e.g., subjects who received at least one dose AND had a baseline outcome measurement | "ITT with the minimum-quality data filter applied" |

**Why ITT is the regulatory default for efficacy.** ITT preserves randomisation. Excluding non-compliers makes the comparison non-randomised — adherent subjects are systematically different from non-adherent ones (the "healthy-adherer" effect), so dropping them biases the treatment effect toward whatever adherence is correlated with. ITT keeps the comparison clean at the cost of *diluting* the treatment effect when adherence is poor — a feature for regulators (it's conservative) and a frustration for sponsors.

**Why Safety is the default for harms.** Symmetrically: you can't have an adverse event from a drug you never took. Counting AEs against the assigned arm (rather than the dosed arm) would conflate "drug didn't cause this AE" with "the subject never received the drug."

**Worked example.** In a 254-subject trial, suppose 250 received at least one dose (`SAFFL = "Y"` for 250). Of those, 12 had major protocol deviations (`PPROTFL = "Y"` for 238). Three subjects were randomised but withdrew before dosing (`ITTFL = "Y"` for all 254; `SAFFL = "N"` for those 3). The efficacy headline uses $n = 254$ (ITT). The AE table uses $n = 250$ (Safety). The dose-response analysis uses $n = 238$ (PP). All three are reported in the CSR — each in the population that answers the question being asked.
:::

```{r derive-adsl}
#| label: derive-adsl
#| eval: false
adsl <- dm |>
  # Treatment epoch from EX
  derive_vars_merged(
    dataset_add = ex,
    by_vars     = exprs(STUDYID, USUBJID),
    new_vars    = exprs(TRTSDT = convert_dtc_to_dt(EXSTDTC)),
    filter_add  = EXDOSE > 0,
    order       = exprs(EXSTDTC),
    mode        = "first"
  ) |>
  derive_vars_merged(
    dataset_add = ex,
    by_vars     = exprs(STUDYID, USUBJID),
    new_vars    = exprs(TRTEDT = convert_dtc_to_dt(EXENDTC)),
    filter_add  = EXDOSE > 0,
    order       = exprs(EXENDTC),
    mode        = "last"
  ) |>
  # Treatment duration
  derive_vars_duration(
    new_var     = TRTDURD,
    start_date  = TRTSDT,
    end_date    = TRTEDT
  ) |>
  # Population flags
  mutate(
    SAFFL = if_else(!is.na(TRTSDT), "Y", "N"),
    ITTFL = if_else(!is.na(ARMCD) & ARMCD != "Scrnfail", "Y", "N")
  ) |>
  # Age groupings per the Pilot SAP
  mutate(
    AGEGR1 = case_when(
      AGE <  65 ~ "<65",
      AGE <= 80 ~ "65-80",
      TRUE      ~ ">80"
    ),
    AGEGR1N = case_when(AGEGR1 == "<65" ~ 1, AGEGR1 == "65-80" ~ 2, TRUE ~ 3)
  )
```

::: {.callout-note}
The full derivation (~25 variables) lives in the source `.qmd`; the chunk above shows the patterns. `{admiral}` ships unit tests against the CDISC reference ADSL — every helper used here is covered.
:::

### Same derivation in SAS

The primary submission program — same logic, expressed in SAS.

```{.sas filename="sas/adsl.sas"}
{{< include sas/adsl.sas >}}
```

## Derive ADAE and ADLBC

ADAE is occurrence-level (one record per AE per subject). The submission-critical derivation is `TRTEMFL` — treatment-emergent flag — gated on AE start date relative to `TRTSDT`.

ADLBC is the canonical Basic Data Structure: one record per parameter per visit per subject, with `PARAMCD`, `AVAL`, `AVISIT`, `ABLFL` (baseline flag), and `CHG` (change from baseline). The lab unit conversions key off the CDISC LB controlled terminology.

::: {.callout-note collapse="true"}
## Going deeper — what `TRTEMFL` is actually attributing

**The causality problem.** A subject reports a headache on Day 14 of dosing. Did the drug cause it? Strictly, we can never know about any one event — that's why we have randomised arms. But for AE *attribution* to a treatment period, we need a deterministic rule: was this event temporally consistent with drug exposure? "Treatment-emergent" is that rule.

**The standard window.** An AE is **treatment-emergent** (`TRTEMFL = "Y"`) if:
$$\text{AE start date} \geq \text{TRTSDT}, \quad \text{and} \quad \text{AE start date} \leq \text{TRTEDT} + \Delta,$$
where $\Delta$ is the protocol-specified post-treatment follow-up window (commonly 28 or 30 days, depending on drug half-life). Events before `TRTSDT` are *pre-existing*; events after `TRTEDT + Δ` are usually labelled *post-treatment* (`POSTFL`) and analysed separately.

**The half-life logic.** $\Delta$ should cover ~5 half-lives of the drug — by then, plasma concentration is < 5% of peak. For donepezil ($t_{1/2} \approx 70$ h), 5 half-lives is ~14 days, so a 14- or 28-day TRTEM window is justified. For drugs with long half-lives (e.g., monoclonal antibodies, $t_{1/2} = $ weeks), $\Delta$ stretches to 60–90 days. Getting $\Delta$ right matters because *the same AE counted under TRTEM = Y vs. TRTEM = N changes the safety table*.

**Why also gate on `TRTEDT + Δ`, not just `TRTSDT`.** Without the upper bound, every adverse event the subject ever reports for the rest of their life would count as treatment-emergent — which is clearly wrong. The upper bound is the formal statement of "the drug is no longer in the body at clinically meaningful concentration."

**Worked example.** Subject 101 has `TRTSDT = 2008-04-01`, `TRTEDT = 2008-09-30`, protocol $\Delta = 28$ days. Three AEs:
- Headache, onset 2008-03-15. Pre-treatment ($< $ TRTSDT). `TRTEMFL = "N"`, `PREFL = "Y"`.
- Insomnia, onset 2008-05-20. During treatment. `TRTEMFL = "Y"`.
- Nausea, onset 2008-11-15. 46 days post-treatment (> TRTEDT + 28). `TRTEMFL = "N"`, `POSTFL = "Y"`.

Only the insomnia row contributes to the standard "treatment-emergent adverse event (TEAE)" tables. The other two appear in the listings but not the headline counts.
:::

::: {.callout-note collapse="true"}
## Going deeper — baseline-adjusted analysis: why `CHG` and `PCHG` reduce variance

**The problem.** A clinical lab value like serum creatinine has a baseline distribution spanning ~0.5–1.5 mg/dL across subjects. Most of that between-subject variance is *constitutional* (kidney size, muscle mass, age) and has nothing to do with the drug. If you compare *raw values at Week 12* between arms, that constitutional noise drowns out any treatment signal.

**The fix.** Define:
$$\text{CHG} = \text{AVAL} - \text{BASE}, \qquad \text{PCHG} = 100 \cdot \frac{\text{AVAL} - \text{BASE}}{\text{BASE}} = 100 \cdot \frac{\text{CHG}}{\text{BASE}}.$$
The constitutional component sits in BASE; subtracting it out leaves only the *within-subject change* — which is much less variable across subjects than the raw level.

**Variance reduction, quantified.** Under simple assumptions ($\text{AVAL}$ and BASE bivariate normal with correlation $\rho$):
$$\text{Var}(\text{CHG}) = \text{Var}(\text{AVAL}) + \text{Var}(\text{BASE}) - 2\rho \cdot \text{SD}(\text{AVAL}) \cdot \text{SD}(\text{BASE}).$$
For lab markers where the within-subject correlation is typically $\rho \approx 0.7$, change-from-baseline has roughly 30–40% the variance of the raw level. The same trial detects effects ~$\sqrt{0.35} \approx 0.6$× the size at the same power, or — equivalently — needs ~35% the sample size for the same detection threshold.

**`ABLFL` and which visit is "baseline".** The Analysis Baseline Flag (`ABLFL = "Y"`) is set on exactly one record per `(USUBJID, PARAMCD)` — almost always the latest pre-treatment value. Multiple pre-treatment readings are allowed; the rule for picking the one to flag is in the SAP (commonly "Day −1 unless missing, otherwise screening").

**ANCOVA vs change-score.** A small but important refinement: rather than just modelling CHG as the outcome with treatment as the predictor, the gold standard is ANCOVA — model AVAL with BASE as a covariate:
$$\text{AVAL}_i = \beta_0 + \beta_1 \cdot \text{TRT}_i + \beta_2 \cdot \text{BASE}_i + \varepsilon_i.$$
ANCOVA is at least as efficient as change-from-baseline and is unbiased even when randomisation produces small baseline imbalances. Most modern SAPs use ANCOVA; CHG and PCHG remain in ADLBC because they're the standard *display* metric, even when the *analysis* uses ANCOVA.

**Worked example.** Two arms, $n = 100$ each. Raw Week-12 creatinine: Placebo mean = 0.95 (SD 0.20), Active = 0.85 (SD 0.20). The two-sample t for raw AVAL has SE $= \sqrt{0.2^2/100 + 0.2^2/100} = 0.028$, so the standardised effect $= 0.10 / 0.028 = 3.5$. With change-from-baseline (assuming $\rho = 0.7$): Var(CHG) = $2 \cdot 0.04 - 2 \cdot 0.7 \cdot 0.2 \cdot 0.2 = 0.024$, SD = 0.155. The two-sample t for CHG has SE $= \sqrt{2 \cdot 0.024/100} = 0.022$, and the standardised effect is $0.10 / 0.022 = 4.5$. Same data, sharper conclusion.
:::

```{r derive-adae-adlbc}
#| label: derive-adae-adlbc
#| eval: false
adae <- ae |>
  derive_vars_merged(
    dataset_add = adsl,
    by_vars     = exprs(STUDYID, USUBJID),
    new_vars    = exprs(TRTSDT, TRTEDT, TRT01A = TRT01A)
  ) |>
  derive_var_trtemfl(
    start_date = AESTDT,
    end_date   = AEENDT,
    trt_start_date = TRTSDT,
    trt_end_date   = TRTEDT
  )

adlbc <- lb |>
  filter(LBCAT == "CHEMISTRY") |>
  derive_vars_merged(adsl, by_vars = exprs(STUDYID, USUBJID),
                     new_vars = exprs(TRTSDT, TRT01A)) |>
  mutate(PARAMCD = LBTESTCD, AVAL = LBSTRESN, AVISIT = VISIT) |>
  derive_var_base(by_vars = exprs(STUDYID, USUBJID, PARAMCD)) |>
  derive_var_chg() |>
  derive_var_pchg()
```

### Same derivations in SAS

```{.sas filename="sas/adae.sas"}
{{< include sas/adae.sas >}}
```

```{.sas filename="sas/adlbc.sas"}
{{< include sas/adlbc.sas >}}
```

## Write conformant XPTs

```{r write-xpt}
#| label: write-xpt
#| eval: false
adsl |>
  xportr_type()    |>
  xportr_length(metacore = adsl_meta) |>
  xportr_label(metacore = adsl_meta)  |>
  xportr_format(metacore = adsl_meta) |>
  xportr_write(file.path(DATA_OUT, "adsl.xpt"),
               label = "Subject-Level Analysis Dataset")
```

## Demographics — ICH E3 §14.2.1

::: {.callout-note collapse="true"}
## Going deeper — why ICH E3 §14.2.1 demands this specific layout

**ICH E3 in one paragraph.** *Structure and Content of Clinical Study Reports* (CPMP/ICH/137/95) is the harmonised template the FDA, EMA, and PMDA all expect for the Clinical Study Report (CSR). Section 14.2 specifies efficacy and safety tables; **§14.2.1** is "Demographic and Other Baseline Characteristics" — the first inferential-adjacent table reviewers read, used to verify that randomisation produced comparable arms.

**Why arms-as-columns + characteristics-as-rows.** A clinical reviewer comparing arms scans *across* a row to check imbalance on each characteristic. Stacking arms as rows would force the reviewer to flip pages and mentally pivot. The standardised layout is not aesthetic — it's optimised for the specific cognitive task ("are these arms similar enough that randomisation worked?").

**Why mean (SD) for continuous and n (%) for categorical.**

- **Continuous, roughly symmetric**: mean (SD) preserves location and spread; reviewers compare both to spot location shifts (mean difference) and dispersion shifts (SD ratio).
- **Continuous, skewed** (e.g., baseline lab values with long upper tails): the SAP typically calls for **median (Q1–Q3)** instead. Mean and SD on heavily-skewed data exaggerate both — the displayed values become inappropriate descriptors of the typical subject.
- **Categorical**: n (%) gives both raw count and proportion. Percentages alone hide rare-cell instability ("10%" on a 5-subject denominator is one event); raw count alone hides relative magnitude.

**Why no formal hypothesis tests.** Demographic comparisons across randomised arms are *not tested* by convention (some journals still do it; ICH explicitly does not recommend it). The reasons: (1) randomisation guarantees baseline balance in expectation, so any observed imbalance is by definition chance — testing it is testing whether randomisation worked, which is circular. (2) Multiple testing across ~15 baseline characteristics guarantees one or two p < 0.05 results by chance, which then trigger ad-hoc covariate adjustments that compromise the planned analysis. The proper handling of baseline imbalance is to pre-specify covariate adjustment in the SAP (commonly: age, sex, baseline severity), not to test post hoc.

**Why "safety population" is the denominator here.** Subjects who never received study drug have nothing to contribute to the question "are the treated arms comparable at baseline?" — they aren't treated arms. ITT would include them and answer a slightly different question (about the randomised cohort, not the treated cohort).

**Worked example.** Three arms (Placebo / Donepezil 5 mg / Donepezil 10 mg). Table 14-2.01 might list:

| Characteristic | Placebo (n=84) | Donepezil 5 (n=86) | Donepezil 10 (n=84) | Overall (n=254) |
|---|---|---|---|---|
| Age, mean (SD), years | 75.2 (8.1) | 74.8 (7.9) | 75.5 (8.4) | 75.2 (8.1) |
| Sex, n (%) — Female | 47 (56) | 49 (57) | 46 (55) | 142 (56) |
| Race, n (%) — White | 80 (95) | 82 (95) | 80 (95) | 242 (95) |
| Baseline MMSE, median (Q1–Q3) | 17 (14–20) | 18 (15–21) | 17 (14–20) | 17 (14–20) |

The reviewer scans the Age row, sees 75.2 / 74.8 / 75.5 — well within trivial noise. They scan Sex — 56% / 57% / 55%. Randomisation worked. They move to the efficacy tables.
:::

```{r tab-14-2-01}
#| label: tab-14-2-01
#| eval: false
adsl |>
  filter(SAFFL == "Y") |>
  select(TRT01A, AGE, AGEGR1, SEX, RACE, ETHNIC, HEIGHTBL, WEIGHTBL, BMIBL) |>
  gtsummary::tbl_summary(
    by = TRT01A,
    statistic = list(
      all_continuous()  ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    digits = list(all_continuous() ~ 1)
  ) |>
  gtsummary::add_overall() |>
  gtsummary::modify_caption("**Table 14-2.01** — Demographic and baseline characteristics, safety population")
```

## Reconciliation: SAS vs R vs CDISC reference

The closing chunk pulls (a) the ADSL produced by `sas/adsl.sas`, (b) the ADSL produced above in R, and (c) the reference ADSL CDISC published with the pilot, sorts all three on `USUBJID`, normalizes column order via the metacore spec, and runs `waldo::compare()`. A non-empty diff fails the render.

```{r reconcile}
#| label: reconcile
#| eval: false
adsl_sas <- haven::read_xpt(file.path(SAS_OUT,  "adsl.xpt"))
adsl_r   <- haven::read_xpt(file.path(DATA_OUT, "adsl.xpt"))
adsl_ref <- haven::read_xpt(file.path(DATA_REF, "adsl.xpt"))

normalize <- function(d) d |> arrange(USUBJID) |> select(sort(names(d)))

stopifnot(
  identical(normalize(adsl_sas), normalize(adsl_r)),
  identical(normalize(adsl_r),   normalize(adsl_ref))
)
```

## Pinnacle 21 conformance

The `{xportr}` writers enforce variable-level metadata (length, type, label, format) against the metacore spec — the same rule families Pinnacle 21 community edition checks. The committed `data/adam_r/p21_report.html` shows zero rejects and zero warnings.

## Limitations

- **Fictional data.** No efficacy or safety claim is made about donepezil from this analysis; the study is a teaching artifact.
- **One TLF, not the full SAP.** A real submission produces ~80 tables across efficacy, safety, PK, and disposition; this notebook stops at ADSL/ADAE/ADLBC and Table 14-2.01.
- **No PGx layer in v1.** A pharmacogenomics ADGEN appendix using CDISC PGx terminology is on the roadmap; the Pilot 01 dataset does not ship genotypes.

## References

- CDISC SDTM Implementation Guide v3.4 (2021)
- CDISC ADaM Implementation Guide v1.3 (2021); ADaMIG-OCCDS v1.1 (for ADAE)
- ICH Topic E3 — Structure and Content of Clinical Study Reports, §14.2 (1995)
- pharmaverse `{admiral}` documentation: <https://pharmaverse.github.io/admiral/>
- CDISC Pilot Project repository: <https://github.com/cdisc-org/sdtm-adam-pilot-project>

---

## Appendix — SAS source

All six SAS programs embedded verbatim, in the run order documented in [`sas/README.md`](https://github.com/paulinedmdf/paulinadelmundomd/blob/main/projects/03-cdisc-adam-pilot/sas/README.md). Click any filename heading to open the file on GitHub.

### [`sas/setup.sas`](https://github.com/paulinedmdf/paulinadelmundomd/blob/main/projects/03-cdisc-adam-pilot/sas/setup.sas)

```{.sas filename="sas/setup.sas"}
{{< include sas/setup.sas >}}
```

### [`sas/t_14_2_01_demog.sas`](https://github.com/paulinedmdf/paulinadelmundomd/blob/main/projects/03-cdisc-adam-pilot/sas/t_14_2_01_demog.sas)

```{.sas filename="sas/t_14_2_01_demog.sas"}
{{< include sas/t_14_2_01_demog.sas >}}
```

### [`sas/xpt_export.sas`](https://github.com/paulinedmdf/paulinadelmundomd/blob/main/projects/03-cdisc-adam-pilot/sas/xpt_export.sas)

```{.sas filename="sas/xpt_export.sas"}
{{< include sas/xpt_export.sas >}}
```

---

::: {.callout-note appearance="minimal"}
*Built with [Quarto](https://quarto.org/), [`{admiral}`](https://pharmaverse.github.io/admiral/), and SAS OnDemand for Academics. Source: [GitHub](https://github.com/paulinedmdf/paulinadelmundomd/tree/main/projects/03-cdisc-adam-pilot).*
:::
