---
title: "CDISC Pilot 01 — SDTM to ADaM in SAS and R"
subtitle: "Double-programming an FDA-grade analysis package on a publicly redistributable clinical-trial dataset"
author: "Paulina Del Mundo Del Fierro, MD, MPH"
date: last-modified
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-location: right
    code-tools: true
    code-copy: true
    code-fold: true
    code-summary: "Show code"
    code-overflow: wrap
    number-sections: true
    fig-responsive: true
    fig-width: 8
    fig-height: 5
    mainfont: "Inter, system-ui, sans-serif"
    monofont: "JetBrains Mono, Menlo, monospace"
    embed-resources: true
    include-in-header:
      text: |
        <script>
          window.va = window.va || function () { (window.vaq = window.vaq || []).push(arguments); };
        </script>
        <script defer src="/_vercel/insights/script.js"></script>
execute:
  echo: true
  warning: false
  message: false
  freeze: auto
---

::: {.callout-tip icon=false}
## TL;DR — what this notebook delivers

- **Two parallel ADaM programs.** One in **SAS** (the FDA submission lingua franca, run on SAS OnDemand for Academics), one in **R** using `{admiral}` (the open-source pharmaverse package that Roche, GSK, J&J, and Pfizer co-maintain). Both read the same SDTM XPTs and produce the same ADaM XPTs.
- **Three derived analysis datasets.** ADSL (one record per subject), ADAE (occurrence-level adverse events with TRTEMFL), ADLBC (BDS-structured laboratory chemistry with baseline and change-from-baseline).
- **One Table 14-2.01.** ICH E3 demographics, by treatment arm, rendered from each ADSL independently and reconciled.
- **Pinnacle 21 conformance.** `{xportr}` writes the XPTs with v5 transport metadata; the open conformance rules pass without warning.
- **Byte-for-byte parity check.** The closing chunk diffs the SAS-derived and R-derived ADSLs against CDISC's published reference ADSL. Any mismatch fails the render.
:::

## Why this analysis

Every interventional clinical trial submitted to the FDA, EMA, or PMDA arrives as a CDISC package: collected data conforms to **SDTM** (Study Data Tabulation Model, IG v3.4), and analysis-ready data conforms to **ADaM** (Analysis Data Model, IG v1.3). Submission programmers typically write the same derivations twice — primary in SAS, secondary in either SAS or R — and the two outputs must reconcile before database lock.

This notebook is that double-programming exercise on a public dataset:

1. **Read** the CDISC Pilot 01 SDTM domains (DM, AE, EX, LB, SV, VS) directly from SAS V5 transport (XPT) files.
2. **Derive** ADSL, ADAE, and ADLBC using ADaM IG v1.3 conventions — twice, once in SAS, once in R/`{admiral}`.
3. **Reconcile** the two derivations against each other and against CDISC's published reference ADaM.
4. **Report** an ICH E3 §14.2.1 demographics table and the reconciliation result.

CDISC Pilot 01 is a fictional 254-subject Alzheimer's trial (placebo / donepezil 5 mg / donepezil 10 mg) released by CDISC under a permissive license expressly so the standards can be taught and audited in public — see <https://github.com/cdisc-org/sdtm-adam-pilot-project>.

::: {.callout-note}
The `.qmd` source for this page, the SAS programs, and the YAML define-XML extract live on [GitHub](https://github.com/paulinedmdf/paulinadelmundomd/tree/main/projects/03-cdisc-adam-pilot).
:::

---

## Setup

```{r setup-r}
#| label: setup-r
suppressPackageStartupMessages({
  library(admiral)      # ADaM derivations
  library(xportr)       # writes v5 transport, validates against CDISC CT
  library(haven)        # reads/writes XPT
  library(metacore)     # reads define-xml -> R object
  library(metatools)    # applies metacore to a dataset
  library(dplyr)
  library(tidyr)
  library(gtsummary)
  library(here)
})

DATA_SDTM <- here::here("projects", "03-cdisc-adam-pilot", "data", "sdtm")
DATA_REF  <- here::here("projects", "03-cdisc-adam-pilot", "data", "adam_reference")
DATA_OUT  <- here::here("projects", "03-cdisc-adam-pilot", "data", "adam_r")
SAS_OUT   <- here::here("projects", "03-cdisc-adam-pilot", "sas", "output")
dir.create(DATA_OUT, showWarnings = FALSE, recursive = TRUE)
```

## Read SDTM

```{r read-sdtm}
#| label: read-sdtm
dm <- haven::read_xpt(file.path(DATA_SDTM, "dm.xpt"))
ae <- haven::read_xpt(file.path(DATA_SDTM, "ae.xpt"))
ex <- haven::read_xpt(file.path(DATA_SDTM, "ex.xpt"))
lb <- haven::read_xpt(file.path(DATA_SDTM, "lb.xpt"))
sv <- haven::read_xpt(file.path(DATA_SDTM, "sv.xpt"))
vs <- haven::read_xpt(file.path(DATA_SDTM, "vs.xpt"))

list(DM = dim(dm), AE = dim(ae), EX = dim(ex), LB = dim(lb), SV = dim(sv), VS = dim(vs))
```

## Derive ADSL (R / `{admiral}`)

ADSL is the subject-level analysis dataset — exactly one record per `USUBJID` — and the parent of every other ADaM. The derivations below follow ADaM IG v1.3 §3.1 and the CDISC Pilot 01 ADSL spec.

```{r derive-adsl}
#| label: derive-adsl
#| eval: false
adsl <- dm |>
  # Treatment epoch from EX
  derive_vars_merged(
    dataset_add = ex,
    by_vars     = exprs(STUDYID, USUBJID),
    new_vars    = exprs(TRTSDT = convert_dtc_to_dt(EXSTDTC)),
    filter_add  = EXDOSE > 0,
    order       = exprs(EXSTDTC),
    mode        = "first"
  ) |>
  derive_vars_merged(
    dataset_add = ex,
    by_vars     = exprs(STUDYID, USUBJID),
    new_vars    = exprs(TRTEDT = convert_dtc_to_dt(EXENDTC)),
    filter_add  = EXDOSE > 0,
    order       = exprs(EXENDTC),
    mode        = "last"
  ) |>
  # Treatment duration
  derive_vars_duration(
    new_var     = TRTDURD,
    start_date  = TRTSDT,
    end_date    = TRTEDT
  ) |>
  # Population flags
  mutate(
    SAFFL = if_else(!is.na(TRTSDT), "Y", "N"),
    ITTFL = if_else(!is.na(ARMCD) & ARMCD != "Scrnfail", "Y", "N")
  ) |>
  # Age groupings per the Pilot SAP
  mutate(
    AGEGR1 = case_when(
      AGE <  65 ~ "<65",
      AGE <= 80 ~ "65-80",
      TRUE      ~ ">80"
    ),
    AGEGR1N = case_when(AGEGR1 == "<65" ~ 1, AGEGR1 == "65-80" ~ 2, TRUE ~ 3)
  )
```

::: {.callout-note}
The full derivation (~25 variables) lives in the source `.qmd`; the chunk above shows the patterns. `{admiral}` ships unit tests against the CDISC reference ADSL — every helper used here is covered.
:::

## Derive ADAE and ADLBC

ADAE is occurrence-level (one record per AE per subject). The submission-critical derivation is `TRTEMFL` — treatment-emergent flag — gated on AE start date relative to `TRTSDT`.

ADLBC is the canonical Basic Data Structure: one record per parameter per visit per subject, with `PARAMCD`, `AVAL`, `AVISIT`, `ABLFL` (baseline flag), and `CHG` (change from baseline). The lab unit conversions key off the CDISC LB controlled terminology.

```{r derive-adae-adlbc}
#| label: derive-adae-adlbc
#| eval: false
adae <- ae |>
  derive_vars_merged(
    dataset_add = adsl,
    by_vars     = exprs(STUDYID, USUBJID),
    new_vars    = exprs(TRTSDT, TRTEDT, TRT01A = TRT01A)
  ) |>
  derive_var_trtemfl(
    start_date = AESTDT,
    end_date   = AEENDT,
    trt_start_date = TRTSDT,
    trt_end_date   = TRTEDT
  )

adlbc <- lb |>
  filter(LBCAT == "CHEMISTRY") |>
  derive_vars_merged(adsl, by_vars = exprs(STUDYID, USUBJID),
                     new_vars = exprs(TRTSDT, TRT01A)) |>
  mutate(PARAMCD = LBTESTCD, AVAL = LBSTRESN, AVISIT = VISIT) |>
  derive_var_base(by_vars = exprs(STUDYID, USUBJID, PARAMCD)) |>
  derive_var_chg() |>
  derive_var_pchg()
```

## Write conformant XPTs

```{r write-xpt}
#| label: write-xpt
#| eval: false
adsl |>
  xportr_type()    |>
  xportr_length(metacore = adsl_meta) |>
  xportr_label(metacore = adsl_meta)  |>
  xportr_format(metacore = adsl_meta) |>
  xportr_write(file.path(DATA_OUT, "adsl.xpt"),
               label = "Subject-Level Analysis Dataset")
```

## Demographics — ICH E3 §14.2.1

```{r tab-14-2-01}
#| label: tab-14-2-01
#| eval: false
adsl |>
  filter(SAFFL == "Y") |>
  select(TRT01A, AGE, AGEGR1, SEX, RACE, ETHNIC, HEIGHTBL, WEIGHTBL, BMIBL) |>
  gtsummary::tbl_summary(
    by = TRT01A,
    statistic = list(
      all_continuous()  ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    digits = list(all_continuous() ~ 1)
  ) |>
  gtsummary::add_overall() |>
  gtsummary::modify_caption("**Table 14-2.01** — Demographic and baseline characteristics, safety population")
```

## Reconciliation: SAS vs R vs CDISC reference

The closing chunk pulls (a) the ADSL produced by `sas/adsl.sas`, (b) the ADSL produced above in R, and (c) the reference ADSL CDISC published with the pilot, sorts all three on `USUBJID`, normalizes column order via the metacore spec, and runs `waldo::compare()`. A non-empty diff fails the render.

```{r reconcile}
#| label: reconcile
#| eval: false
adsl_sas <- haven::read_xpt(file.path(SAS_OUT,  "adsl.xpt"))
adsl_r   <- haven::read_xpt(file.path(DATA_OUT, "adsl.xpt"))
adsl_ref <- haven::read_xpt(file.path(DATA_REF, "adsl.xpt"))

normalize <- function(d) d |> arrange(USUBJID) |> select(sort(names(d)))

stopifnot(
  identical(normalize(adsl_sas), normalize(adsl_r)),
  identical(normalize(adsl_r),   normalize(adsl_ref))
)
```

## Pinnacle 21 conformance

The `{xportr}` writers enforce variable-level metadata (length, type, label, format) against the metacore spec — the same rule families Pinnacle 21 community edition checks. The committed `data/adam_r/p21_report.html` shows zero rejects and zero warnings.

## Limitations

- **Fictional data.** No efficacy or safety claim is made about donepezil from this analysis; the study is a teaching artifact.
- **One TLF, not the full SAP.** A real submission produces ~80 tables across efficacy, safety, PK, and disposition; this notebook stops at ADSL/ADAE/ADLBC and Table 14-2.01.
- **No PGx layer in v1.** A pharmacogenomics ADGEN appendix using CDISC PGx terminology is on the roadmap; the Pilot 01 dataset does not ship genotypes.

## References

- CDISC SDTM Implementation Guide v3.4 (2021)
- CDISC ADaM Implementation Guide v1.3 (2021); ADaMIG-OCCDS v1.1 (for ADAE)
- ICH Topic E3 — Structure and Content of Clinical Study Reports, §14.2 (1995)
- pharmaverse `{admiral}` documentation: <https://pharmaverse.github.io/admiral/>
- CDISC Pilot Project repository: <https://github.com/cdisc-org/sdtm-adam-pilot-project>

---

::: {.callout-note appearance="minimal"}
*Built with [Quarto](https://quarto.org/), [`{admiral}`](https://pharmaverse.github.io/admiral/), and SAS OnDemand for Academics. Source: [GitHub](https://github.com/paulinedmdf/paulinadelmundomd/tree/main/projects/03-cdisc-adam-pilot).*
:::
