---
title: "The $35 Insulin Cap and Medicare Part D"
subtitle: "A difference-in-differences evaluation of the Inflation Reduction Act Section 11406 out-of-pocket cap"
author: "Paulina Del Mundo Del Fierro, MD, MPH"
date: last-modified
format:
  html:
    theme: [cosmo, ../_theme.scss]
    highlight-style: github-dark
    toc: true
    toc-depth: 3
    toc-location: right
    code-tools: true
    code-copy: true
    code-fold: true
    code-summary: "Show code"
    code-overflow: wrap
    number-sections: true
    fig-responsive: true
    fig-width: 8
    fig-height: 5
    embed-resources: true
    include-in-header:
      text: |
        <script>
          window.va = window.va || function () { (window.vaq = window.vaq || []).push(arguments); };
        </script>
        <script defer src="/_vercel/insights/script.js"></script>
execute:
  echo: true
  warning: false
  message: false
  freeze: auto
---

::: {.callout-tip icon=false}
## Key findings — TL;DR

- **The IRA's $35 insulin cap is a clean natural experiment.** Treatment hits one therapeutic class on one date (2023-01-01), with non-insulin antihyperglycemics as an internally valid control group — a setting where **two-way fixed effects (TWFE)** are unbiased without the staggered-adoption corrections that more complex DiD designs require.
- **Primary outcome — log total 30-day fills at the drug-year level.** A positive **ATT (average treatment effect on the treated)** is the expected sign: a binding price cap that lowers beneficiary out-of-pocket cost should raise utilization. The 95% CI and event-study coefficients populate from `data/processed/drug_year_panel.parquet`.
- **Pre-trends are the load-bearing assumption** — and the event-study plot is where it has to be defended. The 2021 coefficient is *not* expected to be zero because the CMS Senior Savings Model partially treated MA-PD insulin enrollees that year; the IRA effect is the marginal jump in 2023 once the cap went universal.
- **Methods used (every claim ties back to a specific chunk):** DuckDB streaming filter of the CMS Part D Prescriber PUFs, drug-year and prescriber-drug-year panels, TWFE DiD with drug-clustered SEs, event-study with a parallel-trends F-test, placebo cap-year tests, leave-one-out drug robustness, prescriber-FE specification, and specialty-stratified heterogeneity. Each step has an inline "**Why this test?**" callout explaining what it tests and why it earns its place at this specific moment in the analysis.
:::

## Why this analysis

Medicare Part D became the headline beneficiary-facing piece of the Inflation Reduction Act of 2022. Section 11406 capped per-month insulin out-of-pocket cost at $35 across all Part D plans starting 2023-01-01 — the first time Medicare imposed a uniform per-product price ceiling on cost-sharing. Whether that cap *moved utilization* (rather than merely shifting cost from beneficiary to plan) is the sharp empirical question, and it has a textbook DiD design baked in: insulin is treated, every other antihyperglycemic is not, treatment is simultaneous, and CMS publishes annual prescriber-drug data for free.

This notebook does four things end-to-end:

1. **Builds the panel.** The DuckDB pipeline in [`prep.py`](prep.py) streams six annual releases of the CMS Medicare Part D Prescribers — by Provider and Drug PUF (2019–2024), filters to insulin and a curated set of non-insulin antihyperglycemic comparators, and writes a small drug-year panel and a larger prescriber-drug-year panel.
2. **Establishes the design.** A pre-period descriptive walk through utilization trajectories, an explicit statement of the parallel-trends assumption, and a discussion of why simultaneous treatment makes plain TWFE unbiased here (Borusyak–Jaravel–Spiess 2024 critique does not bind).
3. **Runs the DiD.** Two-way FE estimate with SEs clustered at the drug level, an event-study with year-by-treatment interactions, a joint F-test on the pre-period leads, and placebo cap-year tests at 2020 and 2021.
4. **Stress-tests it.** Leave-one-out robustness over comparator drugs (drop GLP-1 RAs to absorb the semaglutide / tirzepatide demand shock), a prescriber-FE specification on the larger panel, and heterogeneity stratified by prescriber specialty (endocrinology, primary care, other).

::: {.callout-note}
## Data sources
- **CMS Medicare Part D Prescribers — by Provider and Drug.** Centers for Medicare & Medicaid Services, annual public use files DY2019–DY2024. <https://data.cms.gov/provider-summary-by-type-of-service/medicare-part-d-prescribers>
- **IRA Section 11406** — the Inflation Reduction Act of 2022 statutory text capping Part D insulin OOP at $35/month effective 2023-01-01.
- **CMS Part D Senior Savings Model** — a 2021–2024 demonstration that capped insulin OOP at $35/month for participating MA-PD enrollees, predating the universal IRA cap. <https://www.cms.gov/priorities/innovation/innovation-models/part-d-senior-savings-model>

Raw CSVs stay local under `data/raw/` (six files, ~30 GB unzipped); [`prep.py`](prep.py) emits the three committed Parquets this notebook reads.
:::

---

## Setup

```{r setup-r}
suppressPackageStartupMessages({
  library(arrow)
  library(dplyr)
  library(tidyr)
  library(stringr)
  library(forcats)
  library(scales)
  library(ggplot2)
  library(fixest)
  library(broom)
  library(purrr)
  library(here)
})

# Method packages installed on first render if absent.
for (pkg in c("kableExtra")) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    install.packages(pkg, repos = "https://cloud.r-project.org")
  }
}

theme_set(theme_minimal(base_size = 12))

pretty_kable <- function(x, caption = NULL, ...) {
  knitr::kable(x, caption = caption, ...) |>
    kableExtra::kable_styling(
      bootstrap_options = c("striped", "hover", "condensed", "responsive"),
      full_width        = FALSE,
      position          = "left",
      font_size         = 13
    )
}

data_dir <- here::here("projects", "04-medicare-part-d-insulin-cap",
                      "data", "processed")

drug_year   <- arrow::read_parquet(file.path(data_dir, "drug_year_panel.parquet"))
prescriber  <- arrow::read_parquet(file.path(data_dir, "prescriber_drug_year_panel.parquet"))
classify    <- arrow::read_parquet(file.path(data_dir, "drug_classification.parquet"))

# Reference year for the event study: 2022 is the last full calendar year
# before the IRA cap takes effect, so its treatment-interaction coefficient
# is normalized to zero. The IRA effective date is 2023-01-01.
TREAT_REF_YEAR <- 2022L
TREAT_FIRST_YEAR <- 2023L

# Construct the analytic panel: drop "Other"-classed strays (should be empty
# given the prep filter), require non-zero fills (a generic with zero claims
# nationwide in a year is informationless), and add the post indicator.
panel <- drug_year |>
  filter(drug_class != "Other") |>
  filter(tot_30day_fills > 0) |>
  mutate(
    post = as.integer(year >= TREAT_FIRST_YEAR),
    treated_post = treated * post,
    log_fills    = log(tot_30day_fills),
    log_clms     = log(tot_clms),
    log_supply   = log(tot_day_suply),
    log_cost     = log(tot_drug_cst)
  )
```

## The panel

```{r panel-shape}
panel_shape <- panel |>
  summarise(
    drugs       = n_distinct(gnrc_name),
    insulin_n   = n_distinct(gnrc_name[treated == 1]),
    control_n   = n_distinct(gnrc_name[treated == 0]),
    years       = paste(range(year), collapse = "-"),
    rows        = n()
  )

pretty_kable(panel_shape, caption = "Drug-year panel — shape after prep filter")
```

```{r panel-by-class}
panel |>
  group_by(drug_class, treated) |>
  summarise(
    drugs        = n_distinct(gnrc_name),
    drug_years   = n(),
    tot_30day    = sum(tot_30day_fills),
    .groups = "drop"
  ) |>
  mutate(
    arm = if_else(treated == 1, "Treated (insulin)", "Control (non-insulin)"),
    pct_30day = tot_30day / sum(tot_30day) * 100
  ) |>
  arrange(desc(treated), desc(tot_30day)) |>
  select(arm, drug_class, drugs, drug_years, tot_30day, pct_30day) |>
  pretty_kable(
    caption = "Composition of the analytic panel by drug class and treatment arm",
    digits  = c(0, 0, 0, 0, 0, 1),
    col.names = c("Arm", "Drug class", "Generics",
                  "Drug-years", "Total 30-day fills", "% of total")
  )
```

## Pre-period trends — the assumption to defend

```{r trends-class}
base_year <- min(panel$year)

class_year <- panel |>
  group_by(year, drug_class, treated) |>
  summarise(tot_30day = sum(tot_30day_fills), .groups = "drop") |>
  group_by(drug_class) |>
  mutate(idx = tot_30day / tot_30day[year == base_year] * 100) |>
  ungroup()

ggplot(class_year, aes(year, idx,
                       colour = drug_class,
                       linetype = factor(treated, labels = c("Control", "Treated")))) +
  geom_vline(xintercept = TREAT_FIRST_YEAR - 0.5,
             colour = "grey50", linetype = "dotted") +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.6) +
  scale_x_continuous(breaks = sort(unique(class_year$year))) +
  scale_y_continuous(labels = label_number(suffix = "")) +
  scale_linetype_manual(values = c("Control" = "solid", "Treated" = "longdash")) +
  labs(
    title    = sprintf("Total 30-day fills by drug class, indexed to %d", base_year),
    subtitle = "Vertical line marks the IRA cap (effective 2023-01-01)",
    x = NULL, y = sprintf("Total 30-day fills (%d = 100)", base_year),
    colour = "Drug class", linetype = NULL
  ) +
  theme(legend.position = "right")
```

The visual diagnostic above is the parallel-trends story in pictures. The DiD coefficient below is the same story formalized — but the picture is the one a reviewer trusts first.

::: {.callout-note}
## Why difference-in-differences, and why "parallel trends" is the whole game

A **difference-in-differences (DiD)** asks: *did the gap between the treated and untreated groups change after the policy, beyond what was already changing before?* Concretely, we take two differences:

1. **Within-group change**: insulin volume in 2023 minus insulin volume in 2022.
2. **Between-group difference of those changes**: the insulin within-group change minus the equivalent within-group change for non-insulin antihyperglycemics.

If the gap between groups was stable in the pre-period and then jumped in the post-period, the jump is the causal effect of the policy *under one assumption* — the **parallel-trends assumption**. Parallel trends says: in the absence of the policy, the gap between treated and control would have continued to evolve in parallel. We can never observe the counterfactual ("what would insulin have done in 2023 without the cap?"), so we use the control group's actual 2023 trajectory as the stand-in. That stand-in is only valid if pre-period gaps were stable.

This is why every DiD analysis from here forward is, in some sense, an argument about whether parallel trends hold. The point estimate is one regression coefficient. The credibility of that coefficient depends on whatever evidence we can muster — visual, formal, placebo — that the assumption isn't violated. (Card & Krueger 1994; Angrist & Pischke 2009)
:::

## Primary specification — two-way fixed effects

The specification is

$$
\log \text{Fills}_{it} = \beta \cdot (\text{Treated}_i \times \text{Post}_t) + \alpha_i + \gamma_t + \varepsilon_{it}
$$

where $i$ indexes a generic name, $t$ indexes calendar year, $\alpha_i$ is a drug fixed effect, $\gamma_t$ is a year fixed effect, and $\text{Post}_t = \mathbb{1}[t \geq 2023]$. SEs are clustered at the drug level (the level at which treatment is assigned). $\beta$ is the average treatment effect on the treated (ATT) — the relative change in insulin's log 30-day fills caused by the policy.

::: {.callout-note}
## Why two-way fixed effects, and why cluster the SEs at the drug level?

**Drug fixed effects ($\alpha_i$)** absorb every time-invariant difference between drugs. Insulin glargine is a much higher-volume drug than insulin glulisine, and metformin dwarfs them both — but those baseline differences don't vary across years, so they get soaked up by the drug FE. What's left after the drug FE is *each drug's deviation from its own average over time.*

**Year fixed effects ($\gamma_t$)** absorb every nationwide shock that hit all drugs equally — formulary tier changes, pandemic-era prescribing shifts, broad reimbursement adjustments. After the year FE, what's left is *each drug-year's deviation from its drug's average AND from the year's average.* The TWFE estimator is identifying the treatment effect from exactly this residual variation: the "extra" change in insulin specifically, in 2023 specifically, beyond what the drug-level and year-level baselines predict.

**Clustering SEs at the drug level** corrects for the fact that observations within the same drug across years are not statistically independent. Insulin glargine in 2022 and insulin glargine in 2023 share unobserved supply, formulary, and demand factors. Naive ("classical") SEs would treat each drug-year as independent and produce confidence intervals that are too narrow — a known recipe for false-positive findings. We cluster at the level at which treatment is assigned (the drug); that's the standard rule. (Cameron & Miller 2015; Abadie et al. 2023)
:::

::: {.callout-note collapse="true"}
## Going deeper — the Liang–Zeger cluster-robust ("sandwich") variance

**Plain English.** Ordinary regression assumes every observation is *independent* — like rolling a fresh die each time. But real-world panel data isn't independent: observations within the same drug (or the same patient, or the same firm) are correlated because they share unobserved factors. If you ignore that and use ordinary standard errors, your confidence intervals come out *too narrow* and you'll claim significance for results that aren't really there. The "sandwich" estimator (named because the formula visually looks like bread–meat–bread) fixes this by *estimating* the within-group correlations directly from your residuals instead of assuming independence. You don't have to specify the correlation structure — the sandwich figures it out empirically. The standard rule is "cluster the SEs at the level where treatment was assigned," which here is the drug.

**Intuition.** OLS gives one variance formula assuming all errors are i.i.d.: $\widehat{\text{Var}}(\hat\beta) = \hat\sigma^2 (\mathbf{X}^\top \mathbf{X})^{-1}$. If errors are correlated within clusters (drugs), that formula understates variance because it pretends each within-cluster observation is independent new evidence — when in reality they carry overlapping information. The cluster-robust ("sandwich") estimator replaces the i.i.d. assumption with an empirical estimate of the within-cluster correlation structure, *whatever* it happens to be.

**The sandwich formula.** With $G$ clusters indexed by $g$, residuals $\hat e_g$, and design block $\mathbf{X}_g$ for cluster $g$:
$$\widehat{\text{Var}}_{\text{CR}}(\hat\beta) = (\mathbf{X}^\top \mathbf{X})^{-1} \underbrace{\!\left[ \sum_{g=1}^G \mathbf{X}_g^\top \hat{\mathbf e}_g \hat{\mathbf e}_g^\top \mathbf{X}_g \right]}_{\text{the "meat"}} (\mathbf{X}^\top \mathbf{X})^{-1} \cdot c_G,$$
where $c_G = \tfrac{G}{G-1} \cdot \tfrac{N-1}{N-p}$ is a small-cluster finite-sample correction. The two outer terms (the "bread") are the same as OLS. The middle term (the "meat") sums *across* clusters the *outer product* of within-cluster residuals — this lets within-cluster residuals contribute jointly to the variance, not as independent draws.

**Why "sandwich".** $(\mathbf{X}^\top\mathbf{X})^{-1} \cdot [\text{meat}] \cdot (\mathbf{X}^\top\mathbf{X})^{-1}$ — bread, meat, bread.

**What independence assumption is replaced.** OLS assumes $E[\hat{\mathbf e}\hat{\mathbf e}^\top] = \sigma^2 \mathbf{I}$. Cluster-robust assumes only that the matrix is block-diagonal: zero between clusters, *any* structure within. Within-cluster heteroskedasticity, autocorrelation, and unobserved random effects are all absorbed. (HC1 — heteroskedasticity-robust without clustering — is the special case where each cluster has size 1.)

**Asymptotic regime.** Cluster-robust SEs are consistent as $G \to \infty$ (clusters grow), *not* as $N \to \infty$ within a fixed number of clusters. This is why **the number of clusters, not the total sample size, drives reliability**.

**Worked example.** 30 drug clusters, 6 years each ($N = 180$, $p = 30 + 6 + 1 = 37$). The finite-sample correction is $c_{30} = \tfrac{30}{29} \cdot \tfrac{179}{143} \approx 1.30$. If the classical OLS SE on `treated_post` is 0.020 and the i.i.d. assumption is wildly wrong (residuals are positively autocorrelated within drug), the cluster-robust SE might inflate to 0.045 — more than 2×. The DiD coefficient's 95% CI widens proportionally; what looked statistically significant at the i.i.d. SE could lose significance at the cluster-robust SE — and the cluster-robust version is the honest one.
:::

```{r did-primary}
m_fills  <- feols(log_fills  ~ treated_post | gnrc_name + year, data = panel, cluster = ~gnrc_name)
m_clms   <- feols(log_clms   ~ treated_post | gnrc_name + year, data = panel, cluster = ~gnrc_name)
m_supply <- feols(log_supply ~ treated_post | gnrc_name + year, data = panel, cluster = ~gnrc_name)
m_cost   <- feols(log_cost   ~ treated_post | gnrc_name + year, data = panel, cluster = ~gnrc_name)

did_table <- bind_rows(
  tidy(m_fills,  conf.int = TRUE) |> mutate(outcome = "log 30-day fills"),
  tidy(m_clms,   conf.int = TRUE) |> mutate(outcome = "log claims"),
  tidy(m_supply, conf.int = TRUE) |> mutate(outcome = "log day-supply"),
  tidy(m_cost,   conf.int = TRUE) |> mutate(outcome = "log drug cost")
) |>
  filter(term == "treated_post") |>
  transmute(
    outcome,
    estimate, std.error, conf.low, conf.high,
    p_value = p.value
  )

pretty_kable(
  did_table,
  caption = "Primary DiD — Treated x Post-2023 coefficient (drug-clustered SEs)",
  digits  = 4
)
```

The `treated_post` coefficient is the log-percentage change in the outcome for insulin relative to the non-insulin counterfactual after 2023. Because the outcome is logged, the coefficient is approximately a relative change: $\beta = 0.05$ implies roughly $\exp(0.05) - 1 \approx 5.1\%$ more 30-day fills for insulin than the control-group counterfactual predicts. A negative coefficient says the opposite — insulin grew *less* than controls.

::: {.callout-note}
## Translating the ATT for a non-statistical audience

A policy team or executive sponsor will not read "ATT on log 30-day fills = 0.085 (SE 0.030)" — they need three translation steps before that number is decision-useful:

1. **Relative effect.** The point estimate exponentiates to $e^{0.085} - 1 \approx 8.9\%$. Headline: *"after the IRA cap took effect, insulin 30-day fills grew 8.9% faster than the non-insulin antihyperglycemic counterfactual would have predicted."*
2. **Confidence-interval translation.** The 95% CI in log-space exponentiates asymmetrically. If the log-CI is $[0.026, 0.144]$, the relative-effect CI is roughly $[2.6\%, 15.5\%]$ — wide, but excludes zero.
3. **Absolute-scale anchor.** Multiply the percent change by the pre-period insulin baseline. If insulin averaged 25 million 30-day fills per year before the cap, an 8.9% effect = ~2.2 million additional fills per year attributable to the policy. That's the number policy audiences engage with.

Steps 1 and 2 are arithmetic; step 3 requires the pre-period denominator. Always carry the absolute anchor into the executive summary, not just the log coefficient. The drug-clustered SE and the few-cluster caveats above apply to all three layers of the translation equally — the reported numbers inherit the inference framework, they don't get to bypass it.
:::

::: {.callout-note collapse="true"}
## Going deeper — interpreting a log-outcome coefficient as a percent change

**Plain English.** When your outcome variable in a regression is the *logarithm* of something (log of fills, log of sales), a coefficient of $0.05$ doesn't mean "the outcome increases by 0.05" — it means "the *log* of the outcome increases by 0.05." To translate that to percent change in the actual outcome, you compute $e^{0.05} - 1 \approx 0.0513$, or about 5.1%. For *small* coefficients (under ~0.10) the coefficient itself is close enough to the percent change. For bigger coefficients the gap grows and you should report the exponentiated version explicitly. The reason regressions use log-outcomes in the first place is that *percent* changes — not absolute changes — are usually the meaningful quantity for things like prescription volumes or sales (doubling fills is the same magnitude of event whether you start at 100 or 10,000).

**Setup.** When the outcome is $\log Y$, a one-unit increase in a regressor $X$ changes $\log Y$ by $\beta$ — i.e., $Y$ scales by $e^\beta$. So the percent change in $Y$ is exactly
$$\frac{\Delta Y}{Y} = e^{\beta} - 1.$$
For small $\beta$, the Taylor expansion $e^\beta = 1 + \beta + \tfrac{\beta^2}{2} + \cdots$ gives $e^\beta - 1 \approx \beta$. So *for small effects*, "the coefficient is approximately the percent change" — but the approximation breaks down quickly:

| $\beta$ | $e^\beta - 1$ (exact) | $\beta$ (approx) | Approximation error |
|---:|---:|---:|---:|
| 0.01 | 1.005% | 1.000% | 0.5% |
| 0.05 | 5.127% | 5.000% | 2.5% |
| 0.10 | 10.517% | 10.000% | 5.2% |
| 0.25 | 28.403% | 25.000% | 13.6% |
| 0.50 | 64.872% | 50.000% | 29.7% |

**Rule of thumb.** For $|\beta| < 0.10$ the approximation is within ~5%; report either. For $|\beta| \geq 0.10$ always use the exact $e^\beta - 1$, otherwise you're meaningfully understating the effect.

**The delta-method CI.** A 95% CI for $\beta$ on the log scale, $[\beta_L, \beta_U]$, exponentiates to a 95% CI for the multiplicative effect $e^\beta$: $[e^{\beta_L}, e^{\beta_U}]$. Note this is *asymmetric* on the original scale — $e^x$ is convex, so the upper half of the CI is wider than the lower half once you map back. Reporting "ATT = 7.2% (95% CI: 3.1%, 11.5%)" rather than "$\beta = 0.07$ (SE 0.02)" is the publication-standard transform.

**Why log in the first place.** Three reasons drive the choice:
1. **Multiplicative effects are natural for utilization.** Doubling fills is the same kind of event whether you start at 100 or 10,000; absolute differences are not.
2. **Heteroskedasticity shrinks.** Pharmaceutical fill counts have variance growing with the mean. $\log Y$ stabilises that — the residuals from a log model are usually approximately homoskedastic, satisfying OLS assumptions.
3. **Skewness shrinks.** Fill counts are right-skewed; $\log$ pulls the tail in, producing residuals closer to normal and making asymptotic inference more accurate at modest $N$.

**Worked example.** Suppose the DiD on log 30-day fills returns $\hat\beta = 0.085$ with SE 0.030, $z = 2.83$, 95% CI $[0.026, 0.144]$ on the log scale. Translating:
$$\text{Point estimate}: \; e^{0.085} - 1 = 8.87\%.$$
$$\text{95% CI}: \; [e^{0.026} - 1, \; e^{0.144} - 1] = [2.63\%, \; 15.49\%].$$
"The IRA cap is associated with an 8.9% increase in insulin 30-day fills relative to the non-insulin counterfactual (95% CI: 2.6%, 15.5%)." The naive $\beta \approx 8.5\%$ approximation understates by 0.4 points — small but trending toward "meaningful" at this effect size.
:::

## Event study — does the effect line up with the policy date?

::: {.callout-note}
## Why an event-study, when we already have a single DiD number?

The DiD compresses the entire treatment effect into one number — the average jump after 2023. That's useful for headlines but hides three things a careful reader wants to see:

1. **Pre-trends.** Were treated and control already moving apart *before* the policy? An event-study traces the treated-vs-control gap *year by year*, with the year right before treatment normalized to zero. Pre-period coefficients should hover near zero if parallel trends hold; if they drift, the DiD point estimate is contaminated by a pre-existing trend that has nothing to do with the policy.
2. **Anticipation.** Markets sometimes respond before a policy takes effect — manufacturers, plans, and prescribers anticipated the IRA cap from August 2022 (when the bill was signed) onward. If insulin's 2022 coefficient is already non-zero, that's evidence of anticipation, and the DiD's "post" coefficient is a mix of anticipation + actual policy impact.
3. **Timing of the response.** Did the effect appear sharply in 2023 or build over multiple post-years? A pile of post-period coefficients lets you see the dose-response curve over time, not just the average.

Mechanically, the event-study replaces the single $\text{Treated} \times \text{Post}$ term with $\text{Treated} \times \mathbb{1}[t = k]$ interactions for each year $k$, omitting one reference year (here, 2022 — the last full pre-IRA calendar year). Each coefficient reads as: *the gap between insulin and control in year $k$, relative to the gap in 2022.* (Sun & Abraham 2021; Borusyak, Jaravel & Spiess 2024)
:::

```{r event-study}
# Build year-by-treated interaction with 2022 as the omitted reference.
panel_es <- panel |>
  mutate(year_f = factor(year))

es <- feols(
  log_fills ~ i(year_f, treated, ref = as.character(TREAT_REF_YEAR)) |
              gnrc_name + year_f,
  data = panel_es, cluster = ~gnrc_name
)

es_tidy <- broom::tidy(es, conf.int = TRUE) |>
  filter(str_detect(term, "year_f::")) |>
  mutate(
    year = as.integer(str_extract(term, "(?<=year_f::)\\d{4}")),
    period = if_else(year < TREAT_FIRST_YEAR, "Pre", "Post")
  ) |>
  bind_rows(tibble(
    year = TREAT_REF_YEAR, estimate = 0, conf.low = 0, conf.high = 0,
    period = "Pre", term = "reference"
  )) |>
  arrange(year)

ggplot(es_tidy, aes(year, estimate, colour = period)) +
  geom_hline(yintercept = 0, colour = "grey50") +
  geom_vline(xintercept = TREAT_FIRST_YEAR - 0.5,
             colour = "grey50", linetype = "dotted") +
  geom_pointrange(aes(ymin = conf.low, ymax = conf.high), size = 0.4) +
  geom_line(aes(group = 1), linewidth = 0.4, alpha = 0.6) +
  scale_x_continuous(breaks = sort(unique(es_tidy$year))) +
  scale_colour_manual(values = c(Pre = "#5c5c5c", Post = "#020073")) +
  labs(
    title    = "Event-study coefficients on log(30-day fills)",
    subtitle = sprintf("Reference year %d; vertical line marks the IRA cap effective date",
                       TREAT_REF_YEAR),
    x = NULL, y = "Treated x year coefficient (95% CI)",
    colour = NULL
  )
```

## Parallel-trends F-test

::: {.callout-note}
## Why the parallel-trends F-test, and why a non-rejection isn't proof

The event-study plot is a visual check; the F-test is the formal statement. A **joint Wald F-test** asks whether *all the pre-period coefficients are jointly equal to zero.* If we cannot reject that joint null, the data are at least consistent with parallel trends. If we *do* reject — meaning at least one pre-period coefficient is reliably different from zero — the parallel-trends assumption has measurable violations, and the post-period DiD coefficient is no longer a clean treatment effect.

Two important asymmetries in how to read this test:

1. **A non-rejection ($p > 0.05$) does not prove parallel trends.** It only fails to falsify them. Underpowered designs with few pre-period years and noisy outcomes will fail to reject *any* null. A small number of pre-period coefficients (this smoke test has only one) makes the test almost guaranteed to fail to reject — which is why we explicitly note when the test is statistically meaningful and when it is essentially uninformative.
2. **A rejection ($p < 0.05$) is fatal in a way that's hard to fix.** You can sometimes argue the design is still valid (e.g., if the violation is in a single drug you can drop it via leave-one-out), but the burden of proof shifts to the analyst. Many published DiD papers fail at exactly this check, which is why this section is run *before* anyone is allowed to interpret the post-period coefficient.

We run the F-test twice: once on all pre-period leads, and once dropping 2021 (which is partially treated by the CMS Senior Savings Model and therefore would be expected to deviate from zero even under valid parallel trends). The second test is the cleaner check.
:::

::: {.callout-note collapse="true"}
## Going deeper — the Wald F-statistic formula and what it tests

**Plain English.** Sometimes you want to test, not "is *this one* coefficient zero?", but "are these *three* coefficients all jointly zero, considered together?" The Wald F-test is the tool for that joint question. It's essentially a multidimensional t-test: take your three coefficient estimates, divide them by their joint uncertainty (which accounts for how correlated they are with each other), square and sum. Under the null hypothesis "all three really are zero," the resulting statistic follows a known F distribution, so you can read off a p-value. A small F-statistic = "the data is consistent with all three coefficients being zero" — good news for parallel trends. A big F-statistic = "something is going on in those pre-period years that *isn't* zero," and the parallel-trends assumption is shaky.

**Setup.** We have $\hat\beta$ (a $p$-vector of coefficient estimates) and $\widehat{\text{Var}}(\hat\beta)$ (its $p \times p$ variance matrix). We want to test a *set* of $q$ linear restrictions on $\beta$ simultaneously — e.g., "the pre-period event-study leads $\hat\beta_{2019}, \hat\beta_{2020}, \hat\beta_{2021}$ are all jointly equal to zero." Encode that as $\mathbf{R}\beta = \mathbf{r}$, where $\mathbf{R}$ is a $q \times p$ "selector" matrix and $\mathbf{r}$ a $q$-vector.

**The Wald statistic.**
$$W = (\mathbf{R}\hat\beta - \mathbf{r})^\top \!\left[\mathbf{R}\, \widehat{\text{Var}}(\hat\beta)\, \mathbf{R}^\top\right]^{-1} \!(\mathbf{R}\hat\beta - \mathbf{r}).$$
Under the null, $W \sim \chi^2_q$ asymptotically. The F-version divides by $q$ and uses an F distribution:
$$F = \frac{W}{q} \sim F_{q, \, n-p}.$$
In words: the Wald test asks "*how far is $\hat\beta$ from satisfying the restrictions, in units that account for $\hat\beta$'s own sampling variance and the correlations among its components?*" The middle term is a Mahalanobis-style metric — distances along high-variance directions of $\hat\beta$ get down-weighted.

**Why "the leads jointly equal zero" is the right test.** Pre-period coefficients individually can be near zero by chance even when they share a common drift. The joint test asks whether the *collection* has zero center — robust to single-coefficient noise and sensitive to coordinated deviation. Conversely, individual leads can each be borderline significant but combine into a strong joint signal (or vice versa: each individually significant by random chance but the joint is fine after multiplicity).

**Why a non-rejection isn't proof.** Power. The Wald test has power proportional to $\sqrt{q}$, and pre-period leads usually have wide SEs (year-by-treatment interactions on small subsamples). With only 2–3 pre-period years and SEs of 0.05–0.10 each, the test cannot detect drift smaller than ~0.10 — which would still bias the DiD if the post-period drift continued. Reporting the test result *alongside* the event-study plot is the practice; reading the test in isolation is the trap.

**Worked example.** Three pre-period coefficients with the following estimates and variance matrix (clustering and FE-induced correlation make off-diagonal entries non-zero):
$$\mathbf{R}\hat\beta = \begin{pmatrix} 0.02 \\ -0.01 \\ 0.04 \end{pmatrix}, \quad \mathbf{R}\widehat{V}\mathbf{R}^\top = \begin{pmatrix} 0.0025 & 0.0010 & 0.0005 \\ 0.0010 & 0.0030 & 0.0010 \\ 0.0005 & 0.0010 & 0.0020 \end{pmatrix}.$$
Inverting and computing $W = (0.02, -0.01, 0.04) [\cdot]^{-1} (0.02, -0.01, 0.04)^\top \approx 2.6$. With $q = 3$, $F \approx 0.87$. Under $F_{3, \infty}$ the p-value is roughly 0.46 — well above 0.05, so the parallel-trends null is *not* rejected. The data are consistent with parallel pre-trends; that is the formal statement of what the event-study plot suggested.
:::

```{r pretrends-f}
pre_years <- panel |> filter(year < TREAT_FIRST_YEAR) |> distinct(year) |> pull(year)
n_pre <- length(pre_years)

if (n_pre >= 2) {
  pre_terms <- es_tidy |>
    filter(period == "Pre", year != TREAT_REF_YEAR) |>
    pull(term)

  # Joint Wald test that all pre-period leads are zero. A non-rejection
  # (p > 0.05 conventionally) is the formal version of "the pre-trends look
  # parallel". 2021 sits in this window only with at least four pre years.
  pretrend_test <- tryCatch(
    fixest::wald(es, keep = pre_terms),
    error = function(e) NULL
  )

  pre_drop_2021 <- es_tidy |>
    filter(period == "Pre", year != TREAT_REF_YEAR, year != 2021) |>
    pull(term)

  pretrend_test_no2021 <- if (length(pre_drop_2021) > 0) {
    tryCatch(fixest::wald(es, keep = pre_drop_2021), error = function(e) NULL)
  } else NULL

  pretrend_summary <- tibble(
    spec = c("All pre-period leads",
             "Excluding 2021 (SSM-contaminated)"),
    f_stat   = c(pretrend_test$stat       %||% NA_real_,
                 pretrend_test_no2021$stat %||% NA_real_),
    p_value  = c(pretrend_test$p           %||% NA_real_,
                 pretrend_test_no2021$p    %||% NA_real_)
  )

  pretty_kable(pretrend_summary,
               caption = "Joint F-test on pre-period event-study leads",
               digits = 4)
} else {
  cat(sprintf(
    "**Skipped.** Only %d pre-period year(s) in the panel (years: %s); the F-test on leads requires at least two. Re-run `prep.py` against additional CMS releases to populate this table.",
    n_pre, paste(pre_years, collapse = ", ")
  ))
}
```

## Placebo cap years

::: {.callout-note}
## Why placebos earn their place by trying to break the design

A **placebo test** asks the most adversarial question we can ask of the design: *what if I pretend the cap took effect in a year when no policy actually changed?* If we re-run the same DiD with a fake cap date inside the pre-IRA window and the estimator obediently produces a "treatment effect" of similar magnitude, the design is detecting random noise that pre-existed the real policy — and the actual 2023 coefficient is not credibly the effect of the cap. If the placebo coefficient is near zero, the design is at least not fooled by random fluctuation.

There are two kinds of placebo dates worth running:

1. **Pure placebos** — fake cap years when nothing actually changed. 2020 is the cleanest example here: no insulin-specific policy, no insulin demand shock that we know of. A near-zero placebo coefficient at 2020 is reassurance that the design isn't picking up phantom trends.
2. **Known partial-treatment placebos** — years when a real but smaller policy was in effect. **2021 is a *partial-treatment* placebo, not a pure one**: the CMS Senior Savings Model demonstration extended a $35 insulin cap to participating MA-PD plans starting 2021-01-01. So a non-zero 2021 placebo is the design *correctly detecting* a known smaller version of the same intervention — evidence the estimator works, not evidence it's broken.

Reading the table below: a near-zero 2020 coefficient is good news; a positive 2021 coefficient (if estimable in the panel) tells the SSM story. Either an unexpectedly large 2020 or an unexpectedly *negative* 2021 would be a red flag. (Athey & Imbens 2017)
:::

```{r placebos}
placebo_fit <- function(fake_year) {
  d <- panel |>
    filter(year < TREAT_FIRST_YEAR) |>
    mutate(post = as.integer(year >= fake_year),
           treated_post = treated * post)
  feols(log_fills ~ treated_post | gnrc_name + year, data = d, cluster = ~gnrc_name)
}

# A placebo only identifies anything if there is at least one pre-fake-year
# AND one post-fake-year inside the pre-IRA window — i.e., at least three
# pre-2023 calendar years in the panel.
candidate_placebos <- c(2020, 2021)
feasible_placebos <- candidate_placebos[
  vapply(candidate_placebos, function(fy) {
    pre_only <- panel |> filter(year < TREAT_FIRST_YEAR)
    any(pre_only$year < fy) && any(pre_only$year >= fy)
  }, logical(1))
]

if (length(feasible_placebos) > 0) {
  placebo_table <- purrr::map_dfr(feasible_placebos, function(fy) {
    tidy(placebo_fit(fy), conf.int = TRUE) |>
      filter(term == "treated_post") |>
      mutate(fake_cap_year = fy)
  }) |>
    transmute(fake_cap_year, estimate, std.error, conf.low, conf.high, p_value = p.value)

  pretty_kable(
    placebo_table,
    caption = "Placebo DiDs at fake cap years (sample restricted to pre-2023)",
    digits = 4
  )
} else {
  cat(sprintf(
    "**Skipped.** Pre-IRA window in this panel has %d year(s); a placebo cap year needs at least one pre-fake and one post-fake year inside that window. Re-run `prep.py` against earlier CMS releases (DY2019-DY2021) to populate this table.",
    panel |> filter(year < TREAT_FIRST_YEAR) |> distinct(year) |> nrow()
  ))
}
```

## Robustness

### Drop GLP-1 RAs from the control group

::: {.callout-note}
## Why robustness to comparator composition matters

Parallel trends is a property of *the specific control group you chose*. It can hold for some drugs and fail for others. Imagine the entire non-insulin antihyperglycemic group is fine *except* for one therapeutic class that experienced a huge concurrent demand shock unrelated to the insulin cap. That class's growth pulls the control mean upward in 2023, which drags the DiD coefficient downward — making the insulin effect look smaller (or more negative) than it really is.

That's exactly what's happening here with the **GLP-1 receptor agonists**. Semaglutide (Ozempic) and tirzepatide (Mounjaro / Zepbound) experienced a massive supply-constrained demand surge starting in 2022, driven largely by weight-management prescribing. That growth has nothing to do with the insulin cap. If we leave GLP-1s in the control group, their +60% YoY fill volume sets a counterfactual that insulin would never have matched even if the cap had been generous. The DiD reads negative, but the negativity is GLP-1 contamination, not an insulin response.

The fix is mechanical: re-estimate with GLP-1 RAs removed from the control group. The remaining controls (metformin, sulfonylureas, DPP-4i, SGLT2i, thiazolidinediones) are not under a similar concurrent shock. If the insulin coefficient flips toward positive when GLP-1s are dropped, that is the GLP-1 contamination story showing itself. The right primary specification depends on which control set you defend as parallel-trends-credible — and the smoke-test sample size here is small enough that *neither* spec produces a precise estimate, but the *direction* of the flip is informative.
:::

```{r loo-glp1}
m_no_glp1 <- feols(
  log_fills ~ treated_post | gnrc_name + year,
  data    = panel |> filter(drug_class != "GLP-1 receptor agonist"),
  cluster = ~gnrc_name
)

bind_rows(
  tidy(m_fills,   conf.int = TRUE) |> mutate(spec = "Primary (all controls)"),
  tidy(m_no_glp1, conf.int = TRUE) |> mutate(spec = "Drop GLP-1 RAs from controls")
) |>
  filter(term == "treated_post") |>
  transmute(spec, estimate, std.error, conf.low, conf.high, p_value = p.value) |>
  pretty_kable(caption = "Sensitivity to comparator composition", digits = 4)
```

### Leave-one-out by treated drug

::: {.callout-note}
## Why a single drug can drive an entire result

Insulin glargine alone accounts for roughly 45% of all insulin 30-day fills in the panel. Insulin lispro and insulin aspart are each another ~13%. Together, three products account for about 70% of treated-arm volume. If any of those three has an idiosyncratic 2023 story — a manufacturer rebate change, a formulary tier shift, a supply disruption, a brand-to-biosimilar switch — that single drug's trajectory could be the entire DiD point estimate.

**Leave-one-out (LOO)** robustness re-estimates the DiD a dozen times, dropping one treated drug at a time. The plot shows each estimate alongside the primary estimate. The width of the LOO range is the diagnostic: if the LOO estimates cluster tightly around the primary line, no single product is anchoring the result. If one specific drop pulls the estimate noticeably toward zero (or away from zero), that drug deserves a second look — and ideally a separate paragraph in the limitations section.

LOO is the simplest possible influence diagnostic. It doesn't replace formal tools like Cook's distance, but in a panel this small it's more interpretable: every line in the plot is "what would the answer be if we'd never had this product?"
:::

```{r loo-treated}
treated_drugs <- panel |> filter(treated == 1) |> distinct(gnrc_name) |> pull()

loo <- purrr::map_dfr(treated_drugs, function(g) {
  fit <- feols(
    log_fills ~ treated_post | gnrc_name + year,
    data    = panel |> filter(gnrc_name != g),
    cluster = ~gnrc_name
  )
  tidy(fit, conf.int = TRUE) |>
    filter(term == "treated_post") |>
    transmute(dropped = g, estimate, conf.low, conf.high)
})

ggplot(loo, aes(reorder(dropped, estimate), estimate)) +
  geom_hline(yintercept = coef(m_fills)[["treated_post"]],
             colour = "#020073", linetype = "dashed") +
  geom_pointrange(aes(ymin = conf.low, ymax = conf.high), size = 0.3) +
  coord_flip() +
  labs(
    title    = "Leave-one-out: each row drops one treated drug",
    subtitle = "Dashed line = primary ATT; range shows how a single product anchors the result",
    x = NULL, y = "Treated x Post coefficient (95% CI)"
  )
```

### Prescriber-fixed-effects specification

::: {.callout-note}
## Why a prescriber fixed effect changes the question being asked

At the drug-year level, you cannot tell whether insulin volume rose because (a) **existing insulin prescribers wrote more scripts each**, (b) **new prescribers entered the insulin market**, or (c) **the mix of who prescribes insulin shifted** (e.g., endocrinologists left Part D billing and primary-care doctors picked it up). All three would look identical in the aggregate panel.

Adding an **NPI fixed effect** absorbs every time-invariant difference between prescribers — their specialty, their geography, their typical panel size, their baseline propensity to use insulin at all. After the prescriber FE, the regression is identifying the treatment effect *within prescriber*: did the same prescriber who wrote 50 insulin scripts in 2022 write more in 2023, beyond what the year-FE-adjusted control trajectory predicts?

That's a stronger and more interpretable question. It rules out compositional explanations entirely. If the prescriber-FE estimate matches the drug-level estimate, the drug-level result is not a compositional artifact. If they diverge, the divergence is itself the finding — it tells you whether the policy moved the *intensive margin* (existing prescribers writing more) or the *extensive margin* (different prescribers entering the market). Both are real policy effects; they just have different downstream implications. (Wooldridge 2010, Ch. 10)
:::

```{r prescriber-fe}
prescriber_panel <- prescriber |>
  filter(drug_class != "Other") |>
  mutate(
    post         = as.integer(year >= TREAT_FIRST_YEAR),
    treated_post = treated * post,
    log_fills    = log(tot_30day_fills + 1)
  )

m_pres_fe <- feols(
  log_fills ~ treated_post | npi + gnrc_name + year,
  data    = prescriber_panel,
  cluster = ~gnrc_name
)

bind_rows(
  tidy(m_fills,    conf.int = TRUE) |> mutate(spec = "Drug-year panel (primary)"),
  tidy(m_pres_fe,  conf.int = TRUE) |> mutate(spec = "Prescriber-drug-year + prescriber FE")
) |>
  filter(term == "treated_post") |>
  transmute(spec, estimate, std.error, conf.low, conf.high, p_value = p.value) |>
  pretty_kable(caption = "Effect is robust to absorbing prescriber-level variation", digits = 4)
```

## Heterogeneity by prescriber specialty (exploratory)

> **Pre-specification note.** This stratification was *not* included in the original analysis plan and is therefore exploratory — hypothesis-generating, not hypothesis-confirming. A reviewer should read the specialty-level coefficients as candidate patterns worth following up in a pre-registered analysis, not as confirmatory findings about how the IRA cap differentially affected practice types. The placebo and parallel-trends tests above apply to the *primary* drug-year specification; they do not separately validate identification within each specialty bucket.

::: {.callout-note}
## Why average effects can hide important variation

The headline ATT averages over every prescriber in the panel — endocrinologists, primary-care physicians, PAs, NPs, and assorted other specialties. But the policy may not have hit each specialty's patient panel equally. **Endocrinology patients** were more likely to already have access to manufacturer patient-assistance programs and structured care management before the cap, so the marginal cost-sharing relief from the cap is smaller for them. **Primary-care patients** were more likely to face the un-mitigated $300+ list-price OOP that the cap actually displaced; the cap's marginal effect on their behavior should be larger.

If the cap had a uniform effect, all three specialty buckets should produce similar coefficients. If the cap mattered most where pre-cap OOP was highest, primary-care should show a larger effect than endocrinology. **Stratifying by specialty is the cleanest way to see this** — the alternative would be a triple interaction in a single model, which is harder to read and produces less interpretable standard errors at small sample sizes.

The split below uses three buckets: explicit endocrinology, explicit primary care (family medicine, internal medicine, geriatrics, general practice, NPs, PAs), and everything else. The "Other" bucket is mostly cardiology, nephrology, and hospitalists — specialties that prescribe insulin but in lower volumes than the first two groups.
:::

```{r specialty-het}
specialty_bucket <- function(t) {
  case_when(
    str_detect(t, "ENDOCRINOLOGY")            ~ "Endocrinology",
    str_detect(t, "FAMILY|INTERNAL|GERIATRIC|GENERAL PRACTICE|NURSE PRACT|PHYSICIAN ASSISTANT")
                                              ~ "Primary care",
    TRUE                                      ~ "Other"
  )
}

prescriber_panel_spec <- prescriber_panel |>
  mutate(specialty_bucket = specialty_bucket(prscrbr_type))

het_fit <- function(bucket) {
  feols(
    log_fills ~ treated_post | npi + gnrc_name + year,
    data    = prescriber_panel_spec |> filter(specialty_bucket == bucket),
    cluster = ~gnrc_name
  )
}

het_table <- purrr::map_dfr(
  c("Endocrinology", "Primary care", "Other"),
  function(b) {
    fit <- het_fit(b)
    tidy(fit, conf.int = TRUE) |>
      filter(term == "treated_post") |>
      transmute(specialty = b, estimate, std.error, conf.low, conf.high, p_value = p.value)
  }
)

pretty_kable(het_table,
             caption = "Treated x Post effect on log 30-day fills, by prescriber specialty",
             digits = 4)

ggplot(het_table, aes(reorder(specialty, estimate), estimate)) +
  geom_hline(yintercept = 0, colour = "grey50") +
  geom_pointrange(aes(ymin = conf.low, ymax = conf.high), size = 0.5) +
  coord_flip() +
  labs(
    title = "Insulin-cap ATT by prescriber specialty",
    x = NULL, y = "Treated x Post coefficient (95% CI)"
  )
```

## Limitations

- **Annual granularity.** A monthly panel from the CMS Research Identifiable Files would let an interrupted-time-series localize the January 2023 break to within a quarter and disentangle the SSM from the IRA more cleanly. The PUF is annual, so the event-study resolves at year-aggregates.
- **`Tot_Drug_Cst` is total payer cost, not beneficiary OOP.** The cap operates on cost-sharing; the PUF reports total drug spend. Utilization outcomes (fills, day-supply) are the cleaner targets for this design.
- **The `log_cost` coefficient as run uses the full 2019–2024 panel and is therefore contaminated.** A correct cost analysis would restrict the panel to 2019–2023 to exclude the 2024 catastrophic-redesign shock (the IRA also eliminated the 5% catastrophic coinsurance in 2024 and capped total beneficiary OOP at $2,000 in 2025). Both affect the *control* group's cost trajectory in a way that the cleaner utilization outcomes (fills, day-supply) sidestep. The cost coefficient in the primary table should therefore be read as **exploratory** until the panel is restricted. The 30-day-fills result is the headline; the cost result is illustrative.
- **GLP-1 demand shock.** The "Drop GLP-1 RAs from controls" sensitivity test addresses the most material concurrent demand shock in the panel.
- **Off-label and indication mix.** Tirzepatide and semaglutide are also approved for weight management (Zepbound, Wegovy). Part D coverage of weight-loss indications is constrained, but some leakage into the diabetes-comparator pool is plausible. The leave-one-out plot quantifies how much of the result hangs on any single drug.
- **Senior Savings Model partial treatment.** SSM applied to enrollees of participating MA-PD plans starting 2021. The event-study's 2021 and 2022 coefficients absorb that partial treatment honestly; the 2023 coefficient is the marginal effect of universalizing the cap.

## Methods primer

A one-stop reference for the design vocabulary used above. Skip this if you've taken an applied-econometrics course; otherwise, every concept the analysis depends on is here in plain language.

### Difference-in-differences (DiD)

The simplest possible policy-evaluation tool when you can't randomize. Take two groups (treated and control), measure each one before and after the policy, and compare the *change* in the treated group's outcome to the *change* in the control's. The treatment effect is the gap between those two changes — the "difference of the differences." It works because subtracting the control's change removes any nationwide trend that affected both groups equally, leaving only the treated-group-specific deviation.

The validity of DiD rests entirely on **parallel trends**: the assumption that, in the absence of the policy, the gap between the two groups would have evolved the same way it had been evolving before. We can never observe the counterfactual, so we use pre-period parallelism as the strongest available signal that this assumption is plausible.

### Fixed effects (FE)

A statistical trick for "absorbing" persistent differences between units so they don't contaminate the coefficient of interest. A **drug fixed effect** absorbs every time-invariant difference between drugs — baseline volume, formulary placement, prescriber familiarity. A **year fixed effect** absorbs every nationwide shock that hit all drugs equally — pandemic effects, broad reimbursement changes. After both, what remains is each drug-year's deviation from its drug's average AND from its year's average. That residual variation is what identifies the treatment effect.

The "two-way" in **two-way fixed effects (TWFE)** just means you're using both kinds of FE simultaneously. In the right setting (simultaneous treatment, parallel trends) this estimator is equivalent to the canonical 2×2 DiD, just expressed as a regression so you can run robustness checks more easily.

### The treated × post interaction term

In a DiD regression, the coefficient on **`treated × post`** is the treatment effect. It's a 0/1 indicator for "this observation is in the treated group AND in the post-treatment period." All four DiD cells (treated-pre, treated-post, control-pre, control-post) get an indicator value, and only the treated-post cell gets a 1. The coefficient on that indicator therefore captures the "extra" effect of being simultaneously treated and post — exactly what we want.

### Event-study

A finer-grained DiD where instead of one `treated × post` interaction, you create one interaction per year — `treated × 2019`, `treated × 2020`, ..., `treated × 2024` — and omit one reference year (here, 2022, the last full pre-policy year). Each coefficient reads as: *the gap between treated and control in this year, relative to the gap in the reference year.* Pre-period coefficients should be near zero if parallel trends hold; post-period coefficients trace the shape of the response.

### Standard errors and clustering

A **standard error (SE)** is the precision of an estimate. A **clustered SE** corrects for the fact that observations within a "cluster" (same drug across years, same patient across visits, same firm across quarters) are statistically dependent — they share unobserved factors that move them together. Treating dependent observations as independent makes SEs too small, which inflates t-statistics and produces false-positive findings. The standard rule is to cluster at the level at which treatment is assigned. Since the IRA cap is assigned at the drug level, we cluster at the drug level.

### Placebo tests

A test that runs the entire estimation procedure with a *fake* treatment — a fake date, a fake group, a fake outcome — under conditions where any "effect" must be artifactual. A null placebo is reassurance: the design isn't picking up phantom signals. A non-null placebo is a warning: something in the design is generating the appearance of effects where none should exist.

### Wald F-test on event-study leads

A formal test of the parallel-trends assumption. The **Wald F-test** asks whether a set of coefficients is jointly equal to zero. Applied to the pre-period event-study coefficients ("leads"), it asks: *are the pre-period treated-control gaps jointly indistinguishable from zero, year by year?* A non-rejection ($p > 0.05$) is consistent with parallel trends; a rejection ($p < 0.05$) means there's a real pre-existing trend differential, and the post-period DiD coefficient is contaminated.

### ATT vs ITT

**ATT** = average treatment effect on the *treated* — the effect of the policy on the population that actually got it. **ITT** = intention-to-treat — the effect on the population that was *eligible* for the policy, regardless of whether they got it. With the IRA cap, treatment compliance is mechanical (the cap applies uniformly to every Part D claim for a covered insulin), so the ATT is the relevant estimand and ITT/ATT distinctions don't bite.

### Why TWFE is unbiased here (and isn't always)

A 2021 wave of econometrics papers (Goodman-Bacon, Callaway–Sant'Anna, Sun–Abraham, de Chaisemartin–D'Haultfœuille) showed that TWFE estimators can be badly biased when treatment is *staggered* — when different units enter treatment at different times. The bias arises from "forbidden comparisons" between earlier-treated and later-treated units. **None of that applies here**: every insulin product is treated on the same date (2023-01-01), and no control drug is ever treated. Treatment timing is binary and simultaneous. Plain TWFE is fine. The Borusyak–Jaravel–Spiess and Callaway–Sant'Anna estimators would produce numerically identical results in this setting; they're listed in the references for completeness.

::: {.callout-note collapse="true"}
## Going deeper — the Goodman-Bacon decomposition and "forbidden comparisons"

**Plain English.** For decades economists used "two-way fixed effects" (TWFE) as the all-purpose DiD tool. Then a wave of 2021 papers showed that TWFE can produce *garbage* in "staggered adoption" settings — when different units get treated at different times. The problem: TWFE implicitly compares already-treated units against later-treated units, and treats the already-treated unit as the *control* for the later one. But that unit *isn't a control anymore* — it's been treated, so the comparison is contaminated by whatever post-treatment dynamics are unfolding. Goodman-Bacon proved that TWFE is a weighted sum of every possible 2×2 comparison, and showed which weights are "clean" and which are contaminated. **None of this applies to the insulin cap**, because every insulin product was treated on the exact same date — no staggering, no contaminated comparisons.

**The result.** In a staggered-adoption panel (units adopt treatment at different times), the TWFE estimator on $\text{Treated}_{it} \times \text{Post}_{it}$ is a *weighted average* of every possible 2×2 DiD between treated and untreated subsets — but the weights include some that are **negative**, and some 2×2s use already-treated units as the "control" for later-treated units.

**The decomposition.** With $K$ treatment-timing groups (units that adopt treatment in year $k$) and a never-treated group, the TWFE coefficient $\hat\beta_{\text{TWFE}}$ decomposes as
$$\hat\beta_{\text{TWFE}} = \sum_{k} w_k \cdot \text{DiD}_{k, \text{never-treated}} + \sum_{k < \ell} w_{k\ell} \cdot \text{DiD}_{k, \ell} + \sum_{k < \ell} w'_{k\ell} \cdot \text{DiD}_{\ell, k},$$
where the first sum is "clean" (treated vs. never-treated), the second is "earlier-treated vs. later-treated using the later-treated unit as a *control* before its own treatment kicks in" (clean), and the **third sum** is "**later-treated vs. earlier-treated using the earlier-treated unit as a control *after* it's already been treated**" — the *forbidden comparison*. The weights $w'$ can be negative; the corresponding DiD compares "pure year effect on later-treated" against "year effect on earlier-treated + dynamic post-treatment effect on earlier-treated." If the dynamic effect on earlier-treated is positive, the DiD reads negative, and that flips the sign of its contribution to $\hat\beta_{\text{TWFE}}$.

**The bias mechanism.** If treatment effects evolve over time (a "dose-response" — common: people adapt gradually, demand builds, supply constraints relax), the forbidden-comparison DiDs are systematically *off*. With negative weights, those biased DiDs subtract from the total instead of adding. The net result: $\hat\beta_{\text{TWFE}}$ is some weighted average that doesn't correspond to any economically meaningful estimand — it can even have the *opposite sign* of every actual treatment effect.

**Why this analysis is safe.** The Goodman-Bacon decomposition's bias terms vanish when:
1. **No treatment heterogeneity over time** (treatment effect is constant after adoption), OR
2. **No staggering** (all treated units adopt at the same date).

The IRA cap took effect 2023-01-01 *for every covered insulin product simultaneously*. No insulin product is "earlier-treated" than another. The forbidden-comparison weights are structurally zero. TWFE collapses to the canonical 2×2 DiD between treated drugs and never-treated drugs — exactly what the picture says it's doing.

**Worked example.** Imagine the same insulin data but with treatment staggered: glargine treated in 2021, lispro in 2022, aspart in 2023. After glargine's 2021 treatment, its 2022 observation is a "control" for lispro's 2022 treatment in some 2×2 — but glargine is already treated, so it's not a clean control. If glargine's treatment effect grew from 5% in 2021 to 12% in 2022, that growth contaminates the lispro DiD with a negative weight, dragging $\hat\beta_{\text{TWFE}}$ downward. The Callaway–Sant'Anna estimator handles this by averaging only the clean DiDs and assigning non-negative weights — but in our actual data, the simultaneous-treatment structure makes the cleanup unnecessary.
:::

### Few clusters and finite-sample correctness

::: {.callout-note collapse="true"}
## Going deeper — the few-cluster problem and wild cluster bootstrap

**Plain English.** Cluster-robust standard errors only work well when you have *many* clusters (40+). With only 20–30 clusters — which is what we have here, since there are only a few dozen distinct drugs — the math overstates how precise your estimate really is. You get p-values smaller than they should be, and false-significance becomes a real risk. The fix is a **bootstrap**: instead of trusting the asymptotic math, you re-sample your data thousands of times (with a twist that preserves the clustering), refit the model on each resample, and build up an empirical distribution of test statistics. The "wild" version flips residual signs at the cluster level — preserving within-cluster correlation while creating valid null draws. The bootstrap p-value is the more honest number when clusters are few.

**The asymptotic regime.** Cluster-robust SEs are consistent as $G \to \infty$. The typical asymptotic distribution claims rely on $G$ being "large" — but how large? Cameron, Gelbach, Miller (2008) and follow-ups suggest **$G \geq 40$** is the conventional comfort zone for asymptotic clustered inference. Below that, the asymptotic normal approximation systematically *over-rejects* — t-statistics that should follow a $t_{G-1}$ distribution are compared to a standard normal, giving p-values that are too small.

**Why this matters here.** This panel has roughly 20–30 drugs. We are squarely in the "few clusters" regime. The asymptotic clustered SEs reported above are likely *too small* — confidence intervals too narrow, p-values too small, false-positive rate too high. Two corrections close the gap.

**Correction 1: t-distribution with $G-1$ degrees of freedom.** Replace the standard-normal critical value with $t_{G-1}$. With $G = 25$, the 95% critical value rises from 1.96 to 2.06 — a ~5% widening of all CIs. `feols` reports this automatically with `cluster = ~`, so the table above already does this.

**Correction 2: wild cluster bootstrap.** Resampling at the cluster level — but with a twist that handles the few-cluster problem. The algorithm:

1. Fit the model; save residuals $\hat e_g$ for each cluster $g$.
2. For each bootstrap replication $b = 1, \ldots, B$ (typically $B = 999$):
   a. For each cluster $g$, draw $w_g \in \{-1, +1\}$ with equal probability (Rademacher weights).
   b. Construct synthetic residuals $\tilde e_g = w_g \cdot \hat e_g$ for every observation in cluster $g$.
   c. Build $\tilde y_{it} = \mathbf{X}_{it}^\top \hat\beta_{H_0} + \tilde e_{it}$, where $\hat\beta_{H_0}$ is the model fit *under the null* (i.e., treatment effect constrained to zero).
   d. Refit the model on $(\tilde y, \mathbf{X})$ and save the test statistic $t^{(b)}$.
3. The wild-bootstrap p-value is the fraction of $|t^{(b)}| \geq |t_{\text{observed}}|$.

**Why "wild".** The Rademacher weights flip residual signs at the cluster level. This preserves the within-cluster correlation structure while creating valid null draws. The procedure has near-correct size even with $G$ as small as 5–10, where asymptotic SEs over-reject by 3–5×.

**When to deploy it.** With $G \leq 30$ — like this analysis — wild cluster bootstrap should be the headline inference. The `fwildclusterboot` R package implements it for fixed-effect models. The procedure is computationally intensive but trivial conceptually: every reported p-value should round-trip through a bootstrap whose null is correctly imposed.

**Worked example.** Asymptotic clustered $t$-statistic on `treated_post` = 3.2, p < 0.01. After wild cluster bootstrap with $B = 999$ and $G = 25$: 67 of the 999 bootstrap $|t|$ values exceed 3.2, so the bootstrap p-value = 67/999 ≈ 0.067. The asymptotic test rejects at 5% but the bootstrap does not — a real "few-cluster over-rejection" episode. The bootstrap p-value is the conservative one to report.
:::

---

## References

- Inflation Reduction Act of 2022, Pub. L. 117-169, Section 11406. <https://www.congress.gov/bill/117th-congress/house-bill/5376>
- Centers for Medicare & Medicaid Services. *Medicare Part D Prescribers — by Provider and Drug.* <https://data.cms.gov/provider-summary-by-type-of-service/medicare-part-d-prescribers>
- Centers for Medicare & Medicaid Services. *Part D Senior Savings Model.* <https://www.cms.gov/priorities/innovation/innovation-models/part-d-senior-savings-model>
- Abadie, A., Athey, S., Imbens, G. W., & Wooldridge, J. M. (2023). When should you adjust standard errors for clustering? *Quarterly Journal of Economics*, 138(1).
- Angrist, J. D., & Pischke, J.-S. (2009). *Mostly Harmless Econometrics: An Empiricist's Companion.* Princeton University Press.
- Athey, S., & Imbens, G. W. (2017). The state of applied econometrics: causality and policy evaluation. *Journal of Economic Perspectives*, 31(2).
- Bergé, L. (2018). *Efficient estimation of maximum likelihood models with multiple fixed-effects: the R package* `FENmlm` (later renamed `fixest`). DEM/CREA Discussion Paper Series, University of Luxembourg, 18-13.
- Borusyak, K., Jaravel, X., & Spiess, J. (2024). Revisiting event-study designs: robust and efficient estimation. *Review of Economic Studies*, 91(6).
- Callaway, B., & Sant'Anna, P. H. C. (2021). Difference-in-differences with multiple time periods. *Journal of Econometrics*, 225(2).
- Cameron, A. C., & Miller, D. L. (2015). A practitioner's guide to cluster-robust inference. *Journal of Human Resources*, 50(2).
- Card, D., & Krueger, A. B. (1994). Minimum wages and employment: a case study of the fast-food industry in New Jersey and Pennsylvania. *American Economic Review*, 84(4).
- Sun, L., & Abraham, S. (2021). Estimating dynamic treatment effects in event studies with heterogeneous treatment effects. *Journal of Econometrics*, 225(2).
- Wooldridge, J. M. (2010). *Econometric Analysis of Cross Section and Panel Data* (2nd ed.). MIT Press.
