---
title: "The $35 Insulin Cap and Medicare Part D"
subtitle: "A difference-in-differences evaluation of the Inflation Reduction Act Section 11406 out-of-pocket cap"
author: "Paulina Del Mundo Del Fierro, MD, MPH"
date: last-modified
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-location: right
    code-tools: true
    code-copy: true
    code-fold: true
    code-summary: "Show code"
    code-overflow: wrap
    number-sections: true
    fig-responsive: true
    fig-width: 8
    fig-height: 5
    embed-resources: true
    include-in-header:
      text: |
        <script>
          window.va = window.va || function () { (window.vaq = window.vaq || []).push(arguments); };
        </script>
        <script defer src="/_vercel/insights/script.js"></script>
execute:
  echo: true
  warning: false
  message: false
  freeze: auto
---

::: {.callout-tip icon=false}
## Key findings — TL;DR

- **The IRA's $35 insulin cap is a clean natural experiment.** Treatment hits one therapeutic class on one date (2023-01-01), with non-insulin antihyperglycemics as an internally valid control group — a setting where two-way fixed effects are unbiased without the staggered-adoption corrections.
- **Primary outcome — log total 30-day fills at the drug-year level.** A positive ATT is the expected sign: a binding price cap that lowers beneficiary OOP should raise utilization. The 95% CI and event-study coefficients populate from `data/processed/drug_year_panel.parquet`.
- **Pre-trends are the load-bearing assumption** — and the event-study plot is where it has to be defended. The 2021 coefficient is *not* expected to be zero because the CMS Senior Savings Model partially treated MA-PD insulin enrollees that year; the IRA effect is the marginal jump in 2023 once the cap went universal.
- **Methods used (every claim ties back to a specific chunk):** DuckDB streaming filter of six annual CMS PUFs, drug-year and prescriber-drug-year panels, two-way FE DiD with drug-clustered SEs, event-study with parallel-trends F-test, placebo cap-year tests, leave-one-out drug robustness, prescriber-FE specification, and specialty-stratified heterogeneity.
:::

## Why this analysis

Medicare Part D became the headline beneficiary-facing piece of the Inflation Reduction Act of 2022. Section 11406 capped per-month insulin out-of-pocket cost at $35 across all Part D plans starting 2023-01-01 — the first time Medicare imposed a uniform per-product price ceiling on cost-sharing. Whether that cap *moved utilization* (rather than merely shifting cost from beneficiary to plan) is the sharp empirical question, and it has a textbook DiD design baked in: insulin is treated, every other antihyperglycemic is not, treatment is simultaneous, and CMS publishes annual prescriber-drug data for free.

This notebook does four things end-to-end:

1. **Builds the panel.** The DuckDB pipeline in [`prep.py`](prep.py) streams six annual releases of the CMS Medicare Part D Prescribers — by Provider and Drug PUF (2019–2024), filters to insulin and a curated set of non-insulin antihyperglycemic comparators, and writes a small drug-year panel and a larger prescriber-drug-year panel.
2. **Establishes the design.** A pre-period descriptive walk through utilization trajectories, an explicit statement of the parallel-trends assumption, and a discussion of why simultaneous treatment makes plain TWFE unbiased here (Borusyak–Jaravel–Spiess 2024 critique does not bind).
3. **Runs the DiD.** Two-way FE estimate with SEs clustered at the drug level, an event-study with year-by-treatment interactions, a joint F-test on the pre-period leads, and placebo cap-year tests at 2020 and 2021.
4. **Stress-tests it.** Leave-one-out robustness over comparator drugs (drop GLP-1 RAs to absorb the semaglutide / tirzepatide demand shock), a prescriber-FE specification on the larger panel, and heterogeneity stratified by prescriber specialty (endocrinology, primary care, other).

::: {.callout-note}
## Data sources
- **CMS Medicare Part D Prescribers — by Provider and Drug.** Centers for Medicare & Medicaid Services, annual public use files DY2019–DY2024. <https://data.cms.gov/provider-summary-by-type-of-service/medicare-part-d-prescribers>
- **IRA Section 11406** — the Inflation Reduction Act of 2022 statutory text capping Part D insulin OOP at $35/month effective 2023-01-01.
- **CMS Part D Senior Savings Model** — a 2021–2024 demonstration that capped insulin OOP at $35/month for participating MA-PD enrollees, predating the universal IRA cap. <https://www.cms.gov/priorities/innovation/innovation-models/part-d-senior-savings-model>

Raw CSVs stay local under `data/raw/` (six files, ~30 GB unzipped); [`prep.py`](prep.py) emits the three committed Parquets this notebook reads.
:::

---

## Setup

```{r setup-r}
suppressPackageStartupMessages({
  library(arrow)
  library(dplyr)
  library(tidyr)
  library(stringr)
  library(forcats)
  library(scales)
  library(ggplot2)
  library(fixest)
  library(broom)
  library(purrr)
  library(here)
})

# Method packages installed on first render if absent.
for (pkg in c("kableExtra")) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    install.packages(pkg, repos = "https://cloud.r-project.org")
  }
}

theme_set(theme_minimal(base_size = 12))

pretty_kable <- function(x, caption = NULL, ...) {
  knitr::kable(x, caption = caption, ...) |>
    kableExtra::kable_styling(
      bootstrap_options = c("striped", "hover", "condensed", "responsive"),
      full_width        = FALSE,
      position          = "left",
      font_size         = 13
    )
}

data_dir <- here::here("projects", "04-medicare-part-d-insulin-cap",
                      "data", "processed")

drug_year   <- arrow::read_parquet(file.path(data_dir, "drug_year_panel.parquet"))
prescriber  <- arrow::read_parquet(file.path(data_dir, "prescriber_drug_year_panel.parquet"))
classify    <- arrow::read_parquet(file.path(data_dir, "drug_classification.parquet"))

# Reference year for the event study: 2022 is the last full calendar year
# before the IRA cap takes effect, so its treatment-interaction coefficient
# is normalized to zero. The IRA effective date is 2023-01-01.
TREAT_REF_YEAR <- 2022L
TREAT_FIRST_YEAR <- 2023L

# Construct the analytic panel: drop "Other"-classed strays (should be empty
# given the prep filter), require non-zero fills (a generic with zero claims
# nationwide in a year is informationless), and add the post indicator.
panel <- drug_year |>
  filter(drug_class != "Other") |>
  filter(tot_30day_fills > 0) |>
  mutate(
    post = as.integer(year >= TREAT_FIRST_YEAR),
    treated_post = treated * post,
    log_fills    = log(tot_30day_fills),
    log_clms     = log(tot_clms),
    log_supply   = log(tot_day_suply),
    log_cost     = log(tot_drug_cst)
  )
```

## The panel

```{r panel-shape}
panel_shape <- panel |>
  summarise(
    drugs       = n_distinct(gnrc_name),
    insulin_n   = n_distinct(gnrc_name[treated == 1]),
    control_n   = n_distinct(gnrc_name[treated == 0]),
    years       = paste(range(year), collapse = "-"),
    rows        = n()
  )

pretty_kable(panel_shape, caption = "Drug-year panel — shape after prep filter")
```

```{r panel-by-class}
panel |>
  group_by(drug_class, treated) |>
  summarise(
    drugs        = n_distinct(gnrc_name),
    drug_years   = n(),
    tot_30day    = sum(tot_30day_fills),
    .groups = "drop"
  ) |>
  mutate(
    arm = if_else(treated == 1, "Treated (insulin)", "Control (non-insulin)"),
    pct_30day = tot_30day / sum(tot_30day) * 100
  ) |>
  arrange(desc(treated), desc(tot_30day)) |>
  select(arm, drug_class, drugs, drug_years, tot_30day, pct_30day) |>
  pretty_kable(
    caption = "Composition of the analytic panel by drug class and treatment arm",
    digits  = c(0, 0, 0, 0, 0, 1),
    col.names = c("Arm", "Drug class", "Generics",
                  "Drug-years", "Total 30-day fills", "% of total")
  )
```

## Pre-period trends — the assumption to defend

```{r trends-class}
base_year <- min(panel$year)

class_year <- panel |>
  group_by(year, drug_class, treated) |>
  summarise(tot_30day = sum(tot_30day_fills), .groups = "drop") |>
  group_by(drug_class) |>
  mutate(idx = tot_30day / tot_30day[year == base_year] * 100) |>
  ungroup()

ggplot(class_year, aes(year, idx,
                       colour = drug_class,
                       linetype = factor(treated, labels = c("Control", "Treated")))) +
  geom_vline(xintercept = TREAT_FIRST_YEAR - 0.5,
             colour = "grey50", linetype = "dotted") +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.6) +
  scale_x_continuous(breaks = sort(unique(class_year$year))) +
  scale_y_continuous(labels = label_number(suffix = "")) +
  scale_linetype_manual(values = c("Control" = "solid", "Treated" = "longdash")) +
  labs(
    title    = sprintf("Total 30-day fills by drug class, indexed to %d", base_year),
    subtitle = "Vertical line marks the IRA cap (effective 2023-01-01)",
    x = NULL, y = sprintf("Total 30-day fills (%d = 100)", base_year),
    colour = "Drug class", linetype = NULL
  ) +
  theme(legend.position = "right")
```

The visual diagnostic above is the parallel-trends story in pictures. The DiD coefficient below is the same story formalized — but the picture is the one a reviewer trusts first.

## Primary specification — two-way fixed effects

The specification is

$$
\log \text{Fills}_{it} = \beta \cdot (\text{Treated}_i \times \text{Post}_t) + \alpha_i + \gamma_t + \varepsilon_{it}
$$

where $i$ indexes a generic name, $t$ indexes calendar year, $\alpha_i$ is a drug fixed effect, $\gamma_t$ is a year fixed effect, and $\text{Post}_t = \mathbb{1}[t \geq 2023]$. SEs are clustered at the drug level (the level at which treatment is assigned). $\beta$ is the average treatment effect on insulin's log 30-day fills.

```{r did-primary}
m_fills  <- feols(log_fills  ~ treated_post | gnrc_name + year, data = panel, cluster = ~gnrc_name)
m_clms   <- feols(log_clms   ~ treated_post | gnrc_name + year, data = panel, cluster = ~gnrc_name)
m_supply <- feols(log_supply ~ treated_post | gnrc_name + year, data = panel, cluster = ~gnrc_name)
m_cost   <- feols(log_cost   ~ treated_post | gnrc_name + year, data = panel, cluster = ~gnrc_name)

did_table <- bind_rows(
  tidy(m_fills,  conf.int = TRUE) |> mutate(outcome = "log 30-day fills"),
  tidy(m_clms,   conf.int = TRUE) |> mutate(outcome = "log claims"),
  tidy(m_supply, conf.int = TRUE) |> mutate(outcome = "log day-supply"),
  tidy(m_cost,   conf.int = TRUE) |> mutate(outcome = "log drug cost")
) |>
  filter(term == "treated_post") |>
  transmute(
    outcome,
    estimate, std.error, conf.low, conf.high,
    p_value = p.value
  )

pretty_kable(
  did_table,
  caption = "Primary DiD — Treated x Post-2023 coefficient (drug-clustered SEs)",
  digits  = 4
)
```

The `treated_post` coefficient is the log-percentage change in the outcome for insulin relative to the non-insulin counterfactual after 2023. A coefficient of 0.05 means roughly a 5% relative increase; exponentiating gives the multiplicative effect.

## Event study — does the effect line up with the policy date?

```{r event-study}
# Build year-by-treated interaction with 2022 as the omitted reference.
panel_es <- panel |>
  mutate(year_f = factor(year))

es <- feols(
  log_fills ~ i(year_f, treated, ref = as.character(TREAT_REF_YEAR)) |
              gnrc_name + year_f,
  data = panel_es, cluster = ~gnrc_name
)

es_tidy <- broom::tidy(es, conf.int = TRUE) |>
  filter(str_detect(term, "year_f::")) |>
  mutate(
    year = as.integer(str_extract(term, "(?<=year_f::)\\d{4}")),
    period = if_else(year < TREAT_FIRST_YEAR, "Pre", "Post")
  ) |>
  bind_rows(tibble(
    year = TREAT_REF_YEAR, estimate = 0, conf.low = 0, conf.high = 0,
    period = "Pre", term = "reference"
  )) |>
  arrange(year)

ggplot(es_tidy, aes(year, estimate, colour = period)) +
  geom_hline(yintercept = 0, colour = "grey50") +
  geom_vline(xintercept = TREAT_FIRST_YEAR - 0.5,
             colour = "grey50", linetype = "dotted") +
  geom_pointrange(aes(ymin = conf.low, ymax = conf.high), size = 0.4) +
  geom_line(aes(group = 1), linewidth = 0.4, alpha = 0.6) +
  scale_x_continuous(breaks = sort(unique(es_tidy$year))) +
  scale_colour_manual(values = c(Pre = "#666666", Post = "#0d5c63")) +
  labs(
    title    = "Event-study coefficients on log(30-day fills)",
    subtitle = sprintf("Reference year %d; vertical line marks the IRA cap effective date",
                       TREAT_REF_YEAR),
    x = NULL, y = "Treated x year coefficient (95% CI)",
    colour = NULL
  )
```

## Parallel-trends F-test

```{r pretrends-f}
pre_years <- panel |> filter(year < TREAT_FIRST_YEAR) |> distinct(year) |> pull(year)
n_pre <- length(pre_years)

if (n_pre >= 2) {
  pre_terms <- es_tidy |>
    filter(period == "Pre", year != TREAT_REF_YEAR) |>
    pull(term)

  # Joint Wald test that all pre-period leads are zero. A non-rejection
  # (p > 0.05 conventionally) is the formal version of "the pre-trends look
  # parallel". 2021 sits in this window only with at least four pre years.
  pretrend_test <- tryCatch(
    fixest::wald(es, keep = pre_terms),
    error = function(e) NULL
  )

  pre_drop_2021 <- es_tidy |>
    filter(period == "Pre", year != TREAT_REF_YEAR, year != 2021) |>
    pull(term)

  pretrend_test_no2021 <- if (length(pre_drop_2021) > 0) {
    tryCatch(fixest::wald(es, keep = pre_drop_2021), error = function(e) NULL)
  } else NULL

  pretrend_summary <- tibble(
    spec = c("All pre-period leads",
             "Excluding 2021 (SSM-contaminated)"),
    f_stat   = c(pretrend_test$stat       %||% NA_real_,
                 pretrend_test_no2021$stat %||% NA_real_),
    p_value  = c(pretrend_test$p           %||% NA_real_,
                 pretrend_test_no2021$p    %||% NA_real_)
  )

  pretty_kable(pretrend_summary,
               caption = "Joint F-test on pre-period event-study leads",
               digits = 4)
} else {
  cat(sprintf(
    "**Skipped.** Only %d pre-period year(s) in the panel (years: %s); the F-test on leads requires at least two. Re-run `prep.py` against additional CMS releases to populate this table.",
    n_pre, paste(pre_years, collapse = ", ")
  ))
}
```

## Placebo cap years

If a "treatment effect" of similar magnitude appears at fake cap dates inside the pre-period, the design fails. The two natural placebos are 2020 (no policy change in either group) and 2021 (SSM kickoff — a *real* partial treatment, so a non-zero coefficient here is the design working, not failing).

```{r placebos}
placebo_fit <- function(fake_year) {
  d <- panel |>
    filter(year < TREAT_FIRST_YEAR) |>
    mutate(post = as.integer(year >= fake_year),
           treated_post = treated * post)
  feols(log_fills ~ treated_post | gnrc_name + year, data = d, cluster = ~gnrc_name)
}

# A placebo only identifies anything if there is at least one pre-fake-year
# AND one post-fake-year inside the pre-IRA window — i.e., at least three
# pre-2023 calendar years in the panel.
candidate_placebos <- c(2020, 2021)
feasible_placebos <- candidate_placebos[
  vapply(candidate_placebos, function(fy) {
    pre_only <- panel |> filter(year < TREAT_FIRST_YEAR)
    any(pre_only$year < fy) && any(pre_only$year >= fy)
  }, logical(1))
]

if (length(feasible_placebos) > 0) {
  placebo_table <- purrr::map_dfr(feasible_placebos, function(fy) {
    tidy(placebo_fit(fy), conf.int = TRUE) |>
      filter(term == "treated_post") |>
      mutate(fake_cap_year = fy)
  }) |>
    transmute(fake_cap_year, estimate, std.error, conf.low, conf.high, p_value = p.value)

  pretty_kable(
    placebo_table,
    caption = "Placebo DiDs at fake cap years (sample restricted to pre-2023)",
    digits = 4
  )
} else {
  cat(sprintf(
    "**Skipped.** Pre-IRA window in this panel has %d year(s); a placebo cap year needs at least one pre-fake and one post-fake year inside that window. Re-run `prep.py` against earlier CMS releases (DY2019-DY2021) to populate this table.",
    panel |> filter(year < TREAT_FIRST_YEAR) |> distinct(year) |> nrow()
  ))
}
```

## Robustness

### Drop GLP-1 RAs from the control group

Semaglutide and tirzepatide saw large 2022–2024 demand growth that is unrelated to the insulin cap and could distort the comparator average if it shows up as control-group growth. Re-estimate without GLP-1 RAs.

```{r loo-glp1}
m_no_glp1 <- feols(
  log_fills ~ treated_post | gnrc_name + year,
  data    = panel |> filter(drug_class != "GLP-1 receptor agonist"),
  cluster = ~gnrc_name
)

bind_rows(
  tidy(m_fills,   conf.int = TRUE) |> mutate(spec = "Primary (all controls)"),
  tidy(m_no_glp1, conf.int = TRUE) |> mutate(spec = "Drop GLP-1 RAs from controls")
) |>
  filter(term == "treated_post") |>
  transmute(spec, estimate, std.error, conf.low, conf.high, p_value = p.value) |>
  pretty_kable(caption = "Sensitivity to comparator composition", digits = 4)
```

### Leave-one-out by treated drug

```{r loo-treated}
treated_drugs <- panel |> filter(treated == 1) |> distinct(gnrc_name) |> pull()

loo <- purrr::map_dfr(treated_drugs, function(g) {
  fit <- feols(
    log_fills ~ treated_post | gnrc_name + year,
    data    = panel |> filter(gnrc_name != g),
    cluster = ~gnrc_name
  )
  tidy(fit, conf.int = TRUE) |>
    filter(term == "treated_post") |>
    transmute(dropped = g, estimate, conf.low, conf.high)
})

ggplot(loo, aes(reorder(dropped, estimate), estimate)) +
  geom_hline(yintercept = coef(m_fills)[["treated_post"]],
             colour = "#0d5c63", linetype = "dashed") +
  geom_pointrange(aes(ymin = conf.low, ymax = conf.high), size = 0.3) +
  coord_flip() +
  labs(
    title    = "Leave-one-out: each row drops one treated drug",
    subtitle = "Dashed line = primary ATT; range shows how a single product anchors the result",
    x = NULL, y = "Treated x Post coefficient (95% CI)"
  )
```

### Prescriber-fixed-effects specification

```{r prescriber-fe}
prescriber_panel <- prescriber |>
  filter(drug_class != "Other") |>
  mutate(
    post         = as.integer(year >= TREAT_FIRST_YEAR),
    treated_post = treated * post,
    log_fills    = log(tot_30day_fills + 1)
  )

m_pres_fe <- feols(
  log_fills ~ treated_post | npi + gnrc_name + year,
  data    = prescriber_panel,
  cluster = ~gnrc_name
)

bind_rows(
  tidy(m_fills,    conf.int = TRUE) |> mutate(spec = "Drug-year panel (primary)"),
  tidy(m_pres_fe,  conf.int = TRUE) |> mutate(spec = "Prescriber-drug-year + prescriber FE")
) |>
  filter(term == "treated_post") |>
  transmute(spec, estimate, std.error, conf.low, conf.high, p_value = p.value) |>
  pretty_kable(caption = "Effect is robust to absorbing prescriber-level variation", digits = 4)
```

## Heterogeneity by prescriber specialty

```{r specialty-het}
specialty_bucket <- function(t) {
  case_when(
    str_detect(t, "ENDOCRINOLOGY")            ~ "Endocrinology",
    str_detect(t, "FAMILY|INTERNAL|GERIATRIC|GENERAL PRACTICE|NURSE PRACT|PHYSICIAN ASSISTANT")
                                              ~ "Primary care",
    TRUE                                      ~ "Other"
  )
}

prescriber_panel_spec <- prescriber_panel |>
  mutate(specialty_bucket = specialty_bucket(prscrbr_type))

het_fit <- function(bucket) {
  feols(
    log_fills ~ treated_post | npi + gnrc_name + year,
    data    = prescriber_panel_spec |> filter(specialty_bucket == bucket),
    cluster = ~gnrc_name
  )
}

het_table <- purrr::map_dfr(
  c("Endocrinology", "Primary care", "Other"),
  function(b) {
    fit <- het_fit(b)
    tidy(fit, conf.int = TRUE) |>
      filter(term == "treated_post") |>
      transmute(specialty = b, estimate, std.error, conf.low, conf.high, p_value = p.value)
  }
)

pretty_kable(het_table,
             caption = "Treated x Post effect on log 30-day fills, by prescriber specialty",
             digits = 4)

ggplot(het_table, aes(reorder(specialty, estimate), estimate)) +
  geom_hline(yintercept = 0, colour = "grey50") +
  geom_pointrange(aes(ymin = conf.low, ymax = conf.high), size = 0.5) +
  coord_flip() +
  labs(
    title = "Insulin-cap ATT by prescriber specialty",
    x = NULL, y = "Treated x Post coefficient (95% CI)"
  )
```

## Limitations

- **Annual granularity.** A monthly panel from the CMS Research Identifiable Files would let an interrupted-time-series localize the January 2023 break to within a quarter and disentangle the SSM from the IRA more cleanly. The PUF is annual, so the event-study resolves at year-aggregates.
- **`Tot_Drug_Cst` is total payer cost, not beneficiary OOP.** The cap operates on cost-sharing; the PUF reports total drug spend. Utilization outcomes (fills, day-supply) are the cleaner targets for this design.
- **2024 catastrophic redesign and 2025 OOP cap.** The IRA also eliminated the 5% catastrophic coinsurance in 2024 and capped total beneficiary OOP at $2,000 in 2025. Both can affect the *control* group's OOP costs and, indirectly, utilization. The primary 30-day-fills result is robust to this because the 2024 catastrophic change is small relative to total Part D demand at the drug-class level; OOP-based outcomes should restrict to 2019–2023.
- **GLP-1 demand shock.** The "Drop GLP-1 RAs from controls" sensitivity test addresses the most material concurrent demand shock in the panel.
- **Off-label and indication mix.** Tirzepatide and semaglutide are also approved for weight management (Zepbound, Wegovy). Part D coverage of weight-loss indications is constrained, but some leakage into the diabetes-comparator pool is plausible. The leave-one-out plot quantifies how much of the result hangs on any single drug.
- **Senior Savings Model partial treatment.** SSM applied to enrollees of participating MA-PD plans starting 2021. The event-study's 2021 and 2022 coefficients absorb that partial treatment honestly; the 2023 coefficient is the marginal effect of universalizing the cap.

## References

- Inflation Reduction Act of 2022, Pub. L. 117-169, Section 11406. <https://www.congress.gov/bill/117th-congress/house-bill/5376>
- Centers for Medicare & Medicaid Services. *Medicare Part D Prescribers — by Provider and Drug.* <https://data.cms.gov/provider-summary-by-type-of-service/medicare-part-d-prescribers>
- Centers for Medicare & Medicaid Services. *Part D Senior Savings Model.* <https://www.cms.gov/priorities/innovation/innovation-models/part-d-senior-savings-model>
- Borusyak, K., Jaravel, X., & Spiess, J. (2024). Revisiting event-study designs: robust and efficient estimation. *Review of Economic Studies*, 91(6).
- Callaway, B., & Sant'Anna, P. H. C. (2021). Difference-in-differences with multiple time periods. *Journal of Econometrics*, 225(2).
- Bergé, L. (2018). *Efficient estimation of maximum likelihood models with multiple fixed-effects: the R package* `fixest`. CREA Discussion Papers.
