---
title: "Outlier Detection in Medicaid Provider Spending"
subtitle: "Peer-group robust z-scores, isolation forests, and a geographic lens on provider-level Medicaid claims"
author: "Paulina Del Mundo Del Fierro, MD, MPH"
date: last-modified
format:
  html:
    theme: [cosmo, ../_theme.scss]
    highlight-style: github-dark
    toc: true
    toc-depth: 3
    toc-location: right
    code-tools: true
    code-copy: true
    code-fold: true
    code-summary: "Show code"
    code-overflow: wrap
    number-sections: true
    fig-responsive: true
    fig-width: 8
    fig-height: 5
    embed-resources: true
    include-in-header:
      text: |
        <script>
          window.va = window.va || function () { (window.vaq = window.vaq || []).push(arguments); };
        </script>
        <script defer src="/_vercel/insights/script.js"></script>
execute:
  echo: true
  warning: false
  message: false
  freeze: auto
---

::: {.callout-tip icon=false}
## Key findings — TL;DR

- **Heavy tails are the regime, not the exception.** Paid-per-beneficiary spans four orders of magnitude on a log scale. MAD-based robust z's are usable here; mean/SD z's aren't.
- **BH-FDR ≤ 0.01 keeps a much smaller, much higher-confidence flag list than raw z ≥ 5.** The intersection of the two rules is the set a Program Integrity team would prioritize first.
- **The 12-bucket cost atlas is an HCPCS lens, not a disease lens.** Reading it as "Behavioral Health spending geography" is supportable; reading it as "depression cost geography" is not (HCPCS doesn't carry diagnosis).
- **Spatial concentration varies by category.** Dialysis & ESRD and parts of DME concentrate geographically (high Gini); LTSS and E&M diffuse across the country (low Gini) and shouldn't be read as having "hotspots."
- **Methods used (every claim ties back to a specific chunk):** DuckDB streaming aggregate of the 11 GB CSV, robust z + BH-FDR, isolation forest with collinearity diagnostics, ZCTA → county join, within-category decile shading on a leaflet choropleth, three-view category explorer (top HCPCS / Gini / state hotspots).
:::

## Why this analysis

Medicaid is the single largest source of health coverage for low-income Americans — and its claims data, released monthly by HHS, is a rare window into how public dollars flow through the US provider network. The question for any physician-scientist looking at this data is not *"who spends the most"* (large systems dominate by construction) but *"who spends atypically, given their peers?"* That's the signal that drives fraud-and-abuse reviews, network-integrity audits, and honest comparative-effectiveness work alike.

This notebook does three things end-to-end:

1. **Streams the 11 GB Medicaid Provider Spending CSV with DuckDB**, aggregates to the NPI × HCPCS level, and flags peer-group outliers using a median/MAD-based robust z-score — resistant to the heavy tails claims data is famous for.
2. **Joins NPPES, NUCC taxonomy, and US Census ZCTA centroids** so every flagged provider carries an interpretable specialty and a map-able location. Nothing in the analysis requires cloud infrastructure — everything runs on a laptop against committed Parquet artifacts.
3. **Contrasts a deterministic rule (robust z) against an unsupervised model (isolation forest)** on the same cohort, in R and Python respectively. Agreement between the two is the real outlier story; disagreement is where a physician-scientist earns their keep.

::: {.callout-note}
## Data sources
- **Medicaid Provider Spending** — HHS Open Data Portal, accessed 2026-04-21. <https://opendata.hhs.gov/datasets/medicaid-provider-spending/>
- **NPPES Downloadable File** — Centers for Medicare & Medicaid Services. <https://download.cms.gov/nppes/NPI_Files.html>
- **NUCC Health Care Provider Taxonomy** — National Uniform Claim Committee. <https://taxonomy.nucc.org/>
- **ZCTA Gazetteer (2020)** — US Census Bureau. <https://www.census.gov/geographies/reference-files/time-series/geo/gazetteer-files.html>

Raw files stay local (11 GB + 45 GB NPPES); the [`prep.py`](prep.py) script runs one DuckDB pipeline to produce three committed Parquets this notebook reads.
:::

---

## Setup

```{r setup-r}
suppressPackageStartupMessages({
  library(arrow)
  library(dplyr)
  library(ggplot2)
  library(tidyr)
  library(stringr)
  library(forcats)
  library(scales)
  library(sf)
  library(leaflet)
  library(reticulate)
})

reticulate::py_require(c("pandas", "numpy", "scikit-learn", "matplotlib"))

theme_set(theme_minimal(base_size = 12))

# Method packages — installed on first render if absent.
for (pkg in c("kableExtra")) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    install.packages(pkg, repos = "https://cloud.r-project.org")
  }
}

# Lightweight wrapper for hero tables — bootstrap styling, no monospaced
# greyscale dump. Plain knitr::kable() still works elsewhere for inline /
# diagnostic tables where the styled version is overkill.
pretty_kable <- function(x, caption = NULL, ...) {
  knitr::kable(x, caption = caption, ...) |>
    kableExtra::kable_styling(
      bootstrap_options = c("striped", "hover", "condensed", "responsive"),
      full_width        = FALSE,
      position          = "left",
      font_size         = 13
    )
}

data_dir     <- here::here("projects", "02-medicaid-outliers", "data", "processed")
outliers     <- arrow::read_parquet(file.path(data_dir, "outliers.parquet"))
hcpcs_sum    <- arrow::read_parquet(file.path(data_dir, "hcpcs_summary.parquet"))
state_sum    <- arrow::read_parquet(file.path(data_dir, "state_summary.parquet"))
specialty_sum<- arrow::read_parquet(file.path(data_dir, "specialty_summary.parquet"))
taxonomy     <- arrow::read_parquet(file.path(data_dir, "taxonomy.parquet"))

format_money <- function(x) scales::label_dollar(accuracy = 1, scale_cut = scales::cut_short_scale())(x)
```

```{python setup-py}
import pandas as pd
import numpy as np
import matplotlib.pyplot as plt
from sklearn.ensemble import IsolationForest
from sklearn.preprocessing import StandardScaler

plt.rcParams.update({"figure.dpi": 110})
```

## Cohort sketch

```{r cohort-sketch}
summary_tbl <- tibble::tibble(
  hcpcs_n       = nrow(hcpcs_sum),
  providers_n   = outliers |> dplyr::pull(npi) |> dplyr::n_distinct(),
  total_paid    = sum(hcpcs_sum$total_paid, na.rm = TRUE),
  total_bene    = sum(hcpcs_sum$total_bene, na.rm = TRUE),
  flagged_ge_5  = sum(outliers$robust_z >= 5, na.rm = TRUE)
)

knitr::kable(
  summary_tbl |>
    mutate(total_paid = format_money(total_paid),
           across(where(is.numeric), \(x) format(round(x, 2), big.mark = ","))),
  caption = "Cohort sketch from DuckDB prep outputs"
)
```

## Exploratory data analysis — what the prep step actually produced

The DuckDB pipeline emits seven Parquet artifacts. Before any modeling claim, four sanity checks: **are the artifacts the size and shape we expect, are the heavy-tailed distributions actually heavy-tailed, what's the join coverage between provider ZIPs and county FIPS, and how does the robust-z null distribution sit relative to the ≥ 5 flagging threshold?**

```{r eda-artifact-shape}
#| label: eda-artifact-shape

artifact_summary <- function(name, df) {
  tibble::tibble(
    artifact     = name,
    rows         = nrow(df),
    cols         = ncol(df),
    n_complete   = sum(stats::complete.cases(df)),
    pct_complete = 100 * n_complete / pmax(nrow(df), 1)
  )
}

dplyr::bind_rows(
  artifact_summary("hcpcs_summary",    hcpcs_sum),
  artifact_summary("state_summary",    state_sum),
  artifact_summary("specialty_summary",specialty_sum),
  artifact_summary("outliers",         outliers),
  artifact_summary("taxonomy",         taxonomy)
) |>
  dplyr::mutate(
    rows         = format(rows, big.mark = ","),
    cols         = format(cols, big.mark = ","),
    n_complete   = format(n_complete, big.mark = ","),
    pct_complete = sprintf("%.1f%%", pct_complete)
  ) |>
  knitr::kable(
    col.names = c("Parquet artifact", "Rows", "Cols",
                  "Complete-case rows", "% complete-case"),
    caption = "Row × column shape and complete-case rate of each prep artifact. Aggregates (hcpcs / state / specialty summary) are dense by construction; the outliers detail table carries some NA in lat/lon (provider ZIP didn't resolve to a Census centroid)."
  )
```

```{r eda-distribution}
#| label: eda-distribution
#| fig-cap: "Three distributional sanity checks. **Top-left** — paid per beneficiary across all (NPI × HCPCS) rows on a log10 scale; the long upper tail is exactly why MAD-based robust z's outperform a mean/SD z-score here. **Top-right** — beneficiary count per row, capped at 5,000 for visibility. **Bottom-left** — peer-group sizes per HCPCS (the denominators behind every robust z); HCPCS with peer_n < 25 are filtered out in prep. **Bottom-right** — robust-z values (those with a finite z) showing the long right tail; the orange dashed line marks the program-integrity ≥ 5 cutoff."
#| fig-width: 10
#| fig-height: 6.2

paid_per_bene_plot <- ggplot(
  outliers |> dplyr::filter(paid_per_bene > 0, !is.na(paid_per_bene)),
  aes(paid_per_bene)
) +
  geom_histogram(fill = "#020073", color = "white", bins = 60) +
  scale_x_log10(labels = scales::label_dollar()) +
  labs(x = "Paid per beneficiary (log10)", y = "Rows", subtitle = "Per-row paid/bene")

bene_plot <- ggplot(
  outliers |> dplyr::filter(beneficiaries > 0, beneficiaries <= 5000),
  aes(beneficiaries)
) +
  geom_histogram(fill = "#020073", color = "white", bins = 60) +
  labs(x = "Beneficiaries (≤ 5,000)", y = "Rows", subtitle = "Per-row beneficiary count")

peer_plot <- ggplot(hcpcs_sum, aes(peer_n)) +
  geom_histogram(fill = "#020073", color = "white", bins = 60) +
  scale_x_log10(labels = scales::label_comma()) +
  labs(x = "Peer providers per HCPCS (log10)", y = "HCPCS codes",
       subtitle = "Peer-group sizes")

z_plot <- ggplot(
  outliers |> dplyr::filter(!is.na(robust_z),
                            robust_z >= -10, robust_z <= 50),
  aes(robust_z)
) +
  geom_histogram(fill = "#020073", color = "white", bins = 60) +
  geom_vline(xintercept = 5, linetype = "dashed",
             color = "#FF2E00", linewidth = 0.6) +
  labs(x = "Robust z (clipped to [-10, 50])", y = "Rows",
       subtitle = "Robust-z distribution")

# Use patchwork if available, then gridExtra, then a plain stacked print.
if (requireNamespace("patchwork", quietly = TRUE)) {
  patchwork::wrap_plots(paid_per_bene_plot, bene_plot,
                        peer_plot, z_plot, ncol = 2)
} else if (requireNamespace("gridExtra", quietly = TRUE)) {
  gridExtra::grid.arrange(paid_per_bene_plot, bene_plot,
                          peer_plot, z_plot, ncol = 2)
} else {
  for (p in list(paid_per_bene_plot, bene_plot, peer_plot, z_plot)) print(p)
}
```

```{r eda-geo-coverage}
#| label: eda-geo-coverage

geo_cov <- outliers |>
  dplyr::summarise(
    rows_total      = dplyr::n(),
    rows_with_state = sum(!is.na(state)),
    rows_with_geo   = sum(!is.na(lat) & !is.na(lon)),
    rows_with_spec  = sum(!is.na(specialty_name))
  ) |>
  dplyr::mutate(
    pct_state = 100 * rows_with_state / rows_total,
    pct_geo   = 100 * rows_with_geo   / rows_total,
    pct_spec  = 100 * rows_with_spec  / rows_total
  )

cat(sprintf(paste0(
    "Outlier-table rows total            : %s\n",
    "  with NPPES state                  : %s (%.1f%%)\n",
    "  with NPPES specialty              : %s (%.1f%%)\n",
    "  with ZCTA → lat/lon centroid      : %s (%.1f%%)\n"),
  format(geo_cov$rows_total,      big.mark = ","),
  format(geo_cov$rows_with_state, big.mark = ","), geo_cov$pct_state,
  format(geo_cov$rows_with_spec,  big.mark = ","), geo_cov$pct_spec,
  format(geo_cov$rows_with_geo,   big.mark = ","), geo_cov$pct_geo
))
```

The geo-coverage line is the join-quality check that any geographic claim downstream depends on: a county-level atlas built from a 60% NPPES-state coverage would mean the map is showing 60% of the data, not 100% — and the difference matters for any conclusion about state-level Medicaid policy.

### Overdispersion and zero-inflation diagnostics

Two formal checks before any future modeling on the count-like features. The variance-to-mean ratio (VMR) tests whether Poisson is even on the table — a VMR much greater than 1 means the variance grows faster than the mean and you need negative binomial (or quasi-Poisson). Pairing VMR with the % zero count is what motivates a **zero-inflated NB** rather than a plain NB.

```{r eda-overdispersion}
#| label: eda-overdispersion

vmr <- function(x) {
  x <- x[!is.na(x) & x >= 0]
  if (length(x) == 0 || mean(x) == 0) return(c(mean = NA, var = NA, vmr = NA))
  c(mean = mean(x), var = stats::var(x), vmr = stats::var(x) / mean(x))
}
zero_pct <- function(x) {
  x <- x[!is.na(x)]
  if (!length(x)) return(NA_real_)
  100 * mean(x == 0)
}

vmr_tbl <- rbind(
  paid          = vmr(outliers$paid),
  beneficiaries = vmr(outliers$beneficiaries),
  claims        = vmr(outliers$claims),
  months_active = vmr(outliers$months_active)
)

family_call <- function(v) {
  if (is.na(v)) return("—")
  if (v < 1.2)              "Poisson"
  else if (v < 5)           "Quasi-Poisson / NB"
  else                      "Negative binomial / ZINB"
}

pretty_kable(
  data.frame(
    Variable     = rownames(vmr_tbl),
    Mean         = sprintf("%.1f", vmr_tbl[, "mean"]),
    Variance     = sprintf("%.1f", vmr_tbl[, "var"]),
    VMR          = sprintf("%.1f", vmr_tbl[, "vmr"]),
    `% zeros`    = sprintf("%.2f%%",
                            vapply(list(outliers$paid, outliers$beneficiaries,
                                        outliers$claims, outliers$months_active),
                                   zero_pct, numeric(1))),
    `Family`     = vapply(vmr_tbl[, "vmr"], family_call, character(1)),
    check.names = FALSE
  ),
  caption = "Variance-to-mean ratio (VMR), zero-inflation, and the implied regression family for each count-like feature. Caveat: the outliers parquet is already filtered to robust_z >= 2 OR top-20 spend codes AND beneficiaries >= 11 in prep, so the % zeros reads near 0 by construction. The VMR is still informative because it characterizes the conditional dispersion that any future modeling step will need to absorb."
)
```

::: {.callout-note collapse="true"}
## Going deeper — why $\text{Var} = E$ for Poisson, and what NB / ZINB / hurdle each fix

**Plain English.** When your data is *counts* (number of beneficiaries, claims, visits), the simplest probability model is the **Poisson** — but Poisson has a weird hidden property: its variance always equals its average. If you see data where the variance is 5× the average, Poisson is *wrong*; the data is more spread out than the model can produce. You move to **negative binomial (NB)**, which adds a "fudge factor" for extra spread. If on top of that, *way* too many of your observations are zero (30% of doctors prescribe zero opioids — way more than NB predicts), you need **zero-inflated NB (ZINB)**, which models the zeros separately. The **VMR** (variance-to-mean ratio) is the quick screen: ≈ 1 → Poisson fits; > 2 → use NB; large with excess zeros → use ZINB or a hurdle model.

**The Poisson identity.** A Poisson random variable with rate $\lambda$ has probability mass function $\Pr(X = k) = \frac{\lambda^k e^{-\lambda}}{k!}$. Computing its mean and variance from the moment-generating function $M_X(t) = e^{\lambda(e^t - 1)}$:
$$E[X] = M_X'(0) = \lambda, \qquad E[X^2] = M_X''(0) = \lambda^2 + \lambda \;\Longrightarrow\; \text{Var}(X) = \lambda.$$
So **for Poisson, $\text{Var}(X) = E[X]$ exactly** — they're the same number. The variance-to-mean ratio $\text{VMR} = \text{Var}/E$ equals 1 in expectation, with sampling fluctuations around 1. **VMR meaningfully greater than 1 falsifies Poisson** — the data have *overdispersion*.

**Negative binomial.** Built as a Poisson-Gamma mixture: $X \mid \mu \sim \text{Poisson}(\mu)$ where $\mu$ itself varies (e.g., $\mu \sim \text{Gamma}(\theta)$). The marginal distribution of $X$ has
$$E[X] = \mu_0, \qquad \text{Var}(X) = \mu_0 + \frac{\mu_0^2}{\theta}.$$
The extra $\mu_0^2/\theta$ is unobserved heterogeneity — different observations have different true rates, even after conditioning on covariates. NB is the workhorse fix for overdispersed counts.

**Zero-inflated NB (ZINB).** A *mixture* model with two latent components: with probability $\pi$, the observation is a "structural zero" (always $X = 0$); with probability $1 - \pi$, it's drawn from a regular NB:
$$\Pr(X = 0) = \pi + (1 - \pi)\Pr_{\text{NB}}(0), \qquad \Pr(X = k > 0) = (1 - \pi)\Pr_{\text{NB}}(k).$$
Use ZINB when zeros are *categorically different* from small counts — e.g., a doctor who *never* prescribes opioids vs. one who prescribes them rarely. The structural-zero process is modeled with a logistic regression on a possibly-different predictor set.

**Hurdle model.** Different mechanism: model the zero/positive *threshold* with one distribution (typically logistic), then model the *positive counts only* with a truncated NB. Use when "any vs none" is governed by a different process than "how much" — e.g., insurance enrollment (any claims at all) is determined by enrollment, but conditional on enrolling, claim volume depends on health.

**Worked example.** 1,000 NPIs, each filing a count of beneficiaries. If $E[X] = 40$ and $\text{Var}(X) = 380$, then VMR $= 9.5$ — strongly overdispersed. An NB fit gives $\mu_0 = 40$ and $\theta$ from $380 = 40 + 1600/\theta$, so $\theta = 4.7$ — moderately heavy heterogeneity. If 5% of NPIs have $X = 0$ but the NB predicts 8% zeros, the NB fit is fine; if 25% of NPIs have $X = 0$ vs. NB's 8%, you need ZINB or a hurdle.
:::

### Within-flagged-set selection bias

The outliers parquet is itself a non-random sample of the full DuckDB output (kept rows are those with `robust_z >= 2` OR billed under the top-20 spending HCPCS codes). We can't directly compare against the dropped rows without re-running the 11 GB DuckDB pipeline, but we *can* check whether the **high-z subset** (`robust_z >= 5`, the program-integrity threshold) systematically differs from the **moderate-z subset** (`2 <= robust_z < 5`) on the variables we'd otherwise model — a "what makes the flagged-flagged set different from the merely flagged" check.

```{r eda-within-cohort-bias}
#| label: eda-within-cohort-bias

with_groups <- outliers |>
  dplyr::filter(!is.na(robust_z), robust_z >= 2) |>
  dplyr::mutate(
    z_group = dplyr::if_else(robust_z >= 5, "High z (>= 5)", "Moderate z (2 - 5)")
  )

n_hi  <- sum(with_groups$z_group == "High z (>= 5)")
n_mod <- sum(with_groups$z_group == "Moderate z (2 - 5)")

cont_results <- list(
  beneficiaries  = stats::wilcox.test(beneficiaries  ~ z_group, data = with_groups),
  claims         = stats::wilcox.test(claims         ~ z_group, data = with_groups),
  paid           = stats::wilcox.test(paid           ~ z_group, data = with_groups),
  paid_per_bene  = stats::wilcox.test(paid_per_bene  ~ z_group, data = with_groups),
  months_active  = stats::wilcox.test(months_active  ~ z_group, data = with_groups)
)

state_tab <- table(with_groups$state, with_groups$z_group)
spec_tab  <- table(with_groups$specialty_class, with_groups$z_group)

cat_tests <- list(
  state           = suppressWarnings(stats::chisq.test(state_tab)),
  specialty_class = suppressWarnings(stats::chisq.test(spec_tab))
)

bias_df <- data.frame(
  Variable    = c(names(cont_results), names(cat_tests)),
  Test        = c(rep("Mann-Whitney U", length(cont_results)),
                  rep("Pearson chi-squared", length(cat_tests))),
  Statistic   = c(
    vapply(cont_results, \(r) round(as.numeric(r$statistic), 0), numeric(1)),
    vapply(cat_tests,    \(r) round(as.numeric(r$statistic), 1), numeric(1))
  ),
  `p-value`   = c(
    vapply(cont_results, \(r) format.pval(r$p.value, eps = 1e-300, digits = 2),
           character(1)),
    vapply(cat_tests,    \(r) format.pval(r$p.value, eps = 1e-300, digits = 2),
           character(1))
  ),
  check.names = FALSE
)

cat(sprintf("High-z (>= 5)         : %s rows\nModerate-z (2 - 5)    : %s rows\n\n",
            format(n_hi,  big.mark = ","),
            format(n_mod, big.mark = ",")))
knitr::kable(
  bias_df,
  caption = "Within-flagged-set bias check: do high-z (>= 5) rows differ from moderate-z (2 - 5) rows on the inputs that would feed any peer-comparison model? The 'high-z is just larger providers' null can be ruled out by checking the beneficiary and claim distributions; specialty mix differences expose which clinical areas drive the high-z tail."
)
```

```{r pareto}
#| fig-cap: "Cumulative share of total Medicaid paid dollars, ordered by HCPCS code from highest- to lowest-spending. A steep early curve means a small set of codes drives most public outlay."

hcpcs_totals <- hcpcs_sum |>
  arrange(desc(total_paid)) |>
  mutate(
    rank_pct  = row_number() / n(),
    cum_share = cumsum(total_paid) / sum(total_paid)
  )

ggplot(hcpcs_totals, aes(rank_pct, cum_share)) +
  geom_line(color = "#020073", linewidth = 1) +
  geom_hline(yintercept = 0.8, linetype = "dashed", color = "#FF2E00") +
  scale_x_continuous(labels = percent_format()) +
  scale_y_continuous(labels = percent_format()) +
  labs(x = "HCPCS codes ranked by total paid (ascending)",
       y = "Cumulative share of total paid")
```

## Peer-group outlier detection (R)

> **In one sentence.** Compute a MAD-based robust z per (NPI × HCPCS) row, then apply Benjamini–Hochberg FDR ≤ 0.01 instead of a hard z-cutoff so the flag list comes with a multiplicity guarantee, not just an arbitrary threshold.

::: {.callout-note}
## Why a robust (MAD-based) z and not a normal z?

The **classical z-score** is `(x − mean) / SD`. It works beautifully when the data are roughly normal. It fails completely on heavy-tailed data because **a few extreme outliers blow up the SD**, which then deflates every other observation's z toward zero — *the outliers hide themselves by inflating the noise scale*.

The **robust z** replaces the two non-robust pieces:

- **Median** instead of mean (robust to outliers; doesn't move when you change extreme values)
- **MAD** (median absolute deviation) instead of SD (also robust). Multiplied by **1.4826** so that under a normal distribution, MAD × 1.4826 estimates the SD — making the resulting z-score interpretable on the familiar scale ("z = 5 means about 5 SDs from the median").

Use robust z whenever your data have heavy tails (claims data, gene expression, financial returns, anything log-skewed). Use the classical z when you've genuinely confirmed normality (rare in real-world health data). (Hampel 1974; Rousseeuw & Croux 1993)
:::

::: {.callout-note collapse="true"}
## Going deeper — where the constant 1.4826 comes from

**Plain English.** Imagine a spreadsheet of a thousand numbers, sorted smallest to largest. The middle one is the *median*. For every row, compute "how far is it from the median" — that's the absolute deviation. The median of *those* deviations is the **MAD** (median absolute deviation). MAD by itself is a small number — not the same scale as a standard deviation. The constant 1.4826 is the rescaling factor that turns MAD into something that *equals* the standard deviation, *if your data is normally distributed*. So a "robust z" computed as $(x - \text{median}) / (1.4826 \times \text{MAD})$ reads on the same scale as a regular z-score ("z = 5 means about 5 standard deviations from the median"), but built from numbers that don't blow up when outliers show up.

**Intuition.** We want MAD to estimate $\sigma$ (so robust z and classical z line up under normality). But the *raw* MAD is much smaller than $\sigma$ for normal data — it's the median absolute deviation, not a true standard deviation. The constant 1.4826 is precisely the factor that rescales MAD to be a consistent estimator of $\sigma$ under the normal model.

**Derivation.** Let $X \sim N(\mu, \sigma^2)$. The MAD is $\text{MAD} = \text{median}\!\left(|X - \mu|\right)$. By symmetry of the normal, $|X - \mu|$ has half-normal distribution, and its median satisfies
$$\Pr(|X - \mu| \leq \text{MAD}) = \tfrac{1}{2} \;\Longleftrightarrow\; \Pr\!\left(\tfrac{|X-\mu|}{\sigma} \leq \tfrac{\text{MAD}}{\sigma}\right) = \tfrac{1}{2}.$$
The standardised quantity $|Z| = |X-\mu|/\sigma$ has CDF $2\Phi(z) - 1$. Setting that equal to $\tfrac{1}{2}$ gives
$$\Phi(\text{MAD}/\sigma) = \tfrac{3}{4} \;\Longrightarrow\; \text{MAD}/\sigma = \Phi^{-1}(0.75) \approx 0.6745.$$
Inverting: $\sigma \approx \text{MAD} / 0.6745 = 1.4826 \cdot \text{MAD}$. The constant is exact for normal data; for other symmetric distributions it's an approximation.

**Breakdown point.** The fraction of arbitrarily-bad observations an estimator can tolerate before producing nonsense. The mean has breakdown point 0 — one bad observation moves it arbitrarily. The median has breakdown point 0.5 (you'd need to corrupt more than half the data to move it). MAD has breakdown point 0.5 as well — it's the same robustness as the median, applied to deviations rather than levels.

**Worked example.** Take ten observations: $\{10, 11, 12, 13, 14, 15, 16, 17, 18, 1000\}$. The mean is 112.6; the SD is ~313. A classical z for the value 1000 is $(1000 - 112.6)/313 \approx 2.83$ — barely flagged. The median is 14.5; absolute deviations from the median are $\{4.5, 3.5, 2.5, 1.5, 0.5, 0.5, 1.5, 2.5, 3.5, 985.5\}$; the MAD = 2.5. Robust z for 1000: $(1000 - 14.5)/(1.4826 \cdot 2.5) \approx 266$. The robust estimator sees what the classical one missed because its scale parameter didn't get inflated by the outlier itself.
:::

::: {.callout-note}
## Why Benjamini–Hochberg FDR — and what does "FDR ≤ 0.01" actually mean?

When you test thousands of hypotheses, **some will look significant just by chance**. At p < 0.05 across 100,000 tests, you'd expect 5,000 false positives even if no real signal exists. A hard z ≥ 5 cutoff has the same problem: at z = 5 in a normal null, p ≈ 6 × 10⁻⁷, but across 100,000 tests you'd still expect ~0.06 false positives **per HCPCS code** — and the data have hundreds of HCPCS codes.

The **Benjamini–Hochberg (BH) procedure** controls the **false discovery rate (FDR)** — the *expected fraction of false positives among the rows you flag*. "FDR ≤ 0.01" means: of the rows the rule flags, no more than 1% are expected to be false positives. That's a much more honest claim than "p < 0.05" on each individual test.

How BH works: sort the p-values from smallest to largest. For the i-th smallest, compare it to (i / m) × α, where m is the total number of tests and α is your target FDR. The largest i where the test passes defines the cutoff; flag everything at or below it. Built into base R as `p.adjust(method = "BH")`.

Use BH whenever you're testing many hypotheses simultaneously — genomics, neuroimaging, fraud detection, A/B-testing dashboards. **Bonferroni** (divide α by m) is more conservative; use Bonferroni when you need a strict family-wise error rate guarantee, BH when controlling the false discovery proportion is enough. (Benjamini & Hochberg 1995; Benjamini & Yekutieli 2001)
:::

::: {.callout-note collapse="true"}
## Going deeper — why the BH algorithm actually controls FDR

**Plain English.** You're running 10,000 statistical tests. If you call every test with $p < 0.05$ "significant," about 500 will look significant *just by chance* — even if nothing real is going on. That's the multiple-comparisons problem. **Bonferroni** is the heavy-handed fix: demand $p < 0.05/10{,}000 = 0.000005$ for any test to count. Very strict, but you'll miss real signals because the threshold is so harsh. **BH-FDR** is a smarter middle ground: it sorts your p-values smallest to largest, then sets a cutoff such that *at most 1% of the tests you call significant are false alarms, in expectation*. So 99 out of every 100 flagged hits are real (on average). The guarantee is weaker than Bonferroni's "no false alarms ever," but it's the right tradeoff when you're sifting through thousands of tests and want to keep some real signal.

**Intuition.** The two key insights:
1. **Under the null**, p-values are uniformly distributed on $[0, 1]$. So if you have $m_0$ true nulls, the expected number of them with $p_{(i)} \leq t$ is exactly $m_0 \cdot t$.
2. The BH cutoff is chosen so that *if we declare the smallest $i$ p-values significant*, the **expected** count of false rejections among those $i$ rejections is at most $\alpha \cdot i$.

**The algorithm.** Sort p-values: $p_{(1)} \leq p_{(2)} \leq \cdots \leq p_{(m)}$. Find the largest index $i^*$ such that
$$p_{(i^*)} \leq \frac{i^*}{m} \cdot \alpha.$$
Reject the null for tests $1, 2, \ldots, i^*$. Equivalently, define the **q-value**
$$q_{(i)} = \min_{j \geq i} \frac{m \cdot p_{(j)}}{j},$$
and reject all tests with $q_{(i)} \leq \alpha$. The min-from-the-right ensures monotonicity (we never reject a less-significant test while keeping a more-significant one).

**Why it controls FDR.** Under independence (or "positive regression dependency"), Benjamini & Hochberg proved
$$\text{FDR} = E\!\left[\frac{V}{\max(R, 1)}\right] \leq \frac{m_0}{m} \cdot \alpha \leq \alpha,$$
where $V$ = false discoveries, $R$ = total discoveries, $m_0$ = number of true nulls. The proof uses a martingale argument on the ordered p-values; the key step is that, conditional on a null p-value taking a particular value $t$, the remaining nulls are still uniform on $[t, 1]$ — and the cutoff $i\alpha/m$ scales linearly with $i$, exactly cancelling that uniformity.

**Bonferroni contrast.** Bonferroni rejects $p_{(i)} \leq \alpha/m$ for *all* $i$ — a fixed, much tighter threshold. It controls the **family-wise error rate (FWER)**: the probability of *any* false rejection $\leq \alpha$. That's a much stronger guarantee, and it costs power: with $m = 1{,}000$ tests at $\alpha = 0.05$, Bonferroni demands $p \leq 5 \times 10^{-5}$, while BH could reject all tests up to $p \leq 0.05$ if the empirical distribution supports it.

**Worked example.** Ten tests sorted by p-value: $0.001, 0.008, 0.012, 0.04, 0.05, 0.10, 0.15, 0.30, 0.50, 0.80$. Target $\alpha = 0.05$, $m = 10$. The BH cutoffs $i\alpha/m$ are $0.005, 0.010, 0.015, 0.020, 0.025, 0.030, 0.035, 0.040, 0.045, 0.050$. Compare element-wise:

| $i$ | $p_{(i)}$ | $i\alpha/m$ | Pass? |
|---:|---:|---:|---|
| 1 | 0.001 | 0.005 | ✓ |
| 2 | 0.008 | 0.010 | ✓ |
| 3 | 0.012 | 0.015 | ✓ |
| 4 | 0.040 | 0.020 | ✗ |
| 5 | 0.050 | 0.025 | ✗ |

The largest $i^*$ where the test passes is 3, so BH rejects the first 3 hypotheses. Bonferroni would reject only $p \leq 0.005$ — just test 1. BH found 2 extra discoveries with at most 5% expected false-discovery proportion.
:::

The DuckDB prep computed a robust z-score per HCPCS using the median-of-absolute-deviations (MAD) scaled by 1.4826. Values ≥ 5 correspond roughly to "5σ from the peer median for this code," which is the threshold CMS Program Integrity uses as a first-pass screen — but at this scale of testing (hundreds of thousands of NPI × HCPCS rows), a fixed z-cutoff with no multiplicity control will produce many false positives by chance. We compute Benjamini–Hochberg false discovery rate adjusted q-values alongside the raw z, and use FDR ≤ 0.01 as the publishable flagging threshold.

```{r bh-fdr-flagging}
#| label: bh-fdr-flagging

flagged <- outliers |>
  dplyr::filter(!is.na(robust_z), beneficiaries >= 30) |>
  dplyr::mutate(
    # Two-sided p from a normal approximation on the robust z. This is
    # an approximation — the MAD-z null distribution isn't exactly
    # normal, but a normal reference is the standard CMS PI starting
    # point and gives a conservative FDR.
    p_value  = 2 * stats::pnorm(-abs(robust_z)),
    q_value  = stats::p.adjust(p_value, method = "BH"),
    flag_z5  = robust_z >= 5,
    flag_fdr = q_value <= 0.01 & robust_z > 0
  )

flagging_summary <- data.frame(
  Rule                 = c("Raw z >= 5",
                            "BH-FDR <= 0.01 (one-sided)",
                            "Both rules agree (intersection)"),
  Flagged              = c(sum(flagged$flag_z5),
                            sum(flagged$flag_fdr),
                            sum(flagged$flag_z5 & flagged$flag_fdr))
)
flagging_summary$`% of tested` <- sprintf("%.2f%%",
  100 * flagging_summary$Flagged / nrow(flagged))

pretty_kable(
  flagging_summary,
  caption = sprintf(
    "Flagging counts for %s NPI x HCPCS rows tested. The raw z >= 5 cutoff is what CMS uses as a first-pass screen but applies no multiplicity control. The BH-FDR rule guarantees no more than 1%% of flagged rows are false positives in expectation. The intersection is the highest-confidence subset for downstream review.",
    format(nrow(flagged), big.mark = ",")
  )
)
```

```{r top-outliers}
top_outliers <- flagged |>
  filter(flag_fdr) |>
  arrange(q_value, desc(robust_z)) |>
  slice_head(n = 25) |>
  transmute(
    npi, state, specialty_name, hcpcs,
    beneficiaries, paid,
    paid_per_bene       = paid / beneficiaries,
    hcpcs_median_ppb,
    robust_z            = round(robust_z, 1),
    q_value             = format.pval(q_value, eps = 1e-300, digits = 2)
  )

pretty_kable(
  top_outliers |>
    mutate(
      paid              = format_money(paid),
      paid_per_bene     = format_money(paid_per_bene),
      hcpcs_median_ppb  = format_money(hcpcs_median_ppb)
    ),
  caption = "Top 25 peer-group outliers ranked by BH-FDR q-value (FDR ≤ 0.01, ≥ 30 beneficiaries). The q-value column is the false-discovery-adjusted p — these are the rows the multiplicity-controlled rule keeps."
)
```

```{r specialty-boxplot}
#| fig-cap: "Per-beneficiary spend by specialty (top 15 by flagged-outlier count). Box spans p25–p75 with median; whiskers are p05 / p95 from the full peer distribution computed in the DuckDB prep step. Orange dots are the individual robust-z ≥ 5 outliers layered on top."

top_specialties <- specialty_sum |>
  arrange(desc(flagged_ge_5)) |>
  slice_head(n = 15) |>
  pull(specialty_name)

box_df <- specialty_sum |>
  filter(specialty_name %in% top_specialties) |>
  mutate(specialty_name = fct_reorder(specialty_name, p50))

jitter_df <- outliers |>
  filter(specialty_name %in% top_specialties, robust_z >= 5) |>
  mutate(specialty_name = factor(specialty_name, levels = levels(box_df$specialty_name)))

ggplot(box_df, aes(specialty_name)) +
  geom_boxplot(
    aes(
      ymin   = p05,  lower  = p25,
      middle = p50,  upper  = p75,
      ymax   = p95
    ),
    stat = "identity",
    fill = "#e8edff", color = "#020073", width = 0.7
  ) +
  geom_jitter(data = jitter_df,
              aes(y = paid_per_bene),
              color = "#FF2E00", width = 0.2, size = 0.7, alpha = 0.5) +
  scale_y_log10(labels = scales::label_dollar()) +
  coord_flip() +
  labs(x = NULL, y = "Paid per beneficiary (log scale)")
```

## Unsupervised comparison (Python)

::: {.callout-note}
## Why an isolation forest as a second opinion?

The robust z asks one question: *is this row's paid-per-beneficiary unusually far from the peer median?* It looks at one variable at a time. **Isolation forest** asks a richer question: *is this row unusual considering the joint distribution of all features at once?* A row could look normal on every individual feature but still be an outlier in the joint space (high beneficiaries + low claims + high paid is a different story from high beneficiaries + high claims + high paid).

How it works: the algorithm builds many random binary trees by splitting on random features at random thresholds. **Outliers are easier to isolate** — they get separated into their own leaf in fewer splits. The "anomaly score" for a row is roughly the average path length from root to its leaf across the trees; shorter path = more anomalous. No labels needed.

Use unsupervised methods (isolation forest, local outlier factor, one-class SVM, autoencoders) when you don't have a labeled training set of "fraud" vs "not fraud" — which is almost always the case in healthcare. Treat the agreement between a rule-based method (robust z) and an unsupervised method as your **highest-confidence flag set**, and the disagreement as the *interesting* set worth a human review. (Liu, Ting & Zhou 2008, 2012)
:::

::: {.callout-note collapse="true"}
## Going deeper — isolation forest scoring and the $c(n)$ normalisation

**Plain English.** Suppose you have 10,000 data points in some feature space and you want to flag the weird ones — but you don't know in advance what "weird" looks like. Pick a feature at random, pick a cutoff at random, split the data into two halves. Repeat the split on each half. Keep going until every point sits alone in its own little leaf. **The weirdos get isolated first** — they're sitting out in low-density regions, so a single random cut separates them from the herd. The "anomaly score" for each point is roughly "how few splits did it take to isolate this point?" Run several hundred random trees, average the path lengths, and that's your score. The $c(n)$ formula below is just a rescaling factor that lets you compare scores across different sample sizes.

**The construction.** Each isolation tree (iTree) is built by recursively splitting a random subsample of $\psi$ observations as follows: pick a feature uniformly at random, then a split value uniformly between the feature's min and max. Stop when each row is in its own leaf, or you hit a depth limit. A "normal" point requires many splits to isolate (it has many neighbours in feature space, each requiring a separate cut); an outlier — being in a low-density region — gets isolated quickly. The **path length** $h(x)$ from root to the leaf containing $x$ is the diagnostic.

**Why path length needs normalising.** Path lengths depend on the subsample size $\psi$. To compare across different $\psi$ values (or to interpret on a fixed scale), divide by the *expected* path length of a binary search tree with $\psi$ nodes:
$$c(\psi) = 2\,H(\psi - 1) - \frac{2(\psi - 1)}{\psi}, \quad H(i) = \ln(i) + 0.5772 \;(\text{Euler–Mascheroni}).$$
This is the average path length to a random leaf in an unsuccessful BST search — Knuth's classical result.

**The anomaly score.** With $E[h(x)]$ the average path length across all trees,
$$s(x, \psi) = 2^{-E[h(x)] / c(\psi)}.$$
Properties:
- $s \to 1$ as $E[h(x)] \to 0$ (instantly isolated → strong anomaly).
- $s \to 0$ as $E[h(x)] \to \infty$ (hard to isolate → likely inlier).
- $s \approx 0.5$ when $E[h(x)] \approx c(\psi)$ (average-depth leaf → no signal).

The exponential form maps an unbounded path length onto a bounded $[0, 1]$ score, making the threshold "anomaly if $s > 0.5$" interpretable.

**The contamination parameter.** `contamination=0.01` tells the model to threshold at the empirical 1% upper tail — i.e., flag the top 1% of scores as anomalies. The choice is *not* a property of the data; it's a budget for how many alerts you're willing to investigate. Set higher when you can afford more reviews; set lower when false positives are expensive.

**Worked example.** Sample size $\psi = 256$ (the default). Then $H(255) = \ln(255) + 0.5772 \approx 6.12$, so $c(256) = 12.24 - 1.99 = 10.25$. A row with average path length 4 across 100 trees gets score $s = 2^{-4/10.25} = 0.764$ — strong anomaly. A row with average path length 12 gets $s = 2^{-12/10.25} = 0.446$ — inlier-ish, below threshold.
:::

A peer-group z-score is simple and auditable but scale-sensitive. An unsupervised model sees the joint distribution of features and can flag records that look unusual across multiple axes at once. We fit an isolation forest on the top 20 HCPCS by total spend, standardize, and score.

### Feature collinearity, briefly

Isolation forests are tree-based and don't inflate standard errors the way a linear model would, but heavy collinearity still costs the ensemble *tree diversity* — the splits become redundant across features that move together. A two-minute check before fitting:

```{python iforest-collinearity}
from sklearn.linear_model import LinearRegression

df = r.outliers.copy()
top_codes = (
    df.groupby("hcpcs")["paid"].sum().sort_values(ascending=False).head(20).index
)
sub = df[df["hcpcs"].isin(top_codes)].copy()
sub = sub.dropna(subset=["paid_per_bene", "beneficiaries", "claims", "months_active"])

feats   = ["paid_per_bene", "beneficiaries", "claims", "months_active"]
X_raw   = sub[feats]

print("Pearson correlation matrix among isolation-forest features:")
print(X_raw.corr(method="pearson").round(2).to_string())

Xs   = StandardScaler().fit_transform(X_raw.to_numpy())
sv   = np.linalg.svd(Xs, compute_uv=False)
cond = sv.max() / sv.min()
print(f"\nCondition number (kappa) of standardized design matrix: {cond:.1f}")
print("Rule of thumb: kappa < 10 clean, 10-30 moderate, > 30 problematic.")

vifs = {}
for j, name in enumerate(feats):
    others = [k for k in range(len(feats)) if k != j]
    r2     = LinearRegression().fit(Xs[:, others], Xs[:, j]).score(Xs[:, others], Xs[:, j])
    vifs[name] = round(1.0 / max(1 - r2, 1e-9), 2)

print("\nApproximate VIF per feature (1 / (1 - R^2_other)):")
for k, v in vifs.items():
    print(f"  {k:<16s} : {v}")
```

What the numbers mean here: `claims` and `beneficiaries` are essentially the same construct measured two ways (each claim covers approximately one beneficiary), so r ≈ 0.95+ and a VIF in the double digits on both is the expected reading — not a bug. `paid_per_bene` is by construction orthogonal to `beneficiaries` (it's paid divided by beneficiaries) and `months_active` carries an independent activity-duration axis. We keep all four features anyway: the isolation forest splits don't *over-count* the redundant pair the way a linear regression would, and dropping `claims` would lose the small but real per-beneficiary multi-claim variation in the long tail. The condition-number print just makes that decision auditable rather than implicit.

::: {.callout-note collapse="true"}
## Going deeper — singular values and the condition number $\kappa$

**Plain English.** When you fit a regression with multiple predictors, some of those predictors might be near-duplicates of each other (BMI and waist circumference, total claims and beneficiaries). When two predictors carry essentially the same information, the regression has trouble deciding "which one is doing the work" — the coefficients become unstable, flipping around dramatically with tiny changes to the data. The **condition number** is a single-number diagnostic that quantifies "how nearly-duplicated are my predictors, *across the whole design at once*?" Below 10 = clean. Above 30 = trouble. The math under the hood (the SVD — singular value decomposition) decomposes your predictor matrix into the "most-correlated direction" and the "least-correlated direction"; the condition number is the ratio of their stretches.

**Intuition.** A matrix $\mathbf{X}$ "stretches" different directions of input space by different amounts. The **singular values** $\sigma_1 \geq \sigma_2 \geq \cdots \geq \sigma_p \geq 0$ are the stretch factors along the most-stretched, second-most-stretched, ..., least-stretched orthogonal directions. The **condition number** $\kappa = \sigma_1 / \sigma_p$ asks: how much wider is the most-stretched direction than the least? When $\kappa$ is large, the matrix is *nearly singular* — small changes in $y$ produce huge changes in $\hat\beta$, because the "narrow" direction barely moves the prediction at all.

**Formula.** The **singular value decomposition (SVD)** of any $n \times p$ matrix is $\mathbf{X} = \mathbf{U} \mathbf{\Sigma} \mathbf{V}^\top$ where $\mathbf{U}$ and $\mathbf{V}$ are orthogonal and $\mathbf{\Sigma}$ is diagonal with the $\sigma_i$ on the diagonal. The condition number is
$$\kappa(\mathbf{X}) = \frac{\sigma_1}{\sigma_p}.$$
For the OLS normal equations, the relative error in $\hat\beta$ is bounded by $\kappa$ times the relative error in $y$ — so $\kappa = 100$ means tiny perturbations in $y$ can shift $\hat\beta$ by up to 100×.

**Why $\kappa > 30$ is "problematic".** Belsley, Kuh & Welsch (1980) tied condition-number thresholds to specific levels of numerical instability:
- $\kappa < 10$ — multicollinearity is negligible.
- $10 \leq \kappa < 30$ — moderate; coefficients have inflated SEs but are interpretable.
- $\kappa \geq 30$ — severe; near-collinear directions; coefficients are unstable to small data perturbations.

**Connection to VIF.** Both diagnose the same thing. VIF is per-feature ("how much is *this* coefficient's variance inflated?"); $\kappa$ is global ("how badly conditioned is the *whole* design?"). They're related: $\kappa^2 \geq \max_j \text{VIF}_j$ for a standardised design matrix, so a high VIF implies a high $\kappa$, but $\kappa$ can be elevated by interactions VIF misses.

**Worked example.** Three standardised features. Their covariance matrix's eigenvalues are 2.8, 0.95, 0.05 (the SVD gives $\sigma_i^2 = $ eigenvalues, so $\sigma$ values are $\sqrt{2.8}, \sqrt{0.95}, \sqrt{0.05} \approx 1.67, 0.97, 0.22$). Then $\kappa = 1.67 / 0.22 \approx 7.6$ — moderate, no action needed. If the smallest singular value were 0.01 instead, $\kappa = 167$ — drop a feature, or use ridge regression to add a small constant to the diagonal of $\mathbf{X}^\top\mathbf{X}$ (which is exactly what ridge does to bound $\kappa$).
:::

### Fitting the isolation forest

```{python iforest}
X = StandardScaler().fit_transform(X_raw.to_numpy())

iso = IsolationForest(
    n_estimators=400, contamination=0.01, random_state=42, n_jobs=-1
)
sub["iforest_score"]    = -iso.fit(X).score_samples(X)   # higher = more anomalous
sub["iforest_flag"]     = iso.predict(X) == -1

print(f"Rows analysed: {len(sub):,}")
print(f"Isolation forest flags: {int(sub['iforest_flag'].sum()):,} ({sub['iforest_flag'].mean()*100:.2f}%)")
print("Overlap with robust-z ≥ 5:",
      int((sub['iforest_flag'] & (sub['robust_z'] >= 5)).sum()))
```

```{python iforest-scatter}
#| fig-cap: "Paid-per-beneficiary vs. beneficiary count for the top 20 HCPCS codes by spend. Isolation-forest flags in orange overlap substantially with — but not identically to — the robust-z rule."

fig, ax = plt.subplots(figsize=(7, 5))
_ = ax.scatter(sub["beneficiaries"], sub["paid_per_bene"],
               c=np.where(sub["iforest_flag"], "#FF2E00", "#c3c3c3"),
               s=10, alpha=0.5)
_ = ax.set(xscale="log", yscale="log",
           xlabel="Unique beneficiaries (log)", ylabel="Paid per beneficiary (log)")
_ = ax.spines[["top", "right"]].set_visible(False)
plt.tight_layout()
plt.show()
```

## Where does the dollar go? A Medicaid category cost atlas

> **In one sentence.** A county-level interactive choropleth across all 12 HCPCS spending categories, shaded by within-category decile of paid-per-beneficiary — coloring is comparable inside a category, not across them.

A provider-outlier list is a fraud-review input. Most policy audiences, though, want the opposite question: **which categories of care cost the most, and where do those costs concentrate geographically?** Switching from "who" to "what and where" requires three pieces we've now built upstream in [`prep.py`](prep.py):

1. A rule-based **HCPCS → category** classifier that buckets every procedure, drug, DME, and Medicaid-specific T-code into one of 12 spending categories.
2. A **Census 2020 ZCTA → county** crosswalk so every provider's practice ZIP resolves to a county FIPS.
3. A county × category aggregate with **within-category quintile rankings** so a single color scale reads consistently across panels: darker = this county is in the top 20% of cost-per-beneficiary for *that* category.

These buckets are a mix of clinical areas (Behavioral Health, Dental, Dialysis & ESRD, Prescription Drugs) and service-type groupings that the HCPCS code system itself organizes by (Surgery & Procedures, Radiology, Laboratory, DME, Transportation, LTSS, E&M) — not a true disease taxonomy. Reading the atlas as "categories of Medicaid spend" is more honest than reading it as "diseases": HCPCS doesn't carry diagnosis, so the closest we can get without ICD is spending-category geography.

::: {.callout-note}
## Why county, not HSA or ZIP3?

For a Medicaid cost atlas, counties are the defensible middle path — clean TIGER polygons, HUD/Census maintained crosswalks, and the unit most federal-agency publications (MACPAC, KFF, Urban Institute) use. HSAs (Dartmouth Atlas) are the gold standard for *Medicare* market analysis but were built from Medicare patient flows, so they're slightly miscalibrated for Medicaid. ZIP3 is tempting because it's one `LEFT(zip, 3)` away, but ZIP3 boundaries are USPS logistics geography — a high Modifiable-Areal-Unit-Problem exposure for a spatial-stats story.
:::

```{r cost-atlas-setup}
county_atlas <- arrow::read_parquet(file.path(data_dir, "county_group_atlas.parquet"))
group_totals <- arrow::read_parquet(file.path(data_dir, "group_totals.parquet"))

top_categories <- group_totals |>
  arrange(desc(total_paid)) |>
  pull(disease_group)

# Recompute a within-category decile rank on paid-per-beneficiary. The parquet
# carries quintiles from prep.py; we override with 10 bins here so shading
# shows more granular within-category variation without re-running the
# DuckDB prep. For categories with < 100 billing counties some deciles will
# be tied — the lambda collapses those cleanly.
county_atlas <- county_atlas |>
  group_by(disease_group) |>
  mutate(decile_ppb = dplyr::ntile(paid_per_bene, 10)) |>
  ungroup()
```

```{r group-bar}
#| fig-cap: "Total national Medicaid paid by HCPCS-derived category, 2024. Ranked descending by total paid. 'Other Medical Services' captures 99000-series codes the classifier doesn't yet disambiguate."
#| fig-width: 9
#| fig-height: 4.5

group_totals |>
  mutate(category = fct_reorder(disease_group, total_paid)) |>
  ggplot(aes(category, total_paid)) +
  geom_col(fill = "#746aff") +
  coord_flip() +
  scale_y_continuous(labels = format_money) +
  labs(x = NULL, y = "Total paid (USD)")
```

```{r county-choropleth}
#| fig-cap: "Interactive county-level cost atlas: Medicaid paid per beneficiary ranked into within-category deciles (darker navy = higher cost relative to other counties for that same category). Each panel is its own choropleth so coloring is comparable within category, not across categories. Use the top-right dropdown to switch across all 12 HCPCS spending categories. Alaska, Hawaii, and territories omitted for contiguous-US framing; grey counties have no Medicaid providers billing that category."

options(tigris_use_cache = TRUE)
counties_sf <- tigris::counties(year = 2020, cb = TRUE, resolution = "5m", progress_bar = FALSE) |>
  dplyr::select(GEOID, geometry) |>
  dplyr::filter(!stringr::str_detect(GEOID, "^(02|15|60|66|69|72|78)")) |>
  sf::st_transform(4326)

navies <- grDevices::colorRampPalette(c("#F6F9F7", "#020073"))(10)
pal <- leaflet::colorFactor(navies, domain = factor(1:10))

atlas_sub <- county_atlas |>
  filter(disease_group %in% top_categories)

lmap <- leaflet::leaflet(width = "100%", height = "620px") |>
  leaflet::addProviderTiles(
    leaflet::providers$CartoDB.PositronNoLabels,
    options = leaflet::providerTileOptions(opacity = 0.75)
  ) |>
  leaflet::setView(lng = -96, lat = 38, zoom = 4) |>
  leaflet::addPolygons(
    data        = counties_sf,
    fillColor   = "#e5e5e5",
    fillOpacity = 0.6,
    color       = "white",
    weight      = 0.3,
    smoothFactor = 0.5,
    options     = leaflet::pathOptions(interactive = FALSE)
  )

for (g in top_categories) {
  dat <- counties_sf |>
    dplyr::inner_join(
      atlas_sub |> dplyr::filter(disease_group == g),
      by = c("GEOID" = "county_fips")
    )
  lmap <- lmap |>
    leaflet::addPolygons(
      data         = dat,
      group        = g,
      fillColor    = ~pal(factor(decile_ppb)),
      fillOpacity  = 0.88,
      color        = "white",
      weight       = 0.3,
      smoothFactor = 0.5,
      highlightOptions = leaflet::highlightOptions(
        weight       = 2,
        color        = "#333",
        fillOpacity  = 0.95,
        bringToFront = TRUE
      ),
      popup = ~paste0(
        "<b>", county_name, ", ", state, "</b><br/>",
        "<em>", disease_group, "</em><br/>",
        "Paid / beneficiary: $",
        format(round(paid_per_bene), big.mark = ","), "<br/>",
        "Total paid: $",
        format(round(total_paid / 1e3), big.mark = ","), "K<br/>",
        "Providers: ", providers_n, "<br/>",
        "Within-category decile: ", decile_ppb, " of 10"
      )
    )
}

categories_json <- jsonlite::toJSON(top_categories)
dropdown_js <- sprintf("
function(el, x) {
  var map = this;
  var categories = %s;
  var initial = categories[0];

  function showOnly(picked) {
    categories.forEach(function(g) {
      var fg = map.layerManager.getLayerGroup(g, false);
      if (!fg) return;
      if (g === picked) {
        if (!map.hasLayer(fg)) fg.addTo(map);
      } else {
        if (map.hasLayer(fg)) map.removeLayer(fg);
      }
    });
  }
  showOnly(initial);

  var ctrl = L.control({position: 'topright'});
  ctrl.onAdd = function() {
    var div = L.DomUtil.create('div', 'leaflet-bar disease-group-control');
    div.style.background = 'white';
    div.style.padding = '8px 10px';
    div.style.boxShadow = '0 1px 4px rgba(0,0,0,0.2)';
    div.style.font = '12px/1.3 -apple-system, BlinkMacSystemFont, \"Segoe UI\", sans-serif';

    var label = document.createElement('label');
    label.textContent = 'Category';
    label.style.cssText = 'display:block;font-weight:600;margin-bottom:4px;color:#333;';

    var sel = document.createElement('select');
    sel.style.cssText = 'padding:4px 6px;font-size:12px;border:1px solid #ccc;border-radius:3px;background:white;min-width:240px;';
    categories.forEach(function(g) {
      var opt = document.createElement('option');
      opt.value = g;
      opt.textContent = g;
      if (g === initial) opt.selected = true;
      sel.appendChild(opt);
    });
    sel.addEventListener('change', function(e) { showOnly(e.target.value); });

    div.appendChild(label);
    div.appendChild(sel);
    L.DomEvent.disableClickPropagation(div);
    L.DomEvent.disableScrollPropagation(div);
    return div;
  };
  ctrl.addTo(map);
}
", categories_json)

lmap |>
  htmlwidgets::onRender(dropdown_js) |>
  leaflet::addLegend(
    position = "bottomleft",
    colors   = navies,
    labels   = c("1 (lowest)", "2", "3", "4", "5", "6", "7", "8", "9", "10 (highest)"),
    title    = "Decile of paid<br/>per beneficiary",
    opacity  = 0.85
  )
```

## Category explorer

> **In one sentence.** Three companion views of the atlas — top 5 HCPCS per category, spatial Gini concentration, and state hotspot table — that answer the three follow-up questions a policy reviewer always asks next.

The choropleth shows *where* a category is expensive. The three views below answer three follow-up questions policy audiences almost always ask next: **what HCPCS codes drive the spend inside each category, how spatially concentrated is that spend across counties, and which states sit at the top of the ladder?** Every chunk iterates over all 12 categories — no cherry-picking — so the honest shape of Medicaid dollar flow is visible in one sweep.

```{r hcpcs-classifier}
#| code-fold: show
#| code-summary: "HCPCS → category classifier (mirrors the DuckDB CASE in prep.py)"

classify_hcpcs <- function(code) {
  num    <- suppressWarnings(as.integer(code))
  is_cpt <- !is.na(num) & nchar(code) == 5
  dplyr::case_when(
    stringr::str_starts(code, "J")                               ~ "Prescription Drugs & Biologics",
    stringr::str_starts(code, "D")                               ~ "Dental",
    stringr::str_starts(code, "H")                               ~ "Behavioral Health",
    stringr::str_starts(code, "E") |
      stringr::str_starts(code, "K") |
      stringr::str_starts(code, "L")                             ~ "Durable Medical Equipment & Supplies",
    stringr::str_detect(code, "^T(20|21|22|51)") |
      stringr::str_starts(code, "A04")                           ~ "Transportation",
    stringr::str_starts(code, "T")                               ~ "Long-Term Services & Supports",
    code %in% c("G0257","G0323","G0324","G0325","G0326","G0327") ~ "Dialysis & ESRD",
    is_cpt & num >= 90791 & num <= 90899                         ~ "Behavioral Health",
    is_cpt & num >= 90935 & num <= 90999                         ~ "Dialysis & ESRD",
    is_cpt & num >= 99201 & num <= 99499                         ~ "Evaluation & Management",
    is_cpt & num >= 10021 & num <= 69990                         ~ "Surgery & Outpatient Procedures",
    is_cpt & num >= 70010 & num <= 79999                         ~ "Radiology & Imaging",
    is_cpt & num >= 80047 & num <= 89398                         ~ "Laboratory & Diagnostics",
    TRUE                                                         ~ "Other Medical Services"
  )
}

hcpcs_by_category <- hcpcs_sum |>
  mutate(category = classify_hcpcs(hcpcs)) |>
  filter(category %in% top_categories)
```

### What codes drive each category?

```{r category-hcpcs-mix}
#| fig-cap: "Top 5 HCPCS codes by total Medicaid paid within each category. The codes carrying the weight are very different per category — J-codes (specialty infusibles and injectables) dominate the drug bucket; Transportation collapses onto A0/T20-series non-emergency transport; LTSS is driven by T-series personal-care and case-management codes; E&M is dominated by the 99213/99214 office-visit pair; a single periodic-exam code (D0120) and a pair of restoration codes drive most of Dental."
#| fig-width: 11
#| fig-height: 11

top_hcpcs_per_category <- hcpcs_by_category |>
  dplyr::group_by(category) |>
  dplyr::arrange(dplyr::desc(total_paid), .by_group = TRUE) |>
  dplyr::slice_head(n = 5) |>
  dplyr::ungroup() |>
  dplyr::mutate(
    category = factor(category, levels = top_categories),
    row_id   = dplyr::row_number()
  )

ggplot(top_hcpcs_per_category,
       aes(x = stats::reorder(paste(hcpcs, row_id), total_paid),
           y = total_paid)) +
  geom_col(fill = "#746aff") +
  geom_text(aes(label = format_money(total_paid)),
            hjust = -0.08, size = 3, color = "#333") +
  facet_wrap(~ category, scales = "free", ncol = 3) +
  scale_x_discrete(labels = \(x) sub(" \\d+$", "", x)) +
  scale_y_continuous(labels = format_money,
                     expand = expansion(mult = c(0, 0.22))) +
  coord_flip() +
  labs(x = NULL, y = "Total paid (all billing providers)")
```

### How spatially concentrated is each category?

::: {.callout-note}
## Why the Gini coefficient — and what does "Gini = 0.6" mean?

The Gini coefficient measures **how unequally a quantity is distributed across units**. Originally an income-inequality measure; here we apply it to "Medicaid paid per county within a category."

- **Gini = 0**: perfectly equal — every county has the same total paid in that category.
- **Gini = 1**: one county has everything; the rest have nothing.
- **Gini ≈ 0.4–0.6**: moderate concentration — typical of national health-spending categories.
- **Gini > 0.7**: spend collapses onto a handful of counties — expected for sparse, high-acuity categories like Dialysis & ESRD.

Geometrically, Gini is twice the area between the **Lorenz curve** (cumulative share of paid plotted against cumulative share of counties, sorted by paid) and the diagonal of perfect equality. We pair Gini with the **top-10-county share** because Gini is a single number that hides the shape; the top-10 share answers "concretely, how much does the top 10 carry?"

Use Gini for any concentration analysis: spending across counties, hospitalizations across providers, citations across papers, wealth across households. **Don't use it for symmetric-around-zero quantities** (it's defined on non-negative values). (Lorenz 1905; Gini 1912; Cowell 2011)
:::

::: {.callout-note collapse="true"}
## Going deeper — Gini derived from the Lorenz curve

**Plain English.** Imagine you sort everyone in a country by income, poorest to richest, and plot "what fraction of total income belongs to the poorest X% of people?" If everyone earned exactly the same, the plot is the diagonal line y = x. If one person had all the income, the line would hug the bottom axis and then jump to 1 at the very end. Real inequality lives between. The **Gini coefficient** measures how far the actual line bends away from the diagonal: 0 = perfect equality (line *is* the diagonal), 1 = total inequality (one person has it all). Same idea works for Medicaid spending across counties: 0 = perfectly diffuse, 0.5 = moderately concentrated, > 0.7 = collapses onto a handful of counties.

**The Lorenz curve.** Sort $n$ units ascending by their value $x_{(1)} \leq x_{(2)} \leq \cdots \leq x_{(n)}$. Plot, for each $k$, the *fraction of units* $k/n$ against the *fraction of total* held by the bottom-$k$ units: $L_k = \sum_{i=1}^k x_{(i)} \,/\, \sum_{i=1}^n x_{(i)}$. The resulting step curve, joined at the corners, is the Lorenz curve. Under perfect equality $L(p) = p$ — the diagonal. Under perfect inequality (one unit has everything) the curve hugs the bottom axis until the final point, then jumps to 1.

**Gini = twice the area between the diagonal and the Lorenz curve.** Geometrically,
$$G = \frac{\text{Area between diagonal and Lorenz curve}}{\text{Area under the diagonal}} = 1 - 2 \int_0^1 L(p)\, dp.$$
Since the area under the diagonal is $\tfrac{1}{2}$, $G$ is exactly 2× the gap-area. Maximum $G = 1$ (Lorenz curve hugs the axis); minimum $G = 0$ (Lorenz curve = diagonal).

**The discrete formula.** Substituting the Lorenz step formula and computing the integral via trapezoids gives, for sorted ascending values:
$$G = \frac{2 \sum_{i=1}^n i \cdot x_{(i)}}{n \sum_{i=1}^n x_{(i)}} - \frac{n + 1}{n}.$$
This is the formula in the `gini()` function above. Sanity check: with equal values $x_{(i)} = c$, $\sum i \cdot c = c \cdot n(n+1)/2$ and $\sum x = cn$, so the first term is $\tfrac{n+1}{n}$, giving $G = 0$. ✓

**Theil index** is the natural alternative — entropy-based: $T = \tfrac{1}{n}\sum (x_i / \bar{x})\log(x_i / \bar{x})$. Theil is *decomposable* into within- and between-group components (Gini isn't), so it's the choice when you want to attribute concentration to subgroups.

**Worked example.** Four counties with paid totals $\{10, 20, 50, 200\}$. Sorted, sum $= 280$. Then $\sum i \cdot x_{(i)} = 1\cdot10 + 2\cdot20 + 3\cdot50 + 4\cdot200 = 10 + 40 + 150 + 800 = 1000$. So
$$G = \frac{2 \cdot 1000}{4 \cdot 280} - \frac{5}{4} = \frac{2000}{1120} - 1.25 = 1.786 - 1.25 = 0.536.$$
Substantial concentration — the top county holds 71% of the spend, and Gini $\approx 0.54$ confirms it.
:::

```{r category-concentration}
#| fig-cap: "Gini coefficient of county Medicaid paid within each category, and the share of that category's total spend captured by just the top 10 counties. A Gini near 1 means spend collapses onto a handful of counties — expected for sparse, high-acuity categories like Dialysis & ESRD; Gini near 0 is evenly diffused public outlay, like LTSS and E&M."
#| fig-width: 9
#| fig-height: 5.5

gini <- function(x) {
  x <- sort(x[!is.na(x) & x >= 0])
  n <- length(x)
  if (n <= 1 || sum(x) == 0) return(NA_real_)
  2 * sum(seq_len(n) * x) / (n * sum(x)) - (n + 1) / n
}

spatial_conc <- county_atlas |>
  filter(disease_group %in% top_categories) |>
  group_by(category = disease_group) |>
  summarise(
    counties_n    = dplyr::n(),
    gini_paid     = gini(total_paid),
    top10_share   = sum(sort(total_paid, decreasing = TRUE)[seq_len(min(10, dplyr::n()))],
                        na.rm = TRUE) / sum(total_paid, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(gini_paid))

spatial_conc |>
  mutate(category = fct_reorder(category, gini_paid)) |>
  ggplot(aes(category, gini_paid)) +
  geom_col(fill = "#746aff") +
  geom_text(aes(label = sprintf("Gini %.2f · top-10 share %s",
                                gini_paid,
                                scales::percent(top10_share, accuracy = 1))),
            hjust = -0.05, size = 3.1, color = "#333") +
  coord_flip(clip = "off") +
  scale_y_continuous(limits = c(0, 1.05),
                     labels = scales::number_format(accuracy = 0.01),
                     expand = expansion(mult = c(0, 0))) +
  labs(x = NULL, y = "Gini of county Medicaid paid (within category)") +
  theme(plot.margin = margin(5, 110, 5, 5))
```

### Which states lead each category?

```{r category-state-hotspots}
state_hotspots <- county_atlas |>
  filter(disease_group %in% top_categories, !is.na(state)) |>
  group_by(state, category = disease_group) |>
  summarise(
    state_paid = sum(total_paid,  na.rm = TRUE),
    state_bene = sum(total_bene,  na.rm = TRUE),
    counties   = dplyr::n(),
    .groups = "drop"
  ) |>
  filter(state_bene >= 1000) |>
  mutate(state_ppb = state_paid / state_bene) |>
  group_by(category) |>
  arrange(desc(state_ppb), .by_group = TRUE) |>
  slice_head(n = 5) |>
  ungroup() |>
  mutate(category = factor(category, levels = top_categories)) |>
  arrange(category, desc(state_ppb))

pretty_kable(
  state_hotspots |>
    transmute(
      Category             = category,
      State                = state,
      Counties             = counties,
      `Total paid`         = format_money(state_paid),
      Beneficiaries        = format(state_bene, big.mark = ","),
      `$ per beneficiary`  = format_money(state_ppb)
    ),
  caption = "Top 5 states by Medicaid paid-per-beneficiary within each category (states with at least 1,000 beneficiaries in that category)."
)
```

## State-level intensity

```{r state-intensity}
#| fig-cap: "Peer-group-excess Medicaid spending by state. Height is the sum of paid dollars above each HCPCS's peer median, aggregated at prep time across all providers billing in that state."

state_excess <- state_sum |>
  arrange(desc(excess_paid)) |>
  slice_head(n = 25)

ggplot(state_excess, aes(reorder(state, excess_paid), excess_paid)) +
  geom_col(fill = "#020073") +
  coord_flip() +
  scale_y_continuous(labels = format_money) +
  labs(x = NULL, y = "Paid above HCPCS peer median ($)")
```

## Conclusions

The seven-step pipeline (DuckDB aggregate → robust z → NPPES enrichment → ZCTA-county join → category classifier → atlas → explorer) supports five conclusions that each tie back to a specific diagnostic on this page:

- **Heavy tails are the regime, not the exception.** The paid-per-beneficiary distribution in [`eda-distribution`](#cell-fig-eda-distribution) is log-skewed across four orders of magnitude. That's why a mean/SD z-score is unusable here: the SD is dominated by one or two mega-providers and every other observation looks "normal" by comparison. The MAD-based robust z is invariant to those tail rows, which is why the same threshold (≥ 5) generalizes across HCPCS codes with very different peer-group sizes.
- **Rule-based and unsupervised flags agree on the high-confidence set.** The overlap between robust-z ≥ 5 and the isolation-forest 1%-contamination flag is the set a Program Integrity team would prioritize. Disagreement between the two is the *interesting* set — it's where a tail-only metric (robust z) and a joint-distribution metric (isolation forest) see different things, which is usually a signal worth a second human look.
- **Collinearity didn't break the isolation forest, but it constrained tree diversity.** The condition number on the four standardized features lands above the rule-of-thumb threshold because `claims` and `beneficiaries` are nearly identical by construction. We kept both for interpretability — but the cost is that some trees in the ensemble split on essentially the same axis. A leaner three-feature variant (drop `claims`) would produce a slightly more diverse forest at the cost of a slightly less interpretable feature set; for a fraud-review prompt list, that tradeoff goes the way we chose.
- **Categories are an HCPCS lens, not a disease lens.** The 12-bucket atlas is honest about what it is: a procedure-code taxonomy projected onto county geography, not a disease taxonomy. Reading the choropleth for "where Medicaid spends money on Behavioral Health procedures" is supportable; reading it for "where Medicaid pays for treating depression" is not — depression doesn't have a HCPCS code, only the *services* delivered for depression do. The Gini concentration plot is a useful tie-breaker: categories with high Gini (Dialysis & ESRD, parts of DME) genuinely cluster geographically; categories with low Gini (LTSS, E&M) are diffuse across the country and shouldn't be read as having geographic "hotspots" at all.
- **Geography is the start of an interpretation, never the end.** State-level Medicaid rate schedules, MCO contracting differences, and referral-network composition can all produce county-level patterns that look like billing outliers but are artifacts of policy. The state hotspot tables in the explorer make those patterns visible; deciding which are actionable requires a payer-policy lens this notebook does not pretend to provide.

## Caveats and what would change the picture

- **Single year, no temporal control.** The dataset is the most recent monthly drop; nothing here separates persistent outliers from transient spikes (COVID-era catch-up billing, one-off audits, policy shocks). Rolling 3-month windows would.
- **Provider geocoding gaps shape the map.** The geo-coverage diagnostic in EDA shows the share of outlier rows missing a state or ZCTA centroid; counties whose providers have unresolved addresses are *under-represented* on the choropleth, not just absent. The grey baseline layer makes "no data" visible but doesn't distinguish "no providers billing this category" from "providers exist but didn't geocode."
- **HCPCS coverage is sparse for some categories.** Dialysis & ESRD is billed in roughly 500 counties; the within-category decile shading there is meaningful but bins are sparse, so the visual difference between deciles 4 and 5 is less stable than for densely-covered categories like E&M.

## Next steps

- **Specialty-adjusted peer groups.** Today the peer group is "everyone billing this HCPCS"; tighter peer groups (same specialty, similar patient panel size) will surface a different, higher-quality signal — especially in categories where the peer pool mixes radically different practice types.
- **Time-aware outlier detection.** Repeat the analysis on rolling 3-month windows to separate persistent outliers from transient spikes; the same robust-z framework applies, the only change is the partitioning column in the DuckDB prep.
- **SDoH overlay.** Merge state-level CDC PLACES / ACS indicators to test whether residual excess spending tracks community need vs. billing behavior — a health-equity stress test for the flagged list. The county-FIPS join already exists; adding a tract-level layer is a one-join extension.

## Methods primer

Every statistical test, estimator, or transformation used above, with a one-line definition, what it does here, the key assumption, what to do if that assumption fails, and the canonical primary source. Full citations are in the [References](#references) section that follows.

| Method | What it does | Why used here | Key assumption | If assumption fails | Reference |
|---|---|---|---|---|---|
| **Median, MAD, robust z** | Median = middle value (robust to outliers); MAD = median absolute deviation; robust z = (x − median) / (1.4826 × MAD), interpretable as approximate SDs from the median under normality. | Per-HCPCS peer-comparison flagging — heavy-tailed claims data make mean/SD z's worthless. | Errors roughly symmetric around the median; 1.4826 calibration exact only for normal data. | Use bootstrap-derived peer percentiles. | Hampel 1974; Rousseeuw & Croux 1993 |
| **Variance-to-mean ratio (VMR)** | The ratio var(x) / mean(x). For Poisson VMR = 1; values above 1 indicate overdispersion. | Diagnose whether count-like features need NB or ZINB rather than Poisson in any future modeling. | Variable is non-negative count-like. | None — descriptive. | Cameron & Trivedi 2013 |
| **Zero-inflation diagnostic** | Fraction of observations equal to zero, paired with VMR. | Flag when a count distribution has more zeros than NB can produce → use zero-inflated models. | Zeros are meaningfully distinct from small positive counts. | Use a hurdle model. | Lambert 1992; Mullahy 1986 |
| **Pearson correlation** | Linear association between two continuous variables; r ∈ [−1, +1]. | Detect collinear features before fitting the isolation forest (claims vs beneficiaries are nearly the same construct). | Linear relationship; no extreme outliers. | Use Spearman rank correlation. | Pearson 1896 |
| **Condition number (κ)** | Ratio of largest to smallest singular value of the standardized design matrix. κ > 30 = problematic collinearity. | Single-number summary of multicollinearity across all features at once. | Standardized features (mean 0, var 1). | Drop redundant features or use ridge. | Belsley, Kuh & Welsch 1980 |
| **Variance Inflation Factor (VIF)** | For each feature, regress it on the others; VIF = 1 / (1 − R²). | Per-feature collinearity diagnostic for the isolation-forest input set. | Linear regression for the auxiliary fit. | Drop the redundant feature. | Marquardt 1970; Belsley, Kuh & Welsch 1980 |
| **Two-sided p-value from a normal approximation** | Convert a z-score to a probability by `2 × Φ(−\|z\|)`. | Step needed before applying BH-FDR — turn each robust z into a tail probability. | Null distribution of robust z approximately normal. | Use a permutation-derived empirical null. | Standard normal-tail formula |
| **Benjamini–Hochberg FDR** (`p.adjust(method = "BH")`) | Adjusts a vector of p-values to control the false discovery rate — the expected fraction of false positives among rejections. | Required when testing thousands of NPI × HCPCS rows simultaneously. | Independent or positively-dependent tests. | Use Benjamini–Yekutieli for arbitrary dependence. | Benjamini & Hochberg 1995 |
| **Benjamini–Yekutieli FDR** | FDR correction valid under arbitrary dependence between tests. | Cited as the fallback when BH's positive-dependence assumption is unsafe. | None on dependence structure; trades power for safety. | N/A | Benjamini & Yekutieli 2001 |
| **Mann-Whitney U / Wilcoxon rank-sum** | Non-parametric test that one distribution is shifted relative to another. Operates on ranks. | Compare beneficiary counts and paid amounts between high-z and moderate-z subsets. | Independent observations; no distributional assumption. | Permutation test. | Wilcoxon 1945; Mann & Whitney 1947 |
| **Pearson chi-squared** | Tests independence between two categorical variables. | Compare specialty mix and state distribution between flagged subsets. | Expected cell counts ≥ 5 in most cells. | Use Fisher's exact test. | Pearson 1900 |
| **Isolation forest** | Unsupervised anomaly detector that builds many random binary trees and scores points by how easily they get isolated (shorter average path = more anomalous). | Second-opinion flagging on the joint distribution of features. | Anomalies are "few and different." | Use one-class SVM, LOF, or an autoencoder. | Liu, Ting & Zhou 2008, 2012 |
| **Standardization** (`StandardScaler`) | Centers each feature at zero and scales to unit variance. | Required before isolation forest splits compare across features at the same scale. | Features are roughly continuous. | Use rank-based standardization for heavy-tailed features. | Standard preprocessing |
| **NTILE / decile binning** (`dplyr::ntile`) | Splits a vector into n equal-sized rank bins. | Within-category ranking for the choropleth — coloring comparable inside a category, not across. | Continuous variable with few ties. | Use `cut()` on quantiles to allow unequal bins. | Conover 1999 |
| **Gini coefficient** | Twice the area between the Lorenz curve and the diagonal of perfect equality. 0 = uniform; 1 = one unit has everything. | Quantify spatial concentration of spend within each Medicaid category. | Non-negative quantity; no missing values. | Report top-K share or Theil index. | Gini 1912; Cowell 2011 |
| **Lorenz curve** | Cumulative share of the quantity plotted against cumulative share of units (sorted ascending). | Geometric basis for the Gini coefficient. | Non-negative quantity. | Use a CDF plot. | Lorenz 1905 |
| **Top-K share** | Sum of the K largest values divided by the total. | Companion to Gini — "the top 10 counties carry X% of the spend." | Ordering by the same metric being summed. | None — fully transparent. | Standard descriptive statistic |
| **`patchwork::wrap_plots`** | Composes multiple ggplot objects into one figure with shared layout. | Pack four diagnostic histograms into a single 2×2 panel. | All inputs are ggplot objects. | Use `gridExtra::grid.arrange` or base layouts. | Pedersen 2024 |
| **Provider taxonomy joins (NPPES + NUCC + ZCTA → county)** | Sequence of left joins to enrich each NPI with state, specialty, and county-FIPS via NPPES → NUCC taxonomy and Census ZCTA crosswalks. | Without this enrichment, flagged outliers are just NPIs; with it they're "an internist in Hennepin County, MN." | Crosswalk tables match correctly on join keys. | Report the % of rows that fail the join (geo-coverage diagnostic). | CMS NPPES; NUCC 2025; US Census ZCTA |

## Caveats and limitations primer

- **Peer group is too broad.** "Everyone billing this HCPCS" mixes specialties, panel sizes, and care settings. The publishable peer is `specialty × panel-size band × setting`. Largest single robustness gap.
- **Single monthly drop, no temporal control.** Rolling 3-month windows would separate persistent outliers from transient spikes (COVID catch-up billing, one-off audits, policy shocks).
- **Geocoding gaps.** A fraction of NPPES providers don't resolve to a Census ZCTA centroid; those counties under-represent on the map.
- **Category sparsity.** Dialysis & ESRD is billed in ~500 counties; decile boundaries get visually closer there than for densely-covered categories like E&M.

## References {#references}

Statistical methods (chronological by topic):

- **Lorenz, M. O.** (1905). Methods of measuring the concentration of wealth. *Publications of the American Statistical Association*, 9(70), 209–219.
- **Gini, C.** (1912). *Variabilità e mutabilità.* Tipografia di Paolo Cuppini, Bologna.
- **Pearson, K.** (1896). Mathematical contributions to the theory of evolution. III. Regression, heredity, and panmixia. *Philosophical Transactions of the Royal Society A*, 187, 253–318.
- **Pearson, K.** (1900). On the criterion that a given system of deviations from the probable in the case of a correlated system of variables is such that it can be reasonably supposed to have arisen from random sampling. *Philosophical Magazine*, 50(302), 157–175.
- **Wilcoxon, F.** (1945). Individual comparisons by ranking methods. *Biometrics Bulletin*, 1(6), 80–83.
- **Mann, H. B., & Whitney, D. R.** (1947). On a test of whether one of two random variables is stochastically larger than the other. *Annals of Mathematical Statistics*, 18(1), 50–60.
- **Marquardt, D. W.** (1970). Generalized inverses, ridge regression, biased linear estimation, and nonlinear estimation. *Technometrics*, 12(3), 591–612.
- **Hampel, F. R.** (1974). The influence curve and its role in robust estimation. *Journal of the American Statistical Association*, 69(346), 383–393.
- **Belsley, D. A., Kuh, E., & Welsch, R. E.** (1980). *Regression Diagnostics: Identifying Influential Data and Sources of Collinearity.* Wiley.
- **Mullahy, J.** (1986). Specification and testing of some modified count data models. *Journal of Econometrics*, 33(3), 341–365.
- **Lambert, D.** (1992). Zero-inflated Poisson regression, with an application to defects in manufacturing. *Technometrics*, 34(1), 1–14.
- **Rousseeuw, P. J., & Croux, C.** (1993). Alternatives to the median absolute deviation. *Journal of the American Statistical Association*, 88(424), 1273–1283.
- **Benjamini, Y., & Hochberg, Y.** (1995). Controlling the false discovery rate: a practical and powerful approach to multiple testing. *Journal of the Royal Statistical Society, Series B*, 57(1), 289–300.
- **Conover, W. J.** (1999). *Practical Nonparametric Statistics* (3rd ed.). Wiley.
- **Benjamini, Y., & Yekutieli, D.** (2001). The control of the false discovery rate in multiple testing under dependency. *Annals of Statistics*, 29(4), 1165–1188.
- **Liu, F. T., Ting, K. M., & Zhou, Z.-H.** (2008). Isolation forest. In *Eighth IEEE International Conference on Data Mining* (pp. 413–422).
- **Liu, F. T., Ting, K. M., & Zhou, Z.-H.** (2012). Isolation-based anomaly detection. *ACM Transactions on Knowledge Discovery from Data*, 6(1), Article 3.
- **Cowell, F. A.** (2011). *Measuring Inequality* (3rd ed.). Oxford University Press.
- **Cameron, A. C., & Trivedi, P. K.** (2013). *Regression Analysis of Count Data* (2nd ed.). Cambridge University Press.
- **Pedersen, T. L.** (2024). *patchwork: The Composer of Plots* (R package). <https://patchwork.data-imaginist.com>

Data sources:

- **Centers for Medicare & Medicaid Services (CMS).** Medicaid Provider Spending. HHS Open Data Portal. <https://opendata.hhs.gov/datasets/medicaid-provider-spending/>
- **CMS.** NPPES (National Plan and Provider Enumeration System) Downloadable File. <https://download.cms.gov/nppes/NPI_Files.html>
- **National Uniform Claim Committee (NUCC).** Health Care Provider Taxonomy Code Set, version 25.0 (2025). <https://taxonomy.nucc.org/>
- **U.S. Census Bureau.** 2020 ZIP Code Tabulation Areas (ZCTA) Gazetteer; 2020 ZCTA-County Relationship File. <https://www.census.gov/geographies/reference-files/time-series/geo/gazetteer-files.html>

---

::: {.callout-tip}
### Reading this as code samples
Every cell above is the real executed code — not screenshots. Use the **Code** dropdown at the top-right to toggle visibility, copy blocks, or view the [raw `.qmd` source](https://github.com/paulinedmdf/paulinadelmundomd/blob/main/projects/02-medicaid-outliers/analysis.qmd) on GitHub. The one-shot DuckDB prep step lives in [`prep.py`](https://github.com/paulinedmdf/paulinadelmundomd/blob/main/projects/02-medicaid-outliers/prep.py).
:::
