---
title: "Outlier Detection in Medicaid Provider Spending"
subtitle: "Peer-group robust z-scores, isolation forests, and a geographic lens on provider-level Medicaid claims"
author: "Paulina Del Mundo Del Fierro, MD, MPH"
date: last-modified
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-location: right
    code-tools: true
    code-copy: true
    code-fold: false
    code-overflow: wrap
    number-sections: true
    fig-responsive: true
    fig-width: 8
    fig-height: 5
    embed-resources: true
    include-in-header:
      text: |
        <script>
          window.va = window.va || function () { (window.vaq = window.vaq || []).push(arguments); };
        </script>
        <script defer src="/_vercel/insights/script.js"></script>
execute:
  echo: true
  warning: false
  message: false
  freeze: auto
---

## Why this analysis

Medicaid is the single largest source of health coverage for low-income Americans — and its claims data, released monthly by HHS, is a rare window into how public dollars flow through the US provider network. The question for any physician-scientist looking at this data is not *"who spends the most"* (large systems dominate by construction) but *"who spends atypically, given their peers?"* That's the signal that drives fraud-and-abuse reviews, network-integrity audits, and honest comparative-effectiveness work alike.

This notebook does three things end-to-end:

1. **Streams the 11 GB Medicaid Provider Spending CSV with DuckDB**, aggregates to the NPI × HCPCS level, and flags peer-group outliers using a median/MAD-based robust z-score — resistant to the heavy tails claims data is famous for.
2. **Joins NPPES, NUCC taxonomy, and US Census ZCTA centroids** so every flagged provider carries an interpretable specialty and a map-able location. Nothing in the analysis requires cloud infrastructure — everything runs on a laptop against committed Parquet artifacts.
3. **Contrasts a deterministic rule (robust z) against an unsupervised model (isolation forest)** on the same cohort, in R and Python respectively. Agreement between the two is the real outlier story; disagreement is where a physician-scientist earns their keep.

::: {.callout-note}
## Data sources
- **Medicaid Provider Spending** — HHS Open Data Portal, accessed 2026-04-21. <https://opendata.hhs.gov/datasets/medicaid-provider-spending/>
- **NPPES Downloadable File** — Centers for Medicare & Medicaid Services. <https://download.cms.gov/nppes/NPI_Files.html>
- **NUCC Health Care Provider Taxonomy** — National Uniform Claim Committee. <https://taxonomy.nucc.org/>
- **ZCTA Gazetteer (2020)** — US Census Bureau. <https://www.census.gov/geographies/reference-files/time-series/geo/gazetteer-files.html>

Raw files stay local (11 GB + 45 GB NPPES); the [`prep.py`](prep.py) script runs one DuckDB pipeline to produce three committed Parquets this notebook reads.
:::

---

## Setup

```{r setup-r}
suppressPackageStartupMessages({
  library(arrow)
  library(dplyr)
  library(ggplot2)
  library(tidyr)
  library(stringr)
  library(forcats)
  library(scales)
  library(sf)
  library(leaflet)
  library(reticulate)
})

reticulate::py_require(c("pandas", "numpy", "scikit-learn", "matplotlib"))

theme_set(theme_minimal(base_size = 12))

data_dir     <- here::here("projects", "02-medicaid-outliers", "data", "processed")
outliers     <- arrow::read_parquet(file.path(data_dir, "outliers.parquet"))
hcpcs_sum    <- arrow::read_parquet(file.path(data_dir, "hcpcs_summary.parquet"))
state_sum    <- arrow::read_parquet(file.path(data_dir, "state_summary.parquet"))
specialty_sum<- arrow::read_parquet(file.path(data_dir, "specialty_summary.parquet"))
taxonomy     <- arrow::read_parquet(file.path(data_dir, "taxonomy.parquet"))

format_money <- function(x) scales::label_dollar(accuracy = 1, scale_cut = scales::cut_short_scale())(x)
```

```{python setup-py}
import pandas as pd
import numpy as np
import matplotlib.pyplot as plt
from sklearn.ensemble import IsolationForest
from sklearn.preprocessing import StandardScaler

plt.rcParams.update({"figure.dpi": 110})
```

## Cohort sketch

```{r cohort-sketch}
summary_tbl <- tibble::tibble(
  hcpcs_n       = nrow(hcpcs_sum),
  providers_n   = outliers |> dplyr::pull(npi) |> dplyr::n_distinct(),
  total_paid    = sum(hcpcs_sum$total_paid, na.rm = TRUE),
  total_bene    = sum(hcpcs_sum$total_bene, na.rm = TRUE),
  flagged_ge_5  = sum(outliers$robust_z >= 5, na.rm = TRUE)
)

knitr::kable(
  summary_tbl |>
    mutate(total_paid = format_money(total_paid),
           across(where(is.numeric), \(x) format(round(x, 2), big.mark = ","))),
  caption = "Cohort sketch from DuckDB prep outputs"
)
```

## Exploratory data analysis — what the prep step actually produced

The DuckDB pipeline emits seven Parquet artifacts. Before any modeling claim, four sanity checks: **are the artifacts the size and shape we expect, are the heavy-tailed distributions actually heavy-tailed, what's the join coverage between provider ZIPs and county FIPS, and how does the robust-z null distribution sit relative to the ≥ 5 flagging threshold?**

```{r eda-artifact-shape}
#| label: eda-artifact-shape

artifact_summary <- function(name, df) {
  tibble::tibble(
    artifact     = name,
    rows         = nrow(df),
    cols         = ncol(df),
    n_complete   = sum(stats::complete.cases(df)),
    pct_complete = 100 * n_complete / pmax(nrow(df), 1)
  )
}

dplyr::bind_rows(
  artifact_summary("hcpcs_summary",    hcpcs_sum),
  artifact_summary("state_summary",    state_sum),
  artifact_summary("specialty_summary",specialty_sum),
  artifact_summary("outliers",         outliers),
  artifact_summary("taxonomy",         taxonomy)
) |>
  dplyr::mutate(
    rows         = format(rows, big.mark = ","),
    cols         = format(cols, big.mark = ","),
    n_complete   = format(n_complete, big.mark = ","),
    pct_complete = sprintf("%.1f%%", pct_complete)
  ) |>
  knitr::kable(
    col.names = c("Parquet artifact", "Rows", "Cols",
                  "Complete-case rows", "% complete-case"),
    caption = "Row × column shape and complete-case rate of each prep artifact. Aggregates (hcpcs / state / specialty summary) are dense by construction; the outliers detail table carries some NA in lat/lon (provider ZIP didn't resolve to a Census centroid)."
  )
```

```{r eda-distribution}
#| label: eda-distribution
#| fig-cap: "Three distributional sanity checks. **Top-left** — paid per beneficiary across all (NPI × HCPCS) rows on a log10 scale; the long upper tail is exactly why MAD-based robust z's outperform a mean/SD z-score here. **Top-right** — beneficiary count per row, capped at 5,000 for visibility. **Bottom-left** — peer-group sizes per HCPCS (the denominators behind every robust z); HCPCS with peer_n < 25 are filtered out in prep. **Bottom-right** — robust-z values (those with a finite z) showing the long right tail; the gold dashed line marks the program-integrity ≥ 5 cutoff."
#| fig-width: 10
#| fig-height: 6.2

paid_per_bene_plot <- ggplot(
  outliers |> dplyr::filter(paid_per_bene > 0, !is.na(paid_per_bene)),
  aes(paid_per_bene)
) +
  geom_histogram(fill = "#0d5c63", color = "white", bins = 60) +
  scale_x_log10(labels = scales::label_dollar()) +
  labs(x = "Paid per beneficiary (log10)", y = "Rows", subtitle = "Per-row paid/bene")

bene_plot <- ggplot(
  outliers |> dplyr::filter(beneficiaries > 0, beneficiaries <= 5000),
  aes(beneficiaries)
) +
  geom_histogram(fill = "#0d5c63", color = "white", bins = 60) +
  labs(x = "Beneficiaries (≤ 5,000)", y = "Rows", subtitle = "Per-row beneficiary count")

peer_plot <- ggplot(hcpcs_sum, aes(peer_n)) +
  geom_histogram(fill = "#0d5c63", color = "white", bins = 60) +
  scale_x_log10(labels = scales::label_comma()) +
  labs(x = "Peer providers per HCPCS (log10)", y = "HCPCS codes",
       subtitle = "Peer-group sizes")

z_plot <- ggplot(
  outliers |> dplyr::filter(!is.na(robust_z),
                            robust_z >= -10, robust_z <= 50),
  aes(robust_z)
) +
  geom_histogram(fill = "#0d5c63", color = "white", bins = 60) +
  geom_vline(xintercept = 5, linetype = "dashed",
             color = "#b08a49", linewidth = 0.6) +
  labs(x = "Robust z (clipped to [-10, 50])", y = "Rows",
       subtitle = "Robust-z distribution")

# Use patchwork if available, then gridExtra, then a plain stacked print.
if (requireNamespace("patchwork", quietly = TRUE)) {
  patchwork::wrap_plots(paid_per_bene_plot, bene_plot,
                        peer_plot, z_plot, ncol = 2)
} else if (requireNamespace("gridExtra", quietly = TRUE)) {
  gridExtra::grid.arrange(paid_per_bene_plot, bene_plot,
                          peer_plot, z_plot, ncol = 2)
} else {
  for (p in list(paid_per_bene_plot, bene_plot, peer_plot, z_plot)) print(p)
}
```

```{r eda-geo-coverage}
#| label: eda-geo-coverage

geo_cov <- outliers |>
  dplyr::summarise(
    rows_total      = dplyr::n(),
    rows_with_state = sum(!is.na(state)),
    rows_with_geo   = sum(!is.na(lat) & !is.na(lon)),
    rows_with_spec  = sum(!is.na(specialty_name))
  ) |>
  dplyr::mutate(
    pct_state = 100 * rows_with_state / rows_total,
    pct_geo   = 100 * rows_with_geo   / rows_total,
    pct_spec  = 100 * rows_with_spec  / rows_total
  )

cat(sprintf(paste0(
    "Outlier-table rows total            : %s\n",
    "  with NPPES state                  : %s (%.1f%%)\n",
    "  with NPPES specialty              : %s (%.1f%%)\n",
    "  with ZCTA → lat/lon centroid      : %s (%.1f%%)\n"),
  format(geo_cov$rows_total,      big.mark = ","),
  format(geo_cov$rows_with_state, big.mark = ","), geo_cov$pct_state,
  format(geo_cov$rows_with_spec,  big.mark = ","), geo_cov$pct_spec,
  format(geo_cov$rows_with_geo,   big.mark = ","), geo_cov$pct_geo
))
```

The geo-coverage line is the join-quality check that any geographic claim downstream depends on: a county-level atlas built from a 60% NPPES-state coverage would mean the map is showing 60% of the data, not 100% — and the difference matters for any conclusion about state-level Medicaid policy.

### Overdispersion and zero-inflation diagnostics

Two formal checks before any future modeling on the count-like features. The variance-to-mean ratio (VMR) tests whether Poisson is even on the table — a VMR much greater than 1 means the variance grows faster than the mean and you need negative binomial (or quasi-Poisson). Pairing VMR with the % zero count is what motivates a **zero-inflated NB** rather than a plain NB.

```{r eda-overdispersion}
#| label: eda-overdispersion

vmr <- function(x) {
  x <- x[!is.na(x) & x >= 0]
  if (length(x) == 0 || mean(x) == 0) return(c(mean = NA, var = NA, vmr = NA))
  c(mean = mean(x), var = stats::var(x), vmr = stats::var(x) / mean(x))
}
zero_pct <- function(x) {
  x <- x[!is.na(x)]
  if (!length(x)) return(NA_real_)
  100 * mean(x == 0)
}

vmr_tbl <- rbind(
  paid          = vmr(outliers$paid),
  beneficiaries = vmr(outliers$beneficiaries),
  claims        = vmr(outliers$claims),
  months_active = vmr(outliers$months_active)
)

family_call <- function(v) {
  if (is.na(v)) return("—")
  if (v < 1.2)              "Poisson"
  else if (v < 5)           "Quasi-Poisson / NB"
  else                      "Negative binomial / ZINB"
}

knitr::kable(
  data.frame(
    Variable     = rownames(vmr_tbl),
    Mean         = sprintf("%.1f", vmr_tbl[, "mean"]),
    Variance     = sprintf("%.1f", vmr_tbl[, "var"]),
    VMR          = sprintf("%.1f", vmr_tbl[, "vmr"]),
    `% zeros`    = sprintf("%.2f%%",
                            vapply(list(outliers$paid, outliers$beneficiaries,
                                        outliers$claims, outliers$months_active),
                                   zero_pct, numeric(1))),
    `Family`     = vapply(vmr_tbl[, "vmr"], family_call, character(1)),
    check.names = FALSE
  ),
  caption = "Variance-to-mean ratio (VMR), zero-inflation, and the implied regression family for each count-like feature. Caveat: the outliers parquet is already filtered to robust_z >= 2 OR top-20 spend codes AND beneficiaries >= 11 in prep, so the % zeros reads near 0 by construction. The VMR is still informative because it characterizes the conditional dispersion that any future modeling step will need to absorb."
)
```

### Within-flagged-set selection bias

The outliers parquet is itself a non-random sample of the full DuckDB output (kept rows are those with `robust_z >= 2` OR billed under the top-20 spending HCPCS codes). We can't directly compare against the dropped rows without re-running the 11 GB DuckDB pipeline, but we *can* check whether the **high-z subset** (`robust_z >= 5`, the program-integrity threshold) systematically differs from the **moderate-z subset** (`2 <= robust_z < 5`) on the variables we'd otherwise model — a "what makes the flagged-flagged set different from the merely flagged" check.

```{r eda-within-cohort-bias}
#| label: eda-within-cohort-bias

with_groups <- outliers |>
  dplyr::filter(!is.na(robust_z), robust_z >= 2) |>
  dplyr::mutate(
    z_group = dplyr::if_else(robust_z >= 5, "High z (>= 5)", "Moderate z (2 - 5)")
  )

n_hi  <- sum(with_groups$z_group == "High z (>= 5)")
n_mod <- sum(with_groups$z_group == "Moderate z (2 - 5)")

cont_results <- list(
  beneficiaries  = stats::wilcox.test(beneficiaries  ~ z_group, data = with_groups),
  claims         = stats::wilcox.test(claims         ~ z_group, data = with_groups),
  paid           = stats::wilcox.test(paid           ~ z_group, data = with_groups),
  paid_per_bene  = stats::wilcox.test(paid_per_bene  ~ z_group, data = with_groups),
  months_active  = stats::wilcox.test(months_active  ~ z_group, data = with_groups)
)

state_tab <- table(with_groups$state, with_groups$z_group)
spec_tab  <- table(with_groups$specialty_class, with_groups$z_group)

cat_tests <- list(
  state           = suppressWarnings(stats::chisq.test(state_tab)),
  specialty_class = suppressWarnings(stats::chisq.test(spec_tab))
)

bias_df <- data.frame(
  Variable    = c(names(cont_results), names(cat_tests)),
  Test        = c(rep("Mann-Whitney U", length(cont_results)),
                  rep("Pearson chi-squared", length(cat_tests))),
  Statistic   = c(
    vapply(cont_results, \(r) round(as.numeric(r$statistic), 0), numeric(1)),
    vapply(cat_tests,    \(r) round(as.numeric(r$statistic), 1), numeric(1))
  ),
  `p-value`   = c(
    vapply(cont_results, \(r) format.pval(r$p.value, eps = 1e-300, digits = 2),
           character(1)),
    vapply(cat_tests,    \(r) format.pval(r$p.value, eps = 1e-300, digits = 2),
           character(1))
  ),
  check.names = FALSE
)

cat(sprintf("High-z (>= 5)         : %s rows\nModerate-z (2 - 5)    : %s rows\n\n",
            format(n_hi,  big.mark = ","),
            format(n_mod, big.mark = ",")))
knitr::kable(
  bias_df,
  caption = "Within-flagged-set bias check: do high-z (>= 5) rows differ from moderate-z (2 - 5) rows on the inputs that would feed any peer-comparison model? The 'high-z is just larger providers' null can be ruled out by checking the beneficiary and claim distributions; specialty mix differences expose which clinical areas drive the high-z tail."
)
```

```{r pareto}
#| fig-cap: "Cumulative share of total Medicaid paid dollars, ordered by HCPCS code from highest- to lowest-spending. A steep early curve means a small set of codes drives most public outlay."

hcpcs_totals <- hcpcs_sum |>
  arrange(desc(total_paid)) |>
  mutate(
    rank_pct  = row_number() / n(),
    cum_share = cumsum(total_paid) / sum(total_paid)
  )

ggplot(hcpcs_totals, aes(rank_pct, cum_share)) +
  geom_line(color = "#0d5c63", linewidth = 1) +
  geom_hline(yintercept = 0.8, linetype = "dashed", color = "#b08a49") +
  scale_x_continuous(labels = percent_format()) +
  scale_y_continuous(labels = percent_format()) +
  labs(x = "HCPCS codes ranked by total paid (ascending)",
       y = "Cumulative share of total paid")
```

## Peer-group outlier detection (R)

The DuckDB prep computed a robust z-score per HCPCS using the median-of-absolute-deviations (MAD) scaled by 1.4826. Values ≥ 5 correspond roughly to "5σ from the peer median for this code," which is the threshold CMS Program Integrity uses as a first-pass screen.

```{r top-outliers}
top_outliers <- outliers |>
  filter(robust_z >= 5, beneficiaries >= 30) |>
  arrange(desc(robust_z)) |>
  slice_head(n = 25) |>
  transmute(
    npi, state, specialty_name, hcpcs,
    beneficiaries, paid,
    paid_per_bene       = paid / beneficiaries,
    hcpcs_median_ppb,
    robust_z            = round(robust_z, 1)
  )

knitr::kable(
  top_outliers |>
    mutate(
      paid              = format_money(paid),
      paid_per_bene     = format_money(paid_per_bene),
      hcpcs_median_ppb  = format_money(hcpcs_median_ppb)
    ),
  caption = "Top 25 peer-group outliers (robust z ≥ 5, ≥ 30 beneficiaries)"
)
```

```{r specialty-boxplot}
#| fig-cap: "Per-beneficiary spend by specialty (top 15 by flagged-outlier count). Box spans p25–p75 with median; whiskers are p05 / p95 from the full peer distribution computed in the DuckDB prep step. Gold dots are the individual robust-z ≥ 5 outliers layered on top."

top_specialties <- specialty_sum |>
  arrange(desc(flagged_ge_5)) |>
  slice_head(n = 15) |>
  pull(specialty_name)

box_df <- specialty_sum |>
  filter(specialty_name %in% top_specialties) |>
  mutate(specialty_name = fct_reorder(specialty_name, p50))

jitter_df <- outliers |>
  filter(specialty_name %in% top_specialties, robust_z >= 5) |>
  mutate(specialty_name = factor(specialty_name, levels = levels(box_df$specialty_name)))

ggplot(box_df, aes(specialty_name)) +
  geom_boxplot(
    aes(
      ymin   = p05,  lower  = p25,
      middle = p50,  upper  = p75,
      ymax   = p95
    ),
    stat = "identity",
    fill = "#e6f1f2", color = "#0d5c63", width = 0.7
  ) +
  geom_jitter(data = jitter_df,
              aes(y = paid_per_bene),
              color = "#b08a49", width = 0.2, size = 0.7, alpha = 0.5) +
  scale_y_log10(labels = scales::label_dollar()) +
  coord_flip() +
  labs(x = NULL, y = "Paid per beneficiary (log scale)")
```

## Unsupervised comparison (Python)

A peer-group z-score is simple and auditable but scale-sensitive. An unsupervised model sees the joint distribution of features and can flag records that look unusual across multiple axes at once. We fit an isolation forest on the top 20 HCPCS by total spend, standardize, and score.

### Feature collinearity, briefly

Isolation forests are tree-based and don't inflate standard errors the way a linear model would, but heavy collinearity still costs the ensemble *tree diversity* — the splits become redundant across features that move together. A two-minute check before fitting:

```{python iforest-collinearity}
from sklearn.linear_model import LinearRegression

df = r.outliers.copy()
top_codes = (
    df.groupby("hcpcs")["paid"].sum().sort_values(ascending=False).head(20).index
)
sub = df[df["hcpcs"].isin(top_codes)].copy()
sub = sub.dropna(subset=["paid_per_bene", "beneficiaries", "claims", "months_active"])

feats   = ["paid_per_bene", "beneficiaries", "claims", "months_active"]
X_raw   = sub[feats]

print("Pearson correlation matrix among isolation-forest features:")
print(X_raw.corr(method="pearson").round(2).to_string())

Xs   = StandardScaler().fit_transform(X_raw.to_numpy())
sv   = np.linalg.svd(Xs, compute_uv=False)
cond = sv.max() / sv.min()
print(f"\nCondition number (kappa) of standardized design matrix: {cond:.1f}")
print("Rule of thumb: kappa < 10 clean, 10-30 moderate, > 30 problematic.")

vifs = {}
for j, name in enumerate(feats):
    others = [k for k in range(len(feats)) if k != j]
    r2     = LinearRegression().fit(Xs[:, others], Xs[:, j]).score(Xs[:, others], Xs[:, j])
    vifs[name] = round(1.0 / max(1 - r2, 1e-9), 2)

print("\nApproximate VIF per feature (1 / (1 - R^2_other)):")
for k, v in vifs.items():
    print(f"  {k:<16s} : {v}")
```

What the numbers mean here: `claims` and `beneficiaries` are essentially the same construct measured two ways (each claim covers approximately one beneficiary), so r ≈ 0.95+ and a VIF in the double digits on both is the expected reading — not a bug. `paid_per_bene` is by construction orthogonal to `beneficiaries` (it's paid divided by beneficiaries) and `months_active` carries an independent activity-duration axis. We keep all four features anyway: the isolation forest splits don't *over-count* the redundant pair the way a linear regression would, and dropping `claims` would lose the small but real per-beneficiary multi-claim variation in the long tail. The condition-number print just makes that decision auditable rather than implicit.

### Fitting the isolation forest

```{python iforest}
X = StandardScaler().fit_transform(X_raw.to_numpy())

iso = IsolationForest(
    n_estimators=400, contamination=0.01, random_state=42, n_jobs=-1
)
sub["iforest_score"]    = -iso.fit(X).score_samples(X)   # higher = more anomalous
sub["iforest_flag"]     = iso.predict(X) == -1

print(f"Rows analysed: {len(sub):,}")
print(f"Isolation forest flags: {int(sub['iforest_flag'].sum()):,} ({sub['iforest_flag'].mean()*100:.2f}%)")
print("Overlap with robust-z ≥ 5:",
      int((sub['iforest_flag'] & (sub['robust_z'] >= 5)).sum()))
```

```{python iforest-scatter}
#| fig-cap: "Paid-per-beneficiary vs. beneficiary count for the top 20 HCPCS codes by spend. Isolation-forest flags in gold overlap substantially with — but not identically to — the robust-z rule."

fig, ax = plt.subplots(figsize=(7, 5))
_ = ax.scatter(sub["beneficiaries"], sub["paid_per_bene"],
               c=np.where(sub["iforest_flag"], "#b08a49", "#cccccc"),
               s=10, alpha=0.5)
_ = ax.set(xscale="log", yscale="log",
           xlabel="Unique beneficiaries (log)", ylabel="Paid per beneficiary (log)")
_ = ax.spines[["top", "right"]].set_visible(False)
plt.tight_layout()
plt.show()
```

## Where does the dollar go? A Medicaid category cost atlas

A provider-outlier list is a fraud-review input. Most policy audiences, though, want the opposite question: **which categories of care cost the most, and where do those costs concentrate geographically?** Switching from "who" to "what and where" requires three pieces we've now built upstream in [`prep.py`](prep.py):

1. A rule-based **HCPCS → category** classifier that buckets every procedure, drug, DME, and Medicaid-specific T-code into one of 12 spending categories.
2. A **Census 2020 ZCTA → county** crosswalk so every provider's practice ZIP resolves to a county FIPS.
3. A county × category aggregate with **within-category quintile rankings** so a single color scale reads consistently across panels: darker = this county is in the top 20% of cost-per-beneficiary for *that* category.

These buckets are a mix of clinical areas (Behavioral Health, Dental, Dialysis & ESRD, Prescription Drugs) and service-type groupings that the HCPCS code system itself organizes by (Surgery & Procedures, Radiology, Laboratory, DME, Transportation, LTSS, E&M) — not a true disease taxonomy. Reading the atlas as "categories of Medicaid spend" is more honest than reading it as "diseases": HCPCS doesn't carry diagnosis, so the closest we can get without ICD is spending-category geography.

::: {.callout-note}
## Why county, not HSA or ZIP3?

For a Medicaid cost atlas, counties are the defensible middle path — clean TIGER polygons, HUD/Census maintained crosswalks, and the unit most federal-agency publications (MACPAC, KFF, Urban Institute) use. HSAs (Dartmouth Atlas) are the gold standard for *Medicare* market analysis but were built from Medicare patient flows, so they're slightly miscalibrated for Medicaid. ZIP3 is tempting because it's one `LEFT(zip, 3)` away, but ZIP3 boundaries are USPS logistics geography — a high Modifiable-Areal-Unit-Problem exposure for a spatial-stats story.
:::

```{r cost-atlas-setup}
county_atlas <- arrow::read_parquet(file.path(data_dir, "county_group_atlas.parquet"))
group_totals <- arrow::read_parquet(file.path(data_dir, "group_totals.parquet"))

top_categories <- group_totals |>
  arrange(desc(total_paid)) |>
  pull(disease_group)

# Recompute a within-category decile rank on paid-per-beneficiary. The parquet
# carries quintiles from prep.py; we override with 10 bins here so shading
# shows more granular within-category variation without re-running the
# DuckDB prep. For categories with < 100 billing counties some deciles will
# be tied — the lambda collapses those cleanly.
county_atlas <- county_atlas |>
  group_by(disease_group) |>
  mutate(decile_ppb = dplyr::ntile(paid_per_bene, 10)) |>
  ungroup()
```

```{r group-bar}
#| fig-cap: "Total national Medicaid paid by HCPCS-derived category, 2024. Ranked descending by total paid. 'Other Medical Services' captures 99000-series codes the classifier doesn't yet disambiguate."
#| fig-width: 9
#| fig-height: 4.5

group_totals |>
  mutate(category = fct_reorder(disease_group, total_paid)) |>
  ggplot(aes(category, total_paid)) +
  geom_col(fill = "#756bb1") +
  coord_flip() +
  scale_y_continuous(labels = format_money) +
  labs(x = NULL, y = "Total paid (USD)")
```

```{r county-choropleth}
#| fig-cap: "Interactive county-level cost atlas: Medicaid paid per beneficiary ranked into within-category deciles (darker purple = higher cost relative to other counties for that same category). Each panel is its own choropleth so coloring is comparable within category, not across categories. Use the top-right dropdown to switch across all 12 HCPCS spending categories. Alaska, Hawaii, and territories omitted for contiguous-US framing; grey counties have no Medicaid providers billing that category."

options(tigris_use_cache = TRUE)
counties_sf <- tigris::counties(year = 2020, cb = TRUE, resolution = "5m", progress_bar = FALSE) |>
  dplyr::select(GEOID, geometry) |>
  dplyr::filter(!stringr::str_detect(GEOID, "^(02|15|60|66|69|72|78)")) |>
  sf::st_transform(4326)

purples <- grDevices::colorRampPalette(c("#fcfbfd", "#3f007d"))(10)
pal <- leaflet::colorFactor(purples, domain = factor(1:10))

atlas_sub <- county_atlas |>
  filter(disease_group %in% top_categories)

lmap <- leaflet::leaflet(width = "100%", height = "620px") |>
  leaflet::addProviderTiles(
    leaflet::providers$CartoDB.PositronNoLabels,
    options = leaflet::providerTileOptions(opacity = 0.75)
  ) |>
  leaflet::setView(lng = -96, lat = 38, zoom = 4) |>
  leaflet::addPolygons(
    data        = counties_sf,
    fillColor   = "#e5e5e5",
    fillOpacity = 0.6,
    color       = "white",
    weight      = 0.3,
    smoothFactor = 0.5,
    options     = leaflet::pathOptions(interactive = FALSE)
  )

for (g in top_categories) {
  dat <- counties_sf |>
    dplyr::inner_join(
      atlas_sub |> dplyr::filter(disease_group == g),
      by = c("GEOID" = "county_fips")
    )
  lmap <- lmap |>
    leaflet::addPolygons(
      data         = dat,
      group        = g,
      fillColor    = ~pal(factor(decile_ppb)),
      fillOpacity  = 0.88,
      color        = "white",
      weight       = 0.3,
      smoothFactor = 0.5,
      highlightOptions = leaflet::highlightOptions(
        weight       = 2,
        color        = "#333",
        fillOpacity  = 0.95,
        bringToFront = TRUE
      ),
      popup = ~paste0(
        "<b>", county_name, ", ", state, "</b><br/>",
        "<em>", disease_group, "</em><br/>",
        "Paid / beneficiary: $",
        format(round(paid_per_bene), big.mark = ","), "<br/>",
        "Total paid: $",
        format(round(total_paid / 1e3), big.mark = ","), "K<br/>",
        "Providers: ", providers_n, "<br/>",
        "Within-category decile: ", decile_ppb, " of 10"
      )
    )
}

categories_json <- jsonlite::toJSON(top_categories)
dropdown_js <- sprintf("
function(el, x) {
  var map = this;
  var categories = %s;
  var initial = categories[0];

  function showOnly(picked) {
    categories.forEach(function(g) {
      var fg = map.layerManager.getLayerGroup(g, false);
      if (!fg) return;
      if (g === picked) {
        if (!map.hasLayer(fg)) fg.addTo(map);
      } else {
        if (map.hasLayer(fg)) map.removeLayer(fg);
      }
    });
  }
  showOnly(initial);

  var ctrl = L.control({position: 'topright'});
  ctrl.onAdd = function() {
    var div = L.DomUtil.create('div', 'leaflet-bar disease-group-control');
    div.style.background = 'white';
    div.style.padding = '8px 10px';
    div.style.boxShadow = '0 1px 4px rgba(0,0,0,0.2)';
    div.style.font = '12px/1.3 -apple-system, BlinkMacSystemFont, \"Segoe UI\", sans-serif';

    var label = document.createElement('label');
    label.textContent = 'Category';
    label.style.cssText = 'display:block;font-weight:600;margin-bottom:4px;color:#333;';

    var sel = document.createElement('select');
    sel.style.cssText = 'padding:4px 6px;font-size:12px;border:1px solid #ccc;border-radius:3px;background:white;min-width:240px;';
    categories.forEach(function(g) {
      var opt = document.createElement('option');
      opt.value = g;
      opt.textContent = g;
      if (g === initial) opt.selected = true;
      sel.appendChild(opt);
    });
    sel.addEventListener('change', function(e) { showOnly(e.target.value); });

    div.appendChild(label);
    div.appendChild(sel);
    L.DomEvent.disableClickPropagation(div);
    L.DomEvent.disableScrollPropagation(div);
    return div;
  };
  ctrl.addTo(map);
}
", categories_json)

lmap |>
  htmlwidgets::onRender(dropdown_js) |>
  leaflet::addLegend(
    position = "bottomleft",
    colors   = purples,
    labels   = c("1 (lowest)", "2", "3", "4", "5", "6", "7", "8", "9", "10 (highest)"),
    title    = "Decile of paid<br/>per beneficiary",
    opacity  = 0.85
  )
```

## Category explorer

The choropleth shows *where* a category is expensive. The three views below answer three follow-up questions policy audiences almost always ask next: **what HCPCS codes drive the spend inside each category, how spatially concentrated is that spend across counties, and which states sit at the top of the ladder?** Every chunk iterates over all 12 categories — no cherry-picking — so the honest shape of Medicaid dollar flow is visible in one sweep.

```{r hcpcs-classifier}
#| code-fold: show
#| code-summary: "HCPCS → category classifier (mirrors the DuckDB CASE in prep.py)"

classify_hcpcs <- function(code) {
  num    <- suppressWarnings(as.integer(code))
  is_cpt <- !is.na(num) & nchar(code) == 5
  dplyr::case_when(
    stringr::str_starts(code, "J")                               ~ "Prescription Drugs & Biologics",
    stringr::str_starts(code, "D")                               ~ "Dental",
    stringr::str_starts(code, "H")                               ~ "Behavioral Health",
    stringr::str_starts(code, "E") |
      stringr::str_starts(code, "K") |
      stringr::str_starts(code, "L")                             ~ "Durable Medical Equipment & Supplies",
    stringr::str_detect(code, "^T(20|21|22|51)") |
      stringr::str_starts(code, "A04")                           ~ "Transportation",
    stringr::str_starts(code, "T")                               ~ "Long-Term Services & Supports",
    code %in% c("G0257","G0323","G0324","G0325","G0326","G0327") ~ "Dialysis & ESRD",
    is_cpt & num >= 90791 & num <= 90899                         ~ "Behavioral Health",
    is_cpt & num >= 90935 & num <= 90999                         ~ "Dialysis & ESRD",
    is_cpt & num >= 99201 & num <= 99499                         ~ "Evaluation & Management",
    is_cpt & num >= 10021 & num <= 69990                         ~ "Surgery & Outpatient Procedures",
    is_cpt & num >= 70010 & num <= 79999                         ~ "Radiology & Imaging",
    is_cpt & num >= 80047 & num <= 89398                         ~ "Laboratory & Diagnostics",
    TRUE                                                         ~ "Other Medical Services"
  )
}

hcpcs_by_category <- hcpcs_sum |>
  mutate(category = classify_hcpcs(hcpcs)) |>
  filter(category %in% top_categories)
```

### What codes drive each category?

```{r category-hcpcs-mix}
#| fig-cap: "Top 5 HCPCS codes by total Medicaid paid within each category. The codes carrying the weight are very different per category — J-codes (specialty infusibles and injectables) dominate the drug bucket; Transportation collapses onto A0/T20-series non-emergency transport; LTSS is driven by T-series personal-care and case-management codes; E&M is dominated by the 99213/99214 office-visit pair; a single periodic-exam code (D0120) and a pair of restoration codes drive most of Dental."
#| fig-width: 11
#| fig-height: 11

top_hcpcs_per_category <- hcpcs_by_category |>
  dplyr::group_by(category) |>
  dplyr::arrange(dplyr::desc(total_paid), .by_group = TRUE) |>
  dplyr::slice_head(n = 5) |>
  dplyr::ungroup() |>
  dplyr::mutate(
    category = factor(category, levels = top_categories),
    row_id   = dplyr::row_number()
  )

ggplot(top_hcpcs_per_category,
       aes(x = stats::reorder(paste(hcpcs, row_id), total_paid),
           y = total_paid)) +
  geom_col(fill = "#54278f") +
  geom_text(aes(label = format_money(total_paid)),
            hjust = -0.08, size = 3, color = "#333") +
  facet_wrap(~ category, scales = "free", ncol = 3) +
  scale_x_discrete(labels = \(x) sub(" \\d+$", "", x)) +
  scale_y_continuous(labels = format_money,
                     expand = expansion(mult = c(0, 0.22))) +
  coord_flip() +
  labs(x = NULL, y = "Total paid (all billing providers)")
```

### How spatially concentrated is each category?

```{r category-concentration}
#| fig-cap: "Gini coefficient of county Medicaid paid within each category, and the share of that category's total spend captured by just the top 10 counties. A Gini near 1 means spend collapses onto a handful of counties — expected for sparse, high-acuity categories like Dialysis & ESRD; Gini near 0 is evenly diffused public outlay, like LTSS and E&M."
#| fig-width: 9
#| fig-height: 5.5

gini <- function(x) {
  x <- sort(x[!is.na(x) & x >= 0])
  n <- length(x)
  if (n <= 1 || sum(x) == 0) return(NA_real_)
  2 * sum(seq_len(n) * x) / (n * sum(x)) - (n + 1) / n
}

spatial_conc <- county_atlas |>
  filter(disease_group %in% top_categories) |>
  group_by(category = disease_group) |>
  summarise(
    counties_n    = dplyr::n(),
    gini_paid     = gini(total_paid),
    top10_share   = sum(sort(total_paid, decreasing = TRUE)[seq_len(min(10, dplyr::n()))],
                        na.rm = TRUE) / sum(total_paid, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(gini_paid))

spatial_conc |>
  mutate(category = fct_reorder(category, gini_paid)) |>
  ggplot(aes(category, gini_paid)) +
  geom_col(fill = "#756bb1") +
  geom_text(aes(label = sprintf("Gini %.2f · top-10 share %s",
                                gini_paid,
                                scales::percent(top10_share, accuracy = 1))),
            hjust = -0.05, size = 3.1, color = "#333") +
  coord_flip(clip = "off") +
  scale_y_continuous(limits = c(0, 1.05),
                     labels = scales::number_format(accuracy = 0.01),
                     expand = expansion(mult = c(0, 0))) +
  labs(x = NULL, y = "Gini of county Medicaid paid (within category)") +
  theme(plot.margin = margin(5, 110, 5, 5))
```

### Which states lead each category?

```{r category-state-hotspots}
state_hotspots <- county_atlas |>
  filter(disease_group %in% top_categories, !is.na(state)) |>
  group_by(state, category = disease_group) |>
  summarise(
    state_paid = sum(total_paid,  na.rm = TRUE),
    state_bene = sum(total_bene,  na.rm = TRUE),
    counties   = dplyr::n(),
    .groups = "drop"
  ) |>
  filter(state_bene >= 1000) |>
  mutate(state_ppb = state_paid / state_bene) |>
  group_by(category) |>
  arrange(desc(state_ppb), .by_group = TRUE) |>
  slice_head(n = 5) |>
  ungroup() |>
  mutate(category = factor(category, levels = top_categories)) |>
  arrange(category, desc(state_ppb))

knitr::kable(
  state_hotspots |>
    transmute(
      Category             = category,
      State                = state,
      Counties             = counties,
      `Total paid`         = format_money(state_paid),
      Beneficiaries        = format(state_bene, big.mark = ","),
      `$ per beneficiary`  = format_money(state_ppb)
    ),
  caption = "Top 5 states by Medicaid paid-per-beneficiary within each category (states with at least 1,000 beneficiaries in that category)."
)
```

## State-level intensity

```{r state-intensity}
#| fig-cap: "Peer-group-excess Medicaid spending by state. Height is the sum of paid dollars above each HCPCS's peer median, aggregated at prep time across all providers billing in that state."

state_excess <- state_sum |>
  arrange(desc(excess_paid)) |>
  slice_head(n = 25)

ggplot(state_excess, aes(reorder(state, excess_paid), excess_paid)) +
  geom_col(fill = "#0d5c63") +
  coord_flip() +
  scale_y_continuous(labels = format_money) +
  labs(x = NULL, y = "Paid above HCPCS peer median ($)")
```

## Conclusions

The seven-step pipeline (DuckDB aggregate → robust z → NPPES enrichment → ZCTA-county join → category classifier → atlas → explorer) supports five conclusions that each tie back to a specific diagnostic on this page:

- **Heavy tails are the regime, not the exception.** The paid-per-beneficiary distribution in [`eda-distribution`](#cell-fig-eda-distribution) is log-skewed across four orders of magnitude. That's why a mean/SD z-score is unusable here: the SD is dominated by one or two mega-providers and every other observation looks "normal" by comparison. The MAD-based robust z is invariant to those tail rows, which is why the same threshold (≥ 5) generalizes across HCPCS codes with very different peer-group sizes.
- **Rule-based and unsupervised flags agree on the high-confidence set.** The overlap between robust-z ≥ 5 and the isolation-forest 1%-contamination flag is the set a Program Integrity team would prioritize. Disagreement between the two is the *interesting* set — it's where a tail-only metric (robust z) and a joint-distribution metric (isolation forest) see different things, which is usually a signal worth a second human look.
- **Collinearity didn't break the isolation forest, but it constrained tree diversity.** The condition number on the four standardized features lands above the rule-of-thumb threshold because `claims` and `beneficiaries` are nearly identical by construction. We kept both for interpretability — but the cost is that some trees in the ensemble split on essentially the same axis. A leaner three-feature variant (drop `claims`) would produce a slightly more diverse forest at the cost of a slightly less interpretable feature set; for a fraud-review prompt list, that tradeoff goes the way we chose.
- **Categories are an HCPCS lens, not a disease lens.** The 12-bucket atlas is honest about what it is: a procedure-code taxonomy projected onto county geography, not a disease taxonomy. Reading the choropleth for "where Medicaid spends money on Behavioral Health procedures" is supportable; reading it for "where Medicaid pays for treating depression" is not — depression doesn't have a HCPCS code, only the *services* delivered for depression do. The Gini concentration plot is a useful tie-breaker: categories with high Gini (Dialysis & ESRD, parts of DME) genuinely cluster geographically; categories with low Gini (LTSS, E&M) are diffuse across the country and shouldn't be read as having geographic "hotspots" at all.
- **Geography is the start of an interpretation, never the end.** State-level Medicaid rate schedules, MCO contracting differences, and referral-network composition can all produce county-level patterns that look like billing outliers but are artifacts of policy. The state hotspot tables in the explorer make those patterns visible; deciding which are actionable requires a payer-policy lens this notebook does not pretend to provide.

## Caveats and what would change the picture

- **Single year, no temporal control.** The dataset is the most recent monthly drop; nothing here separates persistent outliers from transient spikes (COVID-era catch-up billing, one-off audits, policy shocks). Rolling 3-month windows would.
- **Provider geocoding gaps shape the map.** The geo-coverage diagnostic in EDA shows the share of outlier rows missing a state or ZCTA centroid; counties whose providers have unresolved addresses are *under-represented* on the choropleth, not just absent. The grey baseline layer makes "no data" visible but doesn't distinguish "no providers billing this category" from "providers exist but didn't geocode."
- **HCPCS coverage is sparse for some categories.** Dialysis & ESRD is billed in roughly 500 counties; the within-category decile shading there is meaningful but bins are sparse, so the visual difference between deciles 4 and 5 is less stable than for densely-covered categories like E&M.

## Next steps

- **Specialty-adjusted peer groups.** Today the peer group is "everyone billing this HCPCS"; tighter peer groups (same specialty, similar patient panel size) will surface a different, higher-quality signal — especially in categories where the peer pool mixes radically different practice types.
- **Time-aware outlier detection.** Repeat the analysis on rolling 3-month windows to separate persistent outliers from transient spikes; the same robust-z framework applies, the only change is the partitioning column in the DuckDB prep.
- **SDoH overlay.** Merge state-level CDC PLACES / ACS indicators to test whether residual excess spending tracks community need vs. billing behavior — a health-equity stress test for the flagged list. The county-FIPS join already exists; adding a tract-level layer is a one-join extension.

---

::: {.callout-tip}
### Reading this as code samples
Every cell above is the real executed code — not screenshots. Use the **Code** dropdown at the top-right to toggle visibility, copy blocks, or view the [raw `.qmd` source](https://github.com/paulinedmdf/paulinadelmundomd/blob/main/projects/02-medicaid-outliers/analysis.qmd) on GitHub. The one-shot DuckDB prep step lives in [`prep.py`](https://github.com/paulinedmdf/paulinadelmundomd/blob/main/projects/02-medicaid-outliers/prep.py).
:::
