---
title: "Outlier Detection in Medicaid Provider Spending"
subtitle: "Peer-group robust z-scores, isolation forests, and a geographic lens on provider-level Medicaid claims"
author: "Paulina Del Mundo Del Fierro, MD, MPH"
date: last-modified
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-location: right
    code-tools: true
    code-copy: true
    code-fold: false
    code-overflow: wrap
    number-sections: true
    fig-responsive: true
    fig-width: 8
    fig-height: 5
    embed-resources: true
    include-in-header:
      text: |
        <script>
          window.va = window.va || function () { (window.vaq = window.vaq || []).push(arguments); };
        </script>
        <script defer src="/_vercel/insights/script.js"></script>
execute:
  echo: true
  warning: false
  message: false
  freeze: auto
---

## Why this analysis

Medicaid is the single largest source of health coverage for low-income Americans — and its claims data, released monthly by HHS, is a rare window into how public dollars flow through the US provider network. The question for any physician-scientist looking at this data is not *"who spends the most"* (large systems dominate by construction) but *"who spends atypically, given their peers?"* That's the signal that drives fraud-and-abuse reviews, network-integrity audits, and honest comparative-effectiveness work alike.

This notebook does three things end-to-end:

1. **Streams the 11 GB Medicaid Provider Spending CSV with DuckDB**, aggregates to the NPI × HCPCS level, and flags peer-group outliers using a median/MAD-based robust z-score — resistant to the heavy tails claims data is famous for.
2. **Joins NPPES, NUCC taxonomy, and US Census ZCTA centroids** so every flagged provider carries an interpretable specialty and a map-able location. Nothing in the analysis requires cloud infrastructure — everything runs on a laptop against committed Parquet artifacts.
3. **Contrasts a deterministic rule (robust z) against an unsupervised model (isolation forest)** on the same cohort, in R and Python respectively. Agreement between the two is the real outlier story; disagreement is where a physician-scientist earns their keep.

::: {.callout-note}
## Data sources
- **Medicaid Provider Spending** — HHS Open Data Portal, accessed 2026-04-21. <https://opendata.hhs.gov/datasets/medicaid-provider-spending/>
- **NPPES Downloadable File** — Centers for Medicare & Medicaid Services. <https://download.cms.gov/nppes/NPI_Files.html>
- **NUCC Health Care Provider Taxonomy** — National Uniform Claim Committee. <https://taxonomy.nucc.org/>
- **ZCTA Gazetteer (2020)** — US Census Bureau. <https://www.census.gov/geographies/reference-files/time-series/geo/gazetteer-files.html>

Raw files stay local (11 GB + 45 GB NPPES); the [`prep.py`](prep.py) script runs one DuckDB pipeline to produce three committed Parquets this notebook reads.
:::

---

## Setup

```{r setup-r}
suppressPackageStartupMessages({
  library(arrow)
  library(dplyr)
  library(ggplot2)
  library(tidyr)
  library(stringr)
  library(forcats)
  library(scales)
  library(sf)
  library(leaflet)
  library(reticulate)
})

reticulate::py_require(c("pandas", "numpy", "scikit-learn", "matplotlib"))

theme_set(theme_minimal(base_size = 12))

data_dir     <- here::here("projects", "02-medicaid-outliers", "data", "processed")
outliers     <- arrow::read_parquet(file.path(data_dir, "outliers.parquet"))
hcpcs_sum    <- arrow::read_parquet(file.path(data_dir, "hcpcs_summary.parquet"))
state_sum    <- arrow::read_parquet(file.path(data_dir, "state_summary.parquet"))
specialty_sum<- arrow::read_parquet(file.path(data_dir, "specialty_summary.parquet"))
taxonomy     <- arrow::read_parquet(file.path(data_dir, "taxonomy.parquet"))

format_money <- function(x) scales::label_dollar(accuracy = 1, scale_cut = scales::cut_short_scale())(x)
```

```{python setup-py}
import pandas as pd
import numpy as np
import matplotlib.pyplot as plt
from sklearn.ensemble import IsolationForest
from sklearn.preprocessing import StandardScaler

plt.rcParams.update({"figure.dpi": 110})
```

## Cohort sketch

```{r cohort-sketch}
summary_tbl <- tibble::tibble(
  hcpcs_n       = nrow(hcpcs_sum),
  providers_n   = outliers |> dplyr::pull(npi) |> dplyr::n_distinct(),
  total_paid    = sum(hcpcs_sum$total_paid, na.rm = TRUE),
  total_bene    = sum(hcpcs_sum$total_bene, na.rm = TRUE),
  flagged_ge_5  = sum(outliers$robust_z >= 5, na.rm = TRUE)
)

knitr::kable(
  summary_tbl |>
    mutate(total_paid = format_money(total_paid),
           across(where(is.numeric), \(x) format(round(x, 2), big.mark = ","))),
  caption = "Cohort sketch from DuckDB prep outputs"
)
```

## Pareto: where does the dollar concentrate?

```{r pareto}
#| fig-cap: "Cumulative share of total Medicaid paid dollars, ordered by HCPCS code from highest- to lowest-spending. A steep early curve means a small set of codes drives most public outlay."

hcpcs_totals <- hcpcs_sum |>
  arrange(desc(total_paid)) |>
  mutate(
    rank_pct  = row_number() / n(),
    cum_share = cumsum(total_paid) / sum(total_paid)
  )

ggplot(hcpcs_totals, aes(rank_pct, cum_share)) +
  geom_line(color = "#0d5c63", linewidth = 1) +
  geom_hline(yintercept = 0.8, linetype = "dashed", color = "#b08a49") +
  scale_x_continuous(labels = percent_format()) +
  scale_y_continuous(labels = percent_format()) +
  labs(x = "HCPCS codes ranked by total paid (ascending)",
       y = "Cumulative share of total paid")
```

## Peer-group outlier detection (R)

The DuckDB prep computed a robust z-score per HCPCS using the median-of-absolute-deviations (MAD) scaled by 1.4826. Values ≥ 5 correspond roughly to "5σ from the peer median for this code," which is the threshold CMS Program Integrity uses as a first-pass screen.

```{r top-outliers}
top_outliers <- outliers |>
  filter(robust_z >= 5, beneficiaries >= 30) |>
  arrange(desc(robust_z)) |>
  slice_head(n = 25) |>
  transmute(
    npi, state, specialty_name, hcpcs,
    beneficiaries, paid,
    paid_per_bene       = paid / beneficiaries,
    hcpcs_median_ppb,
    robust_z            = round(robust_z, 1)
  )

knitr::kable(
  top_outliers |>
    mutate(
      paid              = format_money(paid),
      paid_per_bene     = format_money(paid_per_bene),
      hcpcs_median_ppb  = format_money(hcpcs_median_ppb)
    ),
  caption = "Top 25 peer-group outliers (robust z ≥ 5, ≥ 30 beneficiaries)"
)
```

```{r specialty-boxplot}
#| fig-cap: "Per-beneficiary spend by specialty (top 15 by flagged-outlier count). Box spans p25–p75 with median; whiskers are p05 / p95 from the full peer distribution computed in the DuckDB prep step. Gold dots are the individual robust-z ≥ 5 outliers layered on top."

top_specialties <- specialty_sum |>
  arrange(desc(flagged_ge_5)) |>
  slice_head(n = 15) |>
  pull(specialty_name)

box_df <- specialty_sum |>
  filter(specialty_name %in% top_specialties) |>
  mutate(specialty_name = fct_reorder(specialty_name, p50))

jitter_df <- outliers |>
  filter(specialty_name %in% top_specialties, robust_z >= 5) |>
  mutate(specialty_name = factor(specialty_name, levels = levels(box_df$specialty_name)))

ggplot(box_df, aes(specialty_name)) +
  geom_boxplot(
    aes(
      ymin   = p05,  lower  = p25,
      middle = p50,  upper  = p75,
      ymax   = p95
    ),
    stat = "identity",
    fill = "#e6f1f2", color = "#0d5c63", width = 0.7
  ) +
  geom_jitter(data = jitter_df,
              aes(y = paid_per_bene),
              color = "#b08a49", width = 0.2, size = 0.7, alpha = 0.5) +
  scale_y_log10(labels = scales::label_dollar()) +
  coord_flip() +
  labs(x = NULL, y = "Paid per beneficiary (log scale)")
```

## Unsupervised comparison (Python)

A peer-group z-score is simple and auditable but scale-sensitive. An unsupervised model sees the joint distribution of features and can flag records that look unusual across multiple axes at once. We fit an isolation forest on the top 20 HCPCS by total spend, standardize, and score.

```{python iforest}
df = r.outliers.copy()

top_codes = (
    df.groupby("hcpcs")["paid"].sum().sort_values(ascending=False).head(20).index
)
sub = df[df["hcpcs"].isin(top_codes)].copy()
sub = sub.dropna(subset=["paid_per_bene", "beneficiaries", "claims", "months_active"])

X = sub[["paid_per_bene", "beneficiaries", "claims", "months_active"]].to_numpy()
X = StandardScaler().fit_transform(X)

iso = IsolationForest(
    n_estimators=400, contamination=0.01, random_state=42, n_jobs=-1
)
sub["iforest_score"]    = -iso.fit(X).score_samples(X)   # higher = more anomalous
sub["iforest_flag"]     = iso.predict(X) == -1

print(f"Rows analysed: {len(sub):,}")
print(f"Isolation forest flags: {int(sub['iforest_flag'].sum()):,} ({sub['iforest_flag'].mean()*100:.2f}%)")
print("Overlap with robust-z ≥ 5:",
      int((sub['iforest_flag'] & (sub['robust_z'] >= 5)).sum()))
```

```{python iforest-scatter}
#| fig-cap: "Paid-per-beneficiary vs. beneficiary count for the top 20 HCPCS codes by spend. Isolation-forest flags in gold overlap substantially with — but not identically to — the robust-z rule."

fig, ax = plt.subplots(figsize=(7, 5))
_ = ax.scatter(sub["beneficiaries"], sub["paid_per_bene"],
               c=np.where(sub["iforest_flag"], "#b08a49", "#cccccc"),
               s=10, alpha=0.5)
_ = ax.set(xscale="log", yscale="log",
           xlabel="Unique beneficiaries (log)", ylabel="Paid per beneficiary (log)")
_ = ax.spines[["top", "right"]].set_visible(False)
plt.tight_layout()
plt.show()
```

## Where does the dollar go? A disease-group cost atlas

A provider-outlier list is a fraud-review input. Most policy audiences, though, want the opposite question: **which disease categories cost the most, and where do those costs concentrate geographically?** Switching from "who" to "what and where" requires three pieces we've now built upstream in [`prep.py`](prep.py):

1. A rule-based **HCPCS → category** classifier that buckets every procedure, drug, DME, and Medicaid-specific T-code into one of 12 clinical or service groups.
2. A **Census 2020 ZCTA → county** crosswalk so every provider's practice ZIP resolves to a county FIPS.
3. A county × group aggregate with **within-group quintile rankings** so a single color scale reads consistently across panels: darker = this county is in the top 20% of cost-per-beneficiary for *that* group.

Four of the 12 buckets are **service categories** rather than disease areas — Transportation (Medicaid NEMT), Long-Term Services & Supports, Evaluation & Management, and the Other Medical Services catch-all. They carry real dollars, so the atlas dropdown keeps them but groups them separately from the eight **disease-focused** buckets. Read the disease panels as clinical cost geography; read the service panels as "where does Medicaid spend on non-clinical support." The bar chart below puts both kinds side by side so the split is explicit.

::: {.callout-note}
## Why county, not HSA or ZIP3?

For a Medicaid cost atlas, counties are the defensible middle path — clean TIGER polygons, HUD/Census maintained crosswalks, and the unit most federal-agency publications (MACPAC, KFF, Urban Institute) use. HSAs (Dartmouth Atlas) are the gold standard for *Medicare* market analysis but were built from Medicare patient flows, so they're slightly miscalibrated for Medicaid. ZIP3 is tempting because it's one `LEFT(zip, 3)` away, but ZIP3 boundaries are USPS logistics geography — a high Modifiable-Areal-Unit-Problem exposure for a spatial-stats story.
:::

```{r cost-atlas-setup}
county_atlas <- arrow::read_parquet(file.path(data_dir, "county_group_atlas.parquet"))
group_totals <- arrow::read_parquet(file.path(data_dir, "group_totals.parquet"))

service_buckets <- c(
  "Transportation",
  "Long-Term Services & Supports",
  "Evaluation & Management",
  "Other Medical Services"
)

top_groups <- group_totals |>
  filter(!disease_group %in% service_buckets) |>
  arrange(desc(total_paid)) |>
  slice_head(n = 6) |>
  pull(disease_group)

top_service_groups <- group_totals |>
  filter(disease_group %in% service_buckets) |>
  arrange(desc(total_paid)) |>
  pull(disease_group)

atlas_groups <- c(top_groups, top_service_groups)
```

```{r group-bar}
#| fig-cap: "Total national Medicaid paid by group, 2024. Teal bars are the disease-focused groups that populate the atlas dropdown; gold bars are service categories kept for provenance but excluded from the choropleth because paid-per-beneficiary for transport, LTSS, E&M, and the classifier's catch-all isn't a defensible 'disease cost' metric."
#| fig-width: 9
#| fig-height: 4.5

group_totals |>
  mutate(
    kind = if_else(disease_group %in% service_buckets,
                   "Service category", "Disease-focused"),
    disease_group = fct_reorder(disease_group, total_paid)
  ) |>
  ggplot(aes(disease_group, total_paid, fill = kind)) +
  geom_col() +
  scale_fill_manual(values = c("Service category" = "#b08a49",
                               "Disease-focused"  = "#0d5c63")) +
  coord_flip() +
  scale_y_continuous(labels = format_money) +
  labs(x = NULL, y = "Total paid (USD)", fill = NULL) +
  theme(legend.position = "top")
```

```{r county-choropleth}
#| fig-cap: "Interactive county-level cost atlas: Medicaid paid per beneficiary ranked into within-group quintiles (darker purple = higher cost relative to other counties for that same group). Each panel is its own choropleth so coloring is comparable within group, not across groups. The top-right dropdown groups options as *Disease areas* (top 6 by spend) and *Service categories* (NEMT, LTSS, E&M, and the classifier catch-all). Alaska, Hawaii, and territories omitted for contiguous-US framing; grey counties have no Medicaid providers billing that category."

options(tigris_use_cache = TRUE)
counties_sf <- tigris::counties(year = 2020, cb = TRUE, resolution = "5m", progress_bar = FALSE) |>
  dplyr::select(GEOID, geometry) |>
  dplyr::filter(!stringr::str_detect(GEOID, "^(02|15|60|66|69|72|78)")) |>
  sf::st_transform(4326)

purples <- c("#f2f0f7", "#cbc9e2", "#9e9ac8", "#756bb1", "#54278f")
pal <- leaflet::colorFactor(purples, domain = factor(1:5))

atlas_sub <- county_atlas |>
  filter(disease_group %in% atlas_groups)

lmap <- leaflet::leaflet(width = "100%", height = "620px") |>
  leaflet::addProviderTiles(
    leaflet::providers$CartoDB.PositronNoLabels,
    options = leaflet::providerTileOptions(opacity = 0.75)
  ) |>
  leaflet::setView(lng = -96, lat = 38, zoom = 4) |>
  leaflet::addPolygons(
    data        = counties_sf,
    fillColor   = "#e5e5e5",
    fillOpacity = 0.6,
    color       = "white",
    weight      = 0.3,
    smoothFactor = 0.5,
    options     = leaflet::pathOptions(interactive = FALSE)
  )

for (g in atlas_groups) {
  dat <- counties_sf |>
    dplyr::inner_join(
      atlas_sub |> dplyr::filter(disease_group == g),
      by = c("GEOID" = "county_fips")
    )
  lmap <- lmap |>
    leaflet::addPolygons(
      data         = dat,
      group        = g,
      fillColor    = ~pal(factor(quintile_ppb)),
      fillOpacity  = 0.88,
      color        = "white",
      weight       = 0.3,
      smoothFactor = 0.5,
      highlightOptions = leaflet::highlightOptions(
        weight       = 2,
        color        = "#333",
        fillOpacity  = 0.95,
        bringToFront = TRUE
      ),
      popup = ~paste0(
        "<b>", county_name, ", ", state, "</b><br/>",
        "<em>", disease_group, "</em><br/>",
        "Paid / beneficiary: $",
        format(round(paid_per_bene), big.mark = ","), "<br/>",
        "Total paid: $",
        format(round(total_paid / 1e3), big.mark = ","), "K<br/>",
        "Providers: ", providers_n, "<br/>",
        "Within-group quintile: ", quintile_ppb, " of 5"
      )
    )
}

dropdown_payload <- jsonlite::toJSON(
  list(
    list(label = "Disease areas",      items = as.list(top_groups)),
    list(label = "Service categories", items = as.list(top_service_groups))
  ),
  auto_unbox = TRUE
)
dropdown_js <- sprintf("
function(el, x) {
  var map = this;
  var optgroups = %s;
  var allItems = [];
  optgroups.forEach(function(g) { allItems = allItems.concat(g.items); });
  var initial = allItems[0];

  function showOnly(picked) {
    allItems.forEach(function(g) {
      var fg = map.layerManager.getLayerGroup(g, false);
      if (!fg) return;
      if (g === picked) {
        if (!map.hasLayer(fg)) fg.addTo(map);
      } else {
        if (map.hasLayer(fg)) map.removeLayer(fg);
      }
    });
  }
  showOnly(initial);

  var ctrl = L.control({position: 'topright'});
  ctrl.onAdd = function() {
    var div = L.DomUtil.create('div', 'leaflet-bar disease-group-control');
    div.style.background = 'white';
    div.style.padding = '8px 10px';
    div.style.boxShadow = '0 1px 4px rgba(0,0,0,0.2)';
    div.style.font = '12px/1.3 -apple-system, BlinkMacSystemFont, \"Segoe UI\", sans-serif';

    var label = document.createElement('label');
    label.textContent = 'Category';
    label.style.cssText = 'display:block;font-weight:600;margin-bottom:4px;color:#333;';

    var sel = document.createElement('select');
    sel.style.cssText = 'padding:4px 6px;font-size:12px;border:1px solid #ccc;border-radius:3px;background:white;min-width:240px;';
    optgroups.forEach(function(og) {
      var groupEl = document.createElement('optgroup');
      groupEl.label = og.label;
      og.items.forEach(function(g) {
        var opt = document.createElement('option');
        opt.value = g;
        opt.textContent = g;
        if (g === initial) opt.selected = true;
        groupEl.appendChild(opt);
      });
      sel.appendChild(groupEl);
    });
    sel.addEventListener('change', function(e) { showOnly(e.target.value); });

    div.appendChild(label);
    div.appendChild(sel);
    L.DomEvent.disableClickPropagation(div);
    L.DomEvent.disableScrollPropagation(div);
    return div;
  };
  ctrl.addTo(map);
}
", dropdown_payload)

lmap |>
  htmlwidgets::onRender(dropdown_js) |>
  leaflet::addLegend(
    position = "bottomleft",
    colors   = purples,
    labels   = c("1 (lowest)", "2", "3", "4", "5 (highest)"),
    title    = "Quintile of paid<br/>per beneficiary",
    opacity  = 0.85
  )
```

## Disease-area explorer

The choropleth shows *where* a disease area is expensive. The three views below answer three follow-up questions policy audiences almost always ask next: **what drives the spend inside each disease area, how spatially concentrated is that spend, and which states sit at the top of the ladder?**

```{r hcpcs-classifier}
#| code-fold: show
#| code-summary: "HCPCS → group classifier (mirrors the DuckDB CASE in prep.py)"

classify_hcpcs <- function(code) {
  num    <- suppressWarnings(as.integer(code))
  is_cpt <- !is.na(num) & nchar(code) == 5
  dplyr::case_when(
    stringr::str_starts(code, "J")                               ~ "Prescription Drugs & Biologics",
    stringr::str_starts(code, "D")                               ~ "Dental",
    stringr::str_starts(code, "H")                               ~ "Behavioral Health",
    stringr::str_starts(code, "E") |
      stringr::str_starts(code, "K") |
      stringr::str_starts(code, "L")                             ~ "Durable Medical Equipment & Supplies",
    stringr::str_detect(code, "^T(20|21|22|51)") |
      stringr::str_starts(code, "A04")                           ~ "Transportation",
    stringr::str_starts(code, "T")                               ~ "Long-Term Services & Supports",
    code %in% c("G0257","G0323","G0324","G0325","G0326","G0327") ~ "Dialysis & ESRD",
    is_cpt & num >= 90791 & num <= 90899                         ~ "Behavioral Health",
    is_cpt & num >= 90935 & num <= 90999                         ~ "Dialysis & ESRD",
    is_cpt & num >= 99201 & num <= 99499                         ~ "Evaluation & Management",
    is_cpt & num >= 10021 & num <= 69990                         ~ "Surgery & Outpatient Procedures",
    is_cpt & num >= 70010 & num <= 79999                         ~ "Radiology & Imaging",
    is_cpt & num >= 80047 & num <= 89398                         ~ "Laboratory & Diagnostics",
    TRUE                                                         ~ "Other Medical Services"
  )
}

hcpcs_by_group <- hcpcs_sum |>
  mutate(disease_group = classify_hcpcs(hcpcs)) |>
  filter(disease_group %in% top_groups)
```

### What codes drive each disease area?

```{r disease-area-hcpcs-mix}
#| fig-cap: "Top 5 HCPCS codes by total Medicaid paid within each disease-focused group. The codes carrying the weight are very different per group — J-codes (specialty infusibles and injectables) dominate the drug bucket; a handful of E&M-adjacent therapy codes dominate Behavioral Health; a single periodic-exam code (D0120) and a pair of restoration codes drive most of Dental."
#| fig-width: 10
#| fig-height: 7

top_hcpcs_per_group <- hcpcs_by_group |>
  dplyr::group_by(disease_group) |>
  dplyr::arrange(dplyr::desc(total_paid), .by_group = TRUE) |>
  dplyr::slice_head(n = 5) |>
  dplyr::ungroup() |>
  dplyr::mutate(
    disease_group = factor(disease_group, levels = top_groups),
    row_id        = dplyr::row_number()
  )

ggplot(top_hcpcs_per_group,
       aes(x = stats::reorder(paste(hcpcs, row_id), total_paid),
           y = total_paid)) +
  geom_col(fill = "#54278f") +
  geom_text(aes(label = format_money(total_paid)),
            hjust = -0.08, size = 3, color = "#333") +
  facet_wrap(~ disease_group, scales = "free", ncol = 2) +
  scale_x_discrete(labels = \(x) sub(" \\d+$", "", x)) +
  scale_y_continuous(labels = format_money,
                     expand = expansion(mult = c(0, 0.22))) +
  coord_flip() +
  labs(x = NULL, y = "Total paid (all billing providers)")
```

### How spatially concentrated is each disease area?

```{r disease-area-concentration}
#| fig-cap: "Gini coefficient of county Medicaid paid within each disease-focused group, and the share of that group's total spend captured by just the top 10 counties. A Gini near 1 means spend collapses onto a handful of counties — expected for sparse, high-acuity categories; Gini near 0 is evenly diffused public outlay."
#| fig-width: 9
#| fig-height: 4.5

gini <- function(x) {
  x <- sort(x[!is.na(x) & x >= 0])
  n <- length(x)
  if (n <= 1 || sum(x) == 0) return(NA_real_)
  2 * sum(seq_len(n) * x) / (n * sum(x)) - (n + 1) / n
}

spatial_conc <- county_atlas |>
  filter(disease_group %in% top_groups) |>
  group_by(disease_group) |>
  summarise(
    counties_n    = dplyr::n(),
    gini_paid     = gini(total_paid),
    top10_share   = sum(sort(total_paid, decreasing = TRUE)[seq_len(min(10, dplyr::n()))],
                        na.rm = TRUE) / sum(total_paid, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(gini_paid))

spatial_conc |>
  mutate(disease_group = fct_reorder(disease_group, gini_paid)) |>
  ggplot(aes(disease_group, gini_paid)) +
  geom_col(fill = "#756bb1") +
  geom_text(aes(label = sprintf("Gini %.2f · top-10 share %s",
                                gini_paid,
                                scales::percent(top10_share, accuracy = 1))),
            hjust = -0.05, size = 3.1, color = "#333") +
  coord_flip(clip = "off") +
  scale_y_continuous(limits = c(0, 1.05),
                     labels = scales::number_format(accuracy = 0.01),
                     expand = expansion(mult = c(0, 0))) +
  labs(x = NULL, y = "Gini of county Medicaid paid (within disease group)") +
  theme(plot.margin = margin(5, 110, 5, 5))
```

### Which states lead each disease area?

```{r disease-area-state-hotspots}
state_hotspots <- county_atlas |>
  filter(disease_group %in% top_groups, !is.na(state)) |>
  group_by(state, disease_group) |>
  summarise(
    state_paid = sum(total_paid,  na.rm = TRUE),
    state_bene = sum(total_bene,  na.rm = TRUE),
    counties   = dplyr::n(),
    .groups = "drop"
  ) |>
  filter(state_bene >= 1000) |>
  mutate(state_ppb = state_paid / state_bene) |>
  group_by(disease_group) |>
  arrange(desc(state_ppb), .by_group = TRUE) |>
  slice_head(n = 5) |>
  ungroup() |>
  mutate(disease_group = factor(disease_group, levels = top_groups)) |>
  arrange(disease_group, desc(state_ppb))

knitr::kable(
  state_hotspots |>
    transmute(
      `Disease group`      = disease_group,
      State                = state,
      Counties             = counties,
      `Total paid`         = format_money(state_paid),
      Beneficiaries        = format(state_bene, big.mark = ","),
      `$ per beneficiary`  = format_money(state_ppb)
    ),
  caption = "Top 5 states by Medicaid paid-per-beneficiary within each disease-focused group (states with at least 1,000 beneficiaries in that group)."
)
```

## Service-area explorer — where non-clinical Medicaid dollars go

Disease categories answer *which diseases cost most*. Service categories answer a complementary question: **what non-clinical support Medicaid underwrites at scale.** Transportation (NEMT), Long-Term Services & Supports, the Evaluation & Management bucket of office visits, and the Other Medical Services catch-all together account for a majority of paid dollars in the dataset. The same three views applied to that subset:

```{r service-area-hcpcs-mix}
#| fig-cap: "Top 5 HCPCS codes by total Medicaid paid within each service category. Transportation collapses onto a handful of A0/T20-series non-emergency transport codes; LTSS is driven by T-series personal-care and case-management codes; E&M is dominated by the 99213/99214 office-visit pair; Other Medical Services surfaces the un-disambiguated 99000-series the classifier punts on."
#| fig-width: 10
#| fig-height: 6.5

hcpcs_by_service <- hcpcs_sum |>
  mutate(disease_group = classify_hcpcs(hcpcs)) |>
  filter(disease_group %in% top_service_groups)

top_hcpcs_per_service <- hcpcs_by_service |>
  dplyr::group_by(disease_group) |>
  dplyr::arrange(dplyr::desc(total_paid), .by_group = TRUE) |>
  dplyr::slice_head(n = 5) |>
  dplyr::ungroup() |>
  dplyr::mutate(
    disease_group = factor(disease_group, levels = top_service_groups),
    row_id        = dplyr::row_number()
  )

ggplot(top_hcpcs_per_service,
       aes(x = stats::reorder(paste(hcpcs, row_id), total_paid),
           y = total_paid)) +
  geom_col(fill = "#b08a49") +
  geom_text(aes(label = format_money(total_paid)),
            hjust = -0.08, size = 3, color = "#333") +
  facet_wrap(~ disease_group, scales = "free", ncol = 2) +
  scale_x_discrete(labels = \(x) sub(" \\d+$", "", x)) +
  scale_y_continuous(labels = format_money,
                     expand = expansion(mult = c(0, 0.22))) +
  coord_flip() +
  labs(x = NULL, y = "Total paid (all billing providers)")
```

```{r service-area-concentration}
#| fig-cap: "Spatial concentration of Medicaid paid per service category. LTSS and E&M dollars diffuse across thousands of counties (low Gini); NEMT is more concentrated — not surprising since transport benefits cluster in urban corridors and aging-rural regions with long travel distances to care."
#| fig-width: 9
#| fig-height: 3.8

service_conc <- county_atlas |>
  filter(disease_group %in% top_service_groups) |>
  group_by(disease_group) |>
  summarise(
    counties_n    = dplyr::n(),
    gini_paid     = gini(total_paid),
    top10_share   = sum(sort(total_paid, decreasing = TRUE)[seq_len(min(10, dplyr::n()))],
                        na.rm = TRUE) / sum(total_paid, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(gini_paid))

service_conc |>
  mutate(disease_group = fct_reorder(disease_group, gini_paid)) |>
  ggplot(aes(disease_group, gini_paid)) +
  geom_col(fill = "#b08a49") +
  geom_text(aes(label = sprintf("Gini %.2f · top-10 share %s",
                                gini_paid,
                                scales::percent(top10_share, accuracy = 1))),
            hjust = -0.05, size = 3.1, color = "#333") +
  coord_flip(clip = "off") +
  scale_y_continuous(limits = c(0, 1.05),
                     labels = scales::number_format(accuracy = 0.01),
                     expand = expansion(mult = c(0, 0))) +
  labs(x = NULL, y = "Gini of county Medicaid paid (within service category)") +
  theme(plot.margin = margin(5, 110, 5, 5))
```

```{r service-area-state-hotspots}
service_state_hotspots <- county_atlas |>
  filter(disease_group %in% top_service_groups, !is.na(state)) |>
  group_by(state, disease_group) |>
  summarise(
    state_paid = sum(total_paid,  na.rm = TRUE),
    state_bene = sum(total_bene,  na.rm = TRUE),
    counties   = dplyr::n(),
    .groups = "drop"
  ) |>
  filter(state_bene >= 1000) |>
  mutate(state_ppb = state_paid / state_bene) |>
  group_by(disease_group) |>
  arrange(desc(state_ppb), .by_group = TRUE) |>
  slice_head(n = 5) |>
  ungroup() |>
  mutate(disease_group = factor(disease_group, levels = top_service_groups)) |>
  arrange(disease_group, desc(state_ppb))

knitr::kable(
  service_state_hotspots |>
    transmute(
      `Service category`   = disease_group,
      State                = state,
      Counties             = counties,
      `Total paid`         = format_money(state_paid),
      Beneficiaries        = format(state_bene, big.mark = ","),
      `$ per beneficiary`  = format_money(state_ppb)
    ),
  caption = "Top 5 states by Medicaid paid-per-beneficiary within each service category (states with at least 1,000 beneficiaries in that category)."
)
```

## State-level intensity

```{r state-intensity}
#| fig-cap: "Peer-group-excess Medicaid spending by state. Height is the sum of paid dollars above each HCPCS's peer median, aggregated at prep time across all providers billing in that state."

state_excess <- state_sum |>
  arrange(desc(excess_paid)) |>
  slice_head(n = 25)

ggplot(state_excess, aes(reorder(state, excess_paid), excess_paid)) +
  geom_col(fill = "#0d5c63") +
  coord_flip() +
  scale_y_continuous(labels = format_money) +
  labs(x = NULL, y = "Paid above HCPCS peer median ($)")
```

## Takeaways

- **Heavy tails demand robust statistics.** A naive z-score on claims data is dominated by a few mega-providers; MAD-based scoring is still simple, audit-friendly, and far more defensible.
- **Rule-based and unsupervised flags agree where it counts.** The overlap between a robust-z ≥ 5 rule and an isolation-forest's 1% contamination call is the highest-confidence outlier set — the set a Program Integrity team would actually review first.
- **Geography is where interpretation starts.** Clusters on the map are a prompt, not a conclusion — state-level Medicaid rate structures, MCO contracting, and referral network differences can all produce geographic patterns that look like outliers but are artifacts of policy.

## Next steps

- **Specialty-adjusted peer groups.** Today the peer group is "everyone billing this HCPCS"; tighter peer groups (same specialty, similar patient panel size) will surface a different, higher-quality signal.
- **Time-aware outlier detection.** Repeat the analysis on rolling 3-month windows to separate persistent outliers from transient spikes (COVID, policy shocks, one-off audits).
- **SDoH overlay.** Merge state-level CDC PLACES / ACS indicators to test whether residual excess spending tracks community need vs. billing behavior — a health-equity stress test for the flagged list.

---

::: {.callout-tip}
### Reading this as code samples
Every cell above is the real executed code — not screenshots. Use the **Code** dropdown at the top-right to toggle visibility, copy blocks, or view the [raw `.qmd` source](https://github.com/paulinedmdf/paulinadelmundomd/blob/main/projects/02-medicaid-outliers/analysis.qmd) on GitHub. The one-shot DuckDB prep step lives in [`prep.py`](https://github.com/paulinedmdf/paulinadelmundomd/blob/main/projects/02-medicaid-outliers/prep.py).
:::
