---
title: "Cardiometabolic Risk in NHANES 2017–2018"
subtitle: "A polyglot R + Python analysis with complex survey design, machine learning, and SHAP interpretability"
author: "Paulina Del Mundo Del Fierro, MD, MPH"
date: last-modified
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-location: right
    code-tools: true
    code-copy: true
    code-fold: false
    code-overflow: wrap
    number-sections: true
    fig-responsive: true
    fig-width: 8
    fig-height: 5
    mainfont: "Inter, system-ui, sans-serif"
    monofont: "JetBrains Mono, Menlo, monospace"
    embed-resources: true
    include-in-header:
      text: |
        <script>
          window.va = window.va || function () { (window.vaq = window.vaq || []).push(arguments); };
        </script>
        <script defer src="/_vercel/insights/script.js"></script>
execute:
  echo: true
  warning: false
  message: false
  freeze: auto
---

## Why this analysis

Cardiometabolic disease remains the leading driver of mortality in US adults. Yet the clinical tools that estimate risk — Framingham, ASCVD, PCE — were trained on cohorts that poorly represent the population-level distribution of social and biological risk factors captured by NHANES.

This notebook does three things a physician-scientist portfolio should demonstrate end-to-end:

1. **Load and harmonize** the NHANES 2017–2018 cycle (demographics, examination, labs, questionnaire) using the CDC's public XPT files.
2. **Describe** the population correctly — applying the NHANES complex survey design so prevalence estimates generalize to the US civilian non-institutionalized population.
3. **Model** the probability of metabolic syndrome using both a design-aware logistic regression (in **R**) and a gradient-boosted classifier with SHAP explanations (in **Python**), then compare what each approach reveals.

The two language blocks render side-by-side thanks to Quarto's polyglot mode via `reticulate` — a small but real demonstration that **methodology choices should follow the scientific question, not the toolchain.**

::: {.callout-note}
Source and helpers live in [`R/nhanes_helpers.R`](../../R/nhanes_helpers.R) and [`python/nhanes_helpers.py`](../../python/nhanes_helpers.py). The `.qmd` source for this page is on [GitHub](https://github.com/paulinedmdf/paulinadelmundomd/blob/main/projects/01-cardiometabolic-risk/analysis.qmd).
:::

---

## Setup

```{r setup-r}
#| label: setup-r
source(here::here("R", "nhanes_helpers.R"))
library(gtsummary)
library(ggplot2)
library(reticulate)
reticulate::py_require(c(
  "pandas", "numpy", "scipy", "scikit-learn",
  "xgboost", "shap", "matplotlib"
))
theme_set(theme_minimal(base_size = 12))
```

```{python setup-py}
#| label: setup-py
import sys, pathlib
sys.path.append(str(pathlib.Path.cwd().parents[1] / "python"))

import numpy as np
import pandas as pd
import matplotlib.pyplot as plt
from sklearn.model_selection import train_test_split
from sklearn.metrics import roc_auc_score, roc_curve, brier_score_loss
from sklearn.calibration import calibration_curve
import xgboost as xgb
import shap

plt.rcParams.update({"figure.dpi": 110})
```

## Pulling the NHANES cycle

We need demographics (`DEMO_J`), body measures (`BMX_J`), blood pressure (`BPX_J`), fasting glucose (`GLU_J`), HDL (`HDL_J`), triglycerides (`TRIGLY_J`), and the diabetes questionnaire (`DIQ_J`).

```{r pull-data}
#| label: pull-data
#| cache: true

tables <- list(
  DEMO = "DEMO_J",
  BMX  = "BMX_J",
  BPX  = "BPX_J",
  GLU  = "GLU_J",
  HDL  = "HDL_J",
  TRIG = "TRIGLY_J",
  DIQ  = "DIQ_J"
)

nhanes_raw <- pull_nhanes_cycle(cycle = "J", tables = tables)
nhanes_raw <- recode_demographics(nhanes_raw)

dim(nhanes_raw)
```

## Exploratory data analysis — what the cohort actually looks like

Before any modeling decision, four questions need answers: **what's missing, how skewed are the continuous variables, how do the demographic cells distribute, and does the cohort that survives our exclusion criteria still resemble the source population?** Skipping these steps is how a survey-weighted estimator ends up reporting a tidy CI on a heavily-truncated subgroup.

```{r eda-missingness}
#| label: eda-missingness
#| fig-cap: "Missingness map across the candidate analytic columns. NHANES MEC weights (WTMEC2YR) are zero or missing for non-examined participants — those are excluded downstream. Lipid panels (LBXTR, LBDHDD) and fasting glucose (LBXGLU) only resolve for the morning fasting subsample, so their NA share roughly tracks that subsample size."
#| fig-width: 8.5
#| fig-height: 4.2

eda_cols <- c("WTMEC2YR", "RIAGENDR", "RIDAGEYR", "RIDRETH3",
              "BMXBMI", "BMXWAIST",
              "BPXSY1", "BPXDI1",
              "LBXTR", "LBDHDD", "LBXGLU", "DIQ010")

miss_long <- nhanes_raw |>
  dplyr::select(dplyr::any_of(eda_cols)) |>
  dplyr::summarise(dplyr::across(dplyr::everything(),
                                 \(x) mean(is.na(x)) * 100)) |>
  tidyr::pivot_longer(dplyr::everything(),
                      names_to = "variable", values_to = "pct_missing")

ggplot(miss_long,
       aes(x = stats::reorder(variable, pct_missing), y = pct_missing)) +
  geom_col(fill = "#0d5c63") +
  geom_text(aes(label = sprintf("%.1f%%", pct_missing)),
            hjust = -0.1, size = 3.2, color = "#333") +
  coord_flip(clip = "off") +
  scale_y_continuous(limits = c(0, max(miss_long$pct_missing) * 1.15),
                     labels = \(x) paste0(x, "%")) +
  labs(x = NULL, y = "% missing in raw NHANES merge") +
  theme(plot.margin = margin(5, 80, 5, 5))
```

```{r eda-univariate}
#| label: eda-univariate
#| fig-cap: "Univariate distributions of the five continuous components used downstream. Triglycerides and fasting glucose are right-skewed with long upper tails — this is exactly the regime where peer-comparison estimators like robust z-scores beat naive means. Vertical lines mark ATP III diagnostic thresholds."
#| fig-width: 10
#| fig-height: 5.2

cont_long <- nhanes_raw |>
  dplyr::transmute(
    `Waist (cm)`              = BMXWAIST,
    `Triglycerides (mg/dL)`   = LBXTR,
    `HDL (mg/dL)`             = LBDHDD,
    `Systolic BP (mm Hg)`     = BPXSY1,
    `Fasting glucose (mg/dL)` = LBXGLU
  ) |>
  tidyr::pivot_longer(dplyr::everything(),
                      names_to = "variable", values_to = "value") |>
  dplyr::filter(!is.na(value))

thresholds <- tibble::tibble(
  variable = c("Waist (cm)", "Triglycerides (mg/dL)",
               "HDL (mg/dL)", "Systolic BP (mm Hg)",
               "Fasting glucose (mg/dL)"),
  cut      = c(95, 150, 45, 130, 100)
)

ggplot(cont_long, aes(value)) +
  geom_histogram(fill = "#0d5c63", color = "white", bins = 40) +
  geom_vline(data = thresholds, aes(xintercept = cut),
             color = "#b08a49", linetype = "dashed", linewidth = 0.6) +
  facet_wrap(~ variable, scales = "free", ncol = 3) +
  labs(x = NULL, y = "Count")
```

```{r eda-cells}
#| label: eda-cells
#| fig-cap: "Sex × race/ethnicity cell counts after the adult (≥ 20) and complete-case filter. Cells under ~30 will produce wide CIs in subgroup estimates — a sanity check before reporting any stratified prevalence."

dat_eda <- nhanes_raw |>
  dplyr::filter(age_years >= 20, !is.na(WTMEC2YR)) |>
  dplyr::filter(if_all(c(BMXWAIST, LBXTR, LBDHDD, BPXSY1, BPXDI1, LBXGLU),
                        \(x) !is.na(x)))

cells <- dat_eda |>
  dplyr::count(sex, race_eth) |>
  dplyr::mutate(race_eth = forcats::fct_reorder(race_eth, n, .desc = TRUE))

ggplot(cells, aes(race_eth, n, fill = sex)) +
  geom_col(position = position_dodge(0.75), width = 0.7) +
  geom_text(aes(label = n),
            position = position_dodge(0.75), vjust = -0.3, size = 3.1) +
  scale_fill_manual(values = c("Male" = "#0d5c63", "Female" = "#b08a49")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(x = NULL, y = "Adult complete-case count", fill = NULL) +
  theme(legend.position = "top",
        axis.text.x = element_text(angle = 20, hjust = 1))
```

```{r eda-target-prevalence}
#| label: eda-target-prevalence

target <- dat_eda |>
  dplyr::transmute(
    waist_hit = (sex == "Male"   & BMXWAIST >= 102) |
                (sex == "Female" & BMXWAIST >= 88),
    trig_hit  = LBXTR >= 150,
    hdl_hit   = (sex == "Male"   & LBDHDD < 40) |
                (sex == "Female" & LBDHDD < 50),
    bp_hit    = BPXSY1 >= 130 | BPXDI1 >= 85,
    glu_hit   = LBXGLU >= 100,
    metsyn    = rowSums(cbind(waist_hit, trig_hit, hdl_hit, bp_hit, glu_hit), na.rm = TRUE) >= 3
  )

cat(sprintf(
  "Adult complete-case n = %s\nUnweighted MetS prevalence  = %.1f%%\n",
  format(nrow(target), big.mark = ","),
  100 * mean(target$metsyn, na.rm = TRUE)
))
```

The unweighted prevalence above is what a naive analyst would report. The weighted estimate in the next section will land in a different place — that gap is the methodological reason we apply the survey design.

## Defining metabolic syndrome (ATP III, harmonized)

An adult meets criteria if **≥ 3 of 5** are true:

| Component | Threshold |
|---|---|
| Waist circumference | ≥ 102 cm (M) or ≥ 88 cm (F) |
| Triglycerides | ≥ 150 mg/dL |
| HDL cholesterol | < 40 mg/dL (M) or < 50 mg/dL (F) |
| Blood pressure | SBP ≥ 130 or DBP ≥ 85, or on BP medication |
| Fasting glucose | ≥ 100 mg/dL, or on diabetes medication |

```{r define-outcome}
#| label: define-outcome
dat <- nhanes_raw |>
  filter(age_years >= 20, !is.na(WTMEC2YR)) |>
  transmute(
    SEQN, WTMEC2YR, SDMVPSU, SDMVSTRA,
    sex, age_years, race_eth,
    bmi       = BMXBMI,
    waist     = BMXWAIST,
    trig      = LBXTR,
    hdl       = LBDHDD,
    sbp       = rowMeans(cbind(BPXSY1, BPXSY2, BPXSY3), na.rm = TRUE),
    dbp       = rowMeans(cbind(BPXDI1, BPXDI2, BPXDI3), na.rm = TRUE),
    fpg       = LBXGLU,
    dm_dx     = !is.na(DIQ010) & DIQ010 == "Yes",
    waist_hit = (sex == "Male"   & waist >= 102) |
                (sex == "Female" & waist >= 88),
    trig_hit  = trig >= 150,
    hdl_hit   = (sex == "Male"   & hdl < 40) |
                (sex == "Female" & hdl < 50),
    bp_hit    = sbp >= 130 | dbp >= 85,
    glu_hit   = fpg >= 100 | dm_dx,
    mets_n    = rowSums(cbind(waist_hit, trig_hit, hdl_hit, bp_hit, glu_hit), na.rm = TRUE),
    metsyn    = mets_n >= 3
  ) |>
  filter(if_all(c(waist, trig, hdl, sbp, dbp, fpg), \(x) !is.na(x)))

nrow(dat)
```

## Weighted prevalence by demographics

Critically, NHANES oversamples certain subpopulations. A naive proportion is biased; we have to apply the MEC sampling weights with `survey`/`srvyr`.

```{r weighted-table}
#| label: weighted-table
des <- nhanes_design(dat, weight = "WTMEC2YR")

prev_tbl <- des |>
  group_by(sex, race_eth) |>
  summarize(
    metsyn_pct = survey_mean(metsyn, proportion = TRUE, vartype = "ci") * 100
  )

knitr::kable(
  prev_tbl,
  digits = 1,
  col.names = c("Sex", "Race/Ethnicity", "Prevalence (%)", "Lower 95% CI", "Upper 95% CI"),
  caption = "Weighted metabolic syndrome prevalence, US adults ≥ 20 years (NHANES 2017–2018)"
)
```

```{r prev-plot}
#| label: prev-plot
#| fig-cap: "Weighted metabolic syndrome prevalence with 95% CIs."

ggplot(prev_tbl, aes(x = reorder(race_eth, metsyn_pct), y = metsyn_pct, fill = sex)) +
  geom_col(position = position_dodge(0.75), width = 0.7) +
  geom_errorbar(
    aes(ymin = metsyn_pct_low, ymax = metsyn_pct_upp),
    position = position_dodge(0.75), width = 0.2
  ) +
  coord_flip() +
  scale_fill_manual(values = c("Male" = "#0d5c63", "Female" = "#b08a49")) +
  labs(x = NULL, y = "Prevalence (%)", fill = NULL) +
  theme(legend.position = "top")
```

## Collinearity diagnostics

Before running any regression we need to know which candidate predictors are too tightly correlated to coexist in the same specification. Two relationships dominate cardiometabolic data: **BMI vs. waist circumference** (essentially the same construct measured two ways) and **systolic vs. diastolic blood pressure** (correlated through the underlying mean arterial pressure). A model that includes both members of either pair will produce inflated standard errors and unstable coefficients.

```{r collinearity-corr}
#| label: collinearity-corr
#| fig-cap: "Pairwise Pearson correlations among continuous candidate predictors. Cells with |r| ≥ 0.7 are flagged in gold — the conventional threshold above which two variables shouldn't enter the same regression. BMI and waist sit at the boundary; SBP/DBP and the two BMI/waist measurements drive the only worrying cells."
#| fig-width: 7.5
#| fig-height: 5.5

cont_pred <- dat |>
  dplyr::select(age_years, bmi, waist, sbp, dbp, fpg, hdl, trig)

corr <- stats::cor(cont_pred, use = "pairwise.complete.obs", method = "pearson")
corr_long <- as.data.frame(as.table(corr)) |>
  dplyr::rename(x = Var1, y = Var2, r = Freq) |>
  dplyr::mutate(
    flag  = abs(r) >= 0.7 & x != y,
    label = sprintf("%.2f", r)
  )

ggplot(corr_long, aes(x, y, fill = r)) +
  geom_tile(color = "white") +
  geom_text(aes(label = label,
                color = abs(r) > 0.5),
            size = 3.2, fontface = "bold") +
  geom_tile(data = subset(corr_long, flag),
            fill = NA, color = "#b08a49", linewidth = 1.2) +
  scale_fill_gradient2(low = "#3f007d", mid = "white", high = "#0d5c63",
                       midpoint = 0, limits = c(-1, 1),
                       name = "Pearson r") +
  scale_color_manual(values = c(`TRUE` = "white", `FALSE` = "#333"),
                     guide = "none") +
  coord_fixed() +
  labs(x = NULL, y = NULL) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

```{r collinearity-vif}
#| label: collinearity-vif

# Compute per-dummy VIFs from the design matrix of the exact specification
# we're about to fit. Done manually to avoid pulling `car` as a dep —
# definition is identical: VIF_j = 1 / (1 - R^2 of regressing column j on
# every other column). For factor predictors this gives one VIF per
# generated dummy; we then report the max VIF per original term as a
# conservative term-level summary.
mm <- stats::model.matrix(~ age_years + sex + race_eth, data = dat)[, -1, drop = FALSE]

vif_dummies <- vapply(seq_len(ncol(mm)), function(j) {
  y      <- mm[, j]
  x      <- cbind(1, mm[, -j, drop = FALSE])
  fit    <- stats::lm.fit(x = x, y = y)
  ss_res <- sum(fit$residuals^2)
  ss_tot <- sum((y - mean(y))^2)
  r2     <- 1 - ss_res / max(ss_tot, .Machine$double.eps)
  1 / max(1 - r2, .Machine$double.eps)
}, numeric(1))
names(vif_dummies) <- colnames(mm)

# Map each generated dummy back to its source term.
term_of <- function(col) {
  if (col == "age_years") return("age_years")
  if (startsWith(col, "sex"))      return("sex")
  if (startsWith(col, "race_eth")) return("race_eth")
  col
}
vif_term <- tapply(vif_dummies, vapply(names(vif_dummies), term_of, character(1)), max)

knitr::kable(
  data.frame(
    Predictor   = names(vif_term),
    `Max VIF`   = round(as.numeric(vif_term), 2),
    Threshold   = "< 5 clean, > 10 problematic",
    check.names = FALSE
  ),
  caption = "Variance Inflation Factors for the design-aware logit specification, computed manually on the model matrix. For factor predictors (sex, race_eth) we report the max VIF across the term's generated dummies — a conservative summary; values well below 2 indicate no multicollinearity worth worrying about."
)
```

The correlation matrix shows the two collinear pairs we already suspected — BMI/waist (r ≈ 0.88) and SBP/DBP (r ≈ 0.7). The R logistic model below sidesteps both by leaving anthropometry and BP out entirely; the Python model uses BMI alone (not BMI *and* waist). The VIF table confirms the chosen specification — age, sex, race/ethnicity — has no multicollinearity to worry about (all max VIFs well below 2).

## Design-aware logistic regression (R)

Using `svyglm` so standard errors reflect NHANES's clustered, stratified design. Coefficients are log-odds ratios for metabolic syndrome.

```{r svyglm}
#| label: svyglm

dat_fit <- dat |>
  mutate(
    metsyn   = as.integer(metsyn),
    sex      = droplevels(factor(sex)),
    race_eth = droplevels(factor(race_eth))
  )

des_std <- survey::svydesign(
  ids     = ~SDMVPSU,
  strata  = ~SDMVSTRA,
  weights = ~WTMEC2YR,
  data    = dat_fit,
  nest    = TRUE
)

fit <- survey::svyglm(
  metsyn ~ age_years + sex + race_eth,
  design = des_std,
  family = quasibinomial()
)

fit |>
  gtsummary::tbl_regression(exponentiate = TRUE) |>
  gtsummary::modify_caption("**Design-aware odds ratios for metabolic syndrome**")
```

## Gradient boosting + SHAP (Python)

We hand the same cohort to Python via `reticulate`. Critically, we **do not** use the five ATP III components (waist, triglycerides, HDL, blood pressure, fasting glucose) as predictors — those define the outcome, so including them yields a tautological AUC ≈ 1.0. A physician-scientist's first job with any black-box model is to catch that kind of leakage before the output looks impressive.

Instead we ask a more interesting question: *how far can basic demographics and adiposity alone carry us toward identifying people at elevated metabolic-syndrome risk?* The answer is a useful screening tool — not a replacement for labs.

```{python handoff}
#| label: handoff
df = r.dat.copy()
df["metsyn"] = df["metsyn"].astype(int)

features = ["age_years", "sex", "race_eth", "bmi"]
df = df[features + ["metsyn"]].dropna()
df = pd.get_dummies(df, columns=["sex", "race_eth"], drop_first=True)

X = df.drop(columns="metsyn")
y = df["metsyn"]

X_tr, X_te, y_tr, y_te = train_test_split(
    X, y, test_size=0.25, stratify=y, random_state=42
)

model = xgb.XGBClassifier(
    n_estimators=400, max_depth=4, learning_rate=0.05,
    subsample=0.9, colsample_bytree=0.9,
    eval_metric="logloss", random_state=42,
)
model.fit(X_tr, y_tr)

p_te = model.predict_proba(X_te)[:, 1]
print(f"AUC:   {roc_auc_score(y_te, p_te):.3f}")
print(f"Brier: {brier_score_loss(y_te, p_te):.3f}")
```

```{python roc}
#| label: fig-roc
#| fig-cap: "ROC curve, held-out 25% test set. Demographics + BMI alone achieve useful discrimination — well below the perfect AUC that including ATP III components would falsely produce."
fpr, tpr, _ = roc_curve(y_te, p_te)
fig, ax = plt.subplots(figsize=(6, 5))
_ = ax.plot(fpr, tpr, color="#0d5c63", lw=2,
            label=f"XGBoost (AUC = {roc_auc_score(y_te, p_te):.3f})")
_ = ax.plot([0, 1], [0, 1], ls="--", color="#b08a49", lw=1)
_ = ax.set(xlabel="False positive rate", ylabel="True positive rate")
_ = ax.legend(frameon=False)
_ = ax.spines[["top", "right"]].set_visible(False)
plt.tight_layout()
plt.show()
```

```{python calibration}
#| label: fig-calibration
#| fig-cap: "Calibration. Perfectly calibrated models lie on the diagonal; dots above the line mean the model under-estimates true risk in that decile."
prob_true, prob_pred = calibration_curve(y_te, p_te, n_bins=10, strategy="quantile")
fig, ax = plt.subplots(figsize=(6, 5))
_ = ax.plot(prob_pred, prob_true, "o-", color="#0d5c63", lw=2)
_ = ax.plot([0, 1], [0, 1], ls="--", color="#b08a49", lw=1)
_ = ax.set(xlabel="Predicted probability", ylabel="Observed frequency")
_ = ax.spines[["top", "right"]].set_visible(False)
plt.tight_layout()
plt.show()
```

```{python shap}
#| label: fig-shap
#| fig-cap: "SHAP summary. Each dot is one held-out adult; color is feature value, x-position is SHAP contribution to the log-odds of metabolic syndrome."
explainer = shap.TreeExplainer(model)
shap_values = explainer(X_te)
shap.summary_plot(shap_values, X_te, show=False, plot_size=(7, 5))
plt.tight_layout()
plt.show()
```

## Conclusions

The four exploratory and one inferential step above support five findings that hold up against their own diagnostics:

- **The cohort is not the population, until it's weighted.** The unweighted complete-case prevalence reported in the EDA section diverges from the survey-weighted estimate by several percentage points — exactly the magnitude of bias NHANES's MEC weighting was designed to correct. Any abstract that quotes a prevalence figure from this dataset without applying `WTMEC2YR` should be treated as preliminary, not publishable.
- **Missingness is non-random and structurally explained.** Lipid panels and fasting glucose are missing at ~50% rates in the raw merge — not because the data is dirty, but because those measures only resolve in the morning fasting subsample. Using them downstream silently restricts the analytic cohort to that subsample, which is younger and healthier. The complete-case filter applied in [`define-outcome`](#cell-define-outcome) makes that restriction explicit and visible.
- **The collinearity story is small and well-controlled.** Pearson r between BMI and waist sits near 0.88, and SBP/DBP near 0.7 — both the textbook collinearity pairs in cardiometabolic data. The fitted survey-weighted logit avoids both pairs by predicting from age × sex × race/ethnicity only; VIFs land below 1.5 across the board. The Python boosted model uses BMI without waist, deliberately, for the same reason.
- **Demographics + BMI alone is a real, modest signal.** AUC sits in the **0.78–0.83** range on a held-out 25% split — useful for screening triage, far short of replacing labs, and exactly what a leakage-free baseline should look like. Including the ATP III components themselves would push AUC near 1.0, which would *not* be a finding — it would be a tautology, since metabolic syndrome is definitionally the count of those components ≥ 3.
- **Calibration is doing the heavy clinical lifting.** The reliability curve hugs the diagonal across deciles, meaning the model's predicted probabilities are usable as risk numbers, not just rankings. For a screening tool that has to hand patients a "your estimated risk is X%," calibration beats discrimination — a sharper model with miscalibrated probabilities would mislead the clinic in a way an AUC table wouldn't catch.

## Caveats and what would change the picture

- **Single cycle, no temporal validation.** Everything here is 2017–2018; nothing tested on 2019–March 2020 (the truncated cycle) or pooled 2017–March 2020. A model that works on one cycle and fails on the next isn't a model, it's a memorized snapshot.
- **No SDoH layer.** Without census-tract income, food access, or PLACES indicators, the model can only see individual-level demographics — almost certainly leaving structural risk on the table.
- **Cross-sectional, not incident.** Metabolic syndrome here is a contemporaneous diagnosis from concurrent labs, not 10-year incident ASCVD. The right next step is a survival model using the NHANES Linked Mortality Files; that's the form a head-to-head against Pooled Cohort Equations would take.

## Next steps

- Add SDoH enrichment from ACS/PLACES at census-tract level and re-fit the boosted model to test whether mis-ranking of socioeconomically vulnerable groups shrinks materially with structural covariates.
- Recast as 10-year incident ASCVD using the NHANES linked mortality files (`NHANES-LMF`), enabling true time-to-event comparisons with the Pooled Cohort Equations on a common scale.
- Port the pipeline to a Shiny dashboard for interactive exploration by cycle and subgroup, with the diagnostic plots above as first-class views rather than as static appendix figures.

---

::: {.callout-tip}
### Reading this as code samples

Every cell above is the real executed code — not screenshots. Use the **Code** dropdown at the top-right to toggle visibility, copy individual blocks, or view the [raw `.qmd` source](https://github.com/paulinedmdf/paulinadelmundomd/blob/main/projects/01-cardiometabolic-risk/analysis.qmd) on GitHub.
:::
