---
title: "Cardiometabolic Risk in NHANES 2017–2018"
subtitle: "A polyglot R + Python analysis with complex survey design, machine learning, and SHAP interpretability"
author: "Paulina Del Mundo Del Fierro, MD, MPH"
date: last-modified
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-location: right
    code-tools: true
    code-copy: true
    code-fold: true
    code-summary: "Show code"
    code-overflow: wrap
    number-sections: true
    fig-responsive: true
    fig-width: 8
    fig-height: 5
    mainfont: "Inter, system-ui, sans-serif"
    monofont: "JetBrains Mono, Menlo, monospace"
    embed-resources: true
    include-in-header:
      text: |
        <script>
          window.va = window.va || function () { (window.vaq = window.vaq || []).push(arguments); };
        </script>
        <script defer src="/_vercel/insights/script.js"></script>
execute:
  echo: true
  warning: false
  message: false
  freeze: auto
---

::: {.callout-tip icon=false}
## Key findings — TL;DR

- **Survey-weighted prevalence of metabolic syndrome in US adults ≥ 20 years (NHANES 2015–2018, MI-pooled): ~33%.** The unweighted complete-case figure is ~5 points lower, the gap MI is built to recover.
- **Definition matters by ~5 percentage points.** Spread across NCEP ATP III, IDF 2005, and JIS 2009 brackets the headline; JIS shifts upward in Asian and Mexican-American subpopulations because it uses lower waist cutoffs.
- **PCE inputs beat demographics + BMI on the same outcome — and that's the *expected* result.** PCE inputs include 3 of the 5 MetS components, so the comparison is "what fraction of MetS signal sits in cardiovascular labs vs pure adiposity," not a clean apples-to-apples test.
- **Calibration carries the clinical interpretability.** AUC sits in 0.78–0.83 for the boosted model; the reliability curve hugs the diagonal across deciles. Calibration > discrimination for any tool that hands a patient a "your estimated risk is X%."
- **Methods used (every claim ties back to a specific chunk):** complex-survey design with pooled cycles + WTMEC4YR, MICE imputation pooled by Rubin's rules, design-aware logistic via `svyglm`, gradient-boosted classifier with SHAP, PCE comparator, TRIPOD+AI checklist as Appendix.
:::

## Why this analysis

Cardiometabolic disease remains the leading driver of mortality in US adults. Yet the clinical tools that estimate risk — Framingham, ASCVD, PCE — were trained on cohorts that poorly represent the population-level distribution of social and biological risk factors captured by NHANES.

This notebook does three things a physician-scientist portfolio should demonstrate end-to-end:

1. **Load and harmonize** the NHANES 2017–2018 cycle (demographics, examination, labs, questionnaire) using the CDC's public XPT files.
2. **Describe** the population correctly — applying the NHANES complex survey design so prevalence estimates generalize to the US civilian non-institutionalized population.
3. **Model** the probability of metabolic syndrome using both a design-aware logistic regression (in **R**) and a gradient-boosted classifier with SHAP explanations (in **Python**), then compare what each approach reveals.

The two language blocks render side-by-side thanks to Quarto's polyglot mode via `reticulate` — a small but real demonstration that **methodology choices should follow the scientific question, not the toolchain.**

::: {.callout-note}
Source and helpers live in [`R/nhanes_helpers.R`](../../R/nhanes_helpers.R) and [`python/nhanes_helpers.py`](../../python/nhanes_helpers.py). The `.qmd` source for this page is on [GitHub](https://github.com/paulinedmdf/paulinadelmundomd/blob/main/projects/01-cardiometabolic-risk/analysis.qmd).
:::

---

## Setup

```{r setup-r}
#| label: setup-r
source(here::here("R", "nhanes_helpers.R"))
library(gtsummary)
library(ggplot2)
library(purrr)
library(reticulate)

# Method packages — installed on first render if absent.
for (pkg in c("mice", "mitools", "e1071", "fitdistrplus",
              "performance", "lme4", "kableExtra")) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    install.packages(pkg, repos = "https://cloud.r-project.org")
  }
}

# Lightweight wrapper for hero tables — bootstrap styling, no monospaced
# greyscale dump. Plain knitr::kable() still works elsewhere for inline /
# diagnostic tables where the styled version is overkill.
pretty_kable <- function(x, caption = NULL, ...) {
  knitr::kable(x, caption = caption, ...) |>
    kableExtra::kable_styling(
      bootstrap_options = c("striped", "hover", "condensed", "responsive"),
      full_width        = FALSE,
      position          = "left",
      font_size         = 13
    )
}

reticulate::py_require(c(
  "pandas", "numpy", "scipy", "scikit-learn",
  "xgboost", "shap", "matplotlib"
))
theme_set(theme_minimal(base_size = 12))
```

```{python setup-py}
#| label: setup-py
import sys, pathlib
sys.path.append(str(pathlib.Path.cwd().parents[1] / "python"))

import numpy as np
import pandas as pd
import matplotlib.pyplot as plt
from sklearn.model_selection import train_test_split
from sklearn.metrics import roc_auc_score, roc_curve, brier_score_loss
from sklearn.calibration import calibration_curve
import xgboost as xgb
import shap

plt.rcParams.update({"figure.dpi": 110})
```

## Pulling and pooling the NHANES cycles

A single cycle is too small for stable subgroup estimates. We pool **NHANES 2015–2016 (cycle I)** and **2017–2018 (cycle J)** per the CDC analytic guidelines: each cycle's MEC weight is divided by the number of cycles, strata and PSU pass through unchanged, and SEQNs don't collide. The combined-cycle weight is `WTMEC4YR = WTMEC2YR / 2`.

We also pull three additional tables (`TCHOL`, `SMQ`, `BPQ`) needed for the Pooled Cohort Equations comparator downstream — total cholesterol, current-smoker status, and antihypertensive treatment.

```{r pull-data}
#| label: pull-data
#| cache: true

build_spec <- function(suffix) {
  list(
    DEMO  = paste0("DEMO_",   suffix),
    BMX   = paste0("BMX_",    suffix),
    BPX   = paste0("BPX_",    suffix),
    GLU   = paste0("GLU_",    suffix),
    HDL   = paste0("HDL_",    suffix),
    TRIG  = paste0("TRIGLY_", suffix),
    DIQ   = paste0("DIQ_",    suffix),
    TCHOL = paste0("TCHOL_",  suffix),
    SMQ   = paste0("SMQ_",    suffix),
    BPQ   = paste0("BPQ_",    suffix)
  )
}

cycle_specs <- list(
  "2015-2016" = build_spec("I"),
  "2017-2018" = build_spec("J")
)

nhanes_raw <- pool_nhanes_cycles(cycle_specs)
nhanes_raw <- recode_demographics(nhanes_raw)

cat(sprintf("Pooled rows: %s\n", format(nrow(nhanes_raw), big.mark = ",")))
print(table(cycle = nhanes_raw$cycle))
```

## Exploratory data analysis — what the cohort actually looks like

Before any modeling decision, four questions need answers: **what's missing, how skewed are the continuous variables, how do the demographic cells distribute, and does the cohort that survives our exclusion criteria still resemble the source population?** Skipping these steps is how a survey-weighted estimator ends up reporting a tidy CI on a heavily-truncated subgroup.

```{r eda-missingness}
#| label: eda-missingness
#| fig-cap: "Missingness map across the candidate analytic columns. NHANES MEC weights (WTMEC2YR) are zero or missing for non-examined participants — those are excluded downstream. Lipid panels (LBXTR, LBDHDD) and fasting glucose (LBXGLU) only resolve for the morning fasting subsample, so their NA share roughly tracks that subsample size."
#| fig-width: 8.5
#| fig-height: 4.2

eda_cols <- c("WTMEC2YR", "RIAGENDR", "RIDAGEYR", "RIDRETH3",
              "BMXBMI", "BMXWAIST",
              "BPXSY1", "BPXDI1",
              "LBXTR", "LBDHDD", "LBXGLU", "DIQ010")

miss_long <- nhanes_raw |>
  dplyr::select(dplyr::any_of(eda_cols)) |>
  dplyr::summarise(dplyr::across(dplyr::everything(),
                                 \(x) mean(is.na(x)) * 100)) |>
  tidyr::pivot_longer(dplyr::everything(),
                      names_to = "variable", values_to = "pct_missing")

ggplot(miss_long,
       aes(x = stats::reorder(variable, pct_missing), y = pct_missing)) +
  geom_col(fill = "#0d5c63") +
  geom_text(aes(label = sprintf("%.1f%%", pct_missing)),
            hjust = -0.1, size = 3.2, color = "#333") +
  coord_flip(clip = "off") +
  scale_y_continuous(limits = c(0, max(miss_long$pct_missing) * 1.15),
                     labels = \(x) paste0(x, "%")) +
  labs(x = NULL, y = "% missing in raw NHANES merge") +
  theme(plot.margin = margin(5, 80, 5, 5))
```

```{r eda-univariate}
#| label: eda-univariate
#| fig-cap: "Univariate distributions of the five continuous components used downstream. Triglycerides and fasting glucose are right-skewed with long upper tails — this is exactly the regime where peer-comparison estimators like robust z-scores beat naive means. Vertical lines mark ATP III diagnostic thresholds."
#| fig-width: 10
#| fig-height: 5.2

cont_long <- nhanes_raw |>
  dplyr::transmute(
    `Waist (cm)`              = BMXWAIST,
    `Triglycerides (mg/dL)`   = LBXTR,
    `HDL (mg/dL)`             = LBDHDD,
    `Systolic BP (mm Hg)`     = BPXSY1,
    `Fasting glucose (mg/dL)` = LBXGLU
  ) |>
  tidyr::pivot_longer(dplyr::everything(),
                      names_to = "variable", values_to = "value") |>
  dplyr::filter(!is.na(value))

thresholds <- tibble::tibble(
  variable = c("Waist (cm)", "Triglycerides (mg/dL)",
               "HDL (mg/dL)", "Systolic BP (mm Hg)",
               "Fasting glucose (mg/dL)"),
  cut      = c(95, 150, 45, 130, 100)
)

ggplot(cont_long, aes(value)) +
  geom_histogram(fill = "#0d5c63", color = "white", bins = 40) +
  geom_vline(data = thresholds, aes(xintercept = cut),
             color = "#b08a49", linetype = "dashed", linewidth = 0.6) +
  facet_wrap(~ variable, scales = "free", ncol = 3) +
  labs(x = NULL, y = "Count")
```

```{r eda-cells}
#| label: eda-cells
#| fig-cap: "Sex × race/ethnicity cell counts after the adult (≥ 20) and complete-case filter. Cells under ~30 will produce wide CIs in subgroup estimates — a sanity check before reporting any stratified prevalence."

dat_eda <- nhanes_raw |>
  dplyr::filter(age_years >= 20, !is.na(WTMEC2YR)) |>
  dplyr::filter(if_all(c(BMXWAIST, LBXTR, LBDHDD, BPXSY1, BPXDI1, LBXGLU),
                        \(x) !is.na(x)))

cells <- dat_eda |>
  dplyr::count(sex, race_eth) |>
  dplyr::mutate(race_eth = forcats::fct_reorder(race_eth, n, .desc = TRUE))

ggplot(cells, aes(race_eth, n, fill = sex)) +
  geom_col(position = position_dodge(0.75), width = 0.7) +
  geom_text(aes(label = n),
            position = position_dodge(0.75), vjust = -0.3, size = 3.1) +
  scale_fill_manual(values = c("Male" = "#0d5c63", "Female" = "#b08a49")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(x = NULL, y = "Adult complete-case count", fill = NULL) +
  theme(legend.position = "top",
        axis.text.x = element_text(angle = 20, hjust = 1))
```

```{r eda-target-prevalence}
#| label: eda-target-prevalence

target <- dat_eda |>
  dplyr::transmute(
    waist_hit = (sex == "Male"   & BMXWAIST >= 102) |
                (sex == "Female" & BMXWAIST >= 88),
    trig_hit  = LBXTR >= 150,
    hdl_hit   = (sex == "Male"   & LBDHDD < 40) |
                (sex == "Female" & LBDHDD < 50),
    bp_hit    = BPXSY1 >= 130 | BPXDI1 >= 85,
    glu_hit   = LBXGLU >= 100,
    metsyn    = rowSums(cbind(waist_hit, trig_hit, hdl_hit, bp_hit, glu_hit), na.rm = TRUE) >= 3
  )

cat(sprintf(
  "Adult complete-case n = %s\nUnweighted MetS prevalence  = %.1f%%\n",
  format(nrow(target), big.mark = ","),
  100 * mean(target$metsyn, na.rm = TRUE)
))
```

The unweighted prevalence above is what a naive analyst would report. The weighted estimate in the next section will land in a different place — that gap is the methodological reason we apply the survey design.

## Defining metabolic syndrome (ATP III, harmonized)

An adult meets criteria if **≥ 3 of 5** are true:

| Component | Threshold |
|---|---|
| Waist circumference | ≥ 102 cm (M) or ≥ 88 cm (F) |
| Triglycerides | ≥ 150 mg/dL |
| HDL cholesterol | < 40 mg/dL (M) or < 50 mg/dL (F) |
| Blood pressure | SBP ≥ 130 or DBP ≥ 85, or on BP medication |
| Fasting glucose | ≥ 100 mg/dL, or on diabetes medication |

```{r define-outcome}
#| label: define-outcome
# Pre-complete-case cohort — keep one copy with NAs intact for MICE downstream,
# and one filtered copy for the design-aware analysis the existing chunks
# already use. The outcome derivation is identical; only the row filter
# differs.
dat_full <- nhanes_raw |>
  filter(age_years >= 20, !is.na(WTMEC4YR), WTMEC4YR > 0) |>
  transmute(
    SEQN, cycle, WTMEC4YR, SDMVPSU, SDMVSTRA,
    sex, age_years, race_eth,
    bmi          = BMXBMI,
    waist        = BMXWAIST,
    trig         = LBXTR,
    hdl          = LBDHDD,
    tc           = LBXTC,
    sbp          = rowMeans(cbind(BPXSY1, BPXSY2, BPXSY3), na.rm = TRUE),
    dbp          = rowMeans(cbind(BPXDI1, BPXDI2, BPXDI3), na.rm = TRUE),
    fpg          = LBXGLU,
    dm_dx        = !is.na(DIQ010) & DIQ010 == "Yes",
    bp_treated   = !is.na(BPQ050A) & BPQ050A == "Yes",
    smoker       = !is.na(SMQ040) & SMQ040 %in% c("Every day", "Some days"),
    waist_hit    = (sex == "Male"   & waist >= 102) |
                   (sex == "Female" & waist >= 88),
    trig_hit     = trig >= 150,
    hdl_hit      = (sex == "Male"   & hdl < 40) |
                   (sex == "Female" & hdl < 50),
    bp_hit       = sbp >= 130 | dbp >= 85,
    glu_hit      = fpg >= 100 | dm_dx,
    mets_n       = rowSums(cbind(waist_hit, trig_hit, hdl_hit, bp_hit, glu_hit),
                           na.rm = TRUE),
    metsyn       = mets_n >= 3
  )

# Coerce sbp/dbp NaN (when all three readings missing) back to NA so MICE
# treats them consistently.
dat_full <- dat_full |>
  mutate(across(c(sbp, dbp), \(x) ifelse(is.nan(x), NA, x)))

dat <- dat_full |>
  filter(if_all(c(waist, trig, hdl, sbp, dbp, fpg), \(x) !is.na(x)))

cat(sprintf(
  "Pre-CC pooled cohort  : %s\nComplete-case cohort  : %s (drop %.1f%%)\n",
  format(nrow(dat_full), big.mark = ","),
  format(nrow(dat),      big.mark = ","),
  100 * (1 - nrow(dat) / nrow(dat_full))
))
```

## Diagnostics on the analytic cohort

Three publication-grade checks before any modeling: a **formal distributional fit** for each continuous candidate (skew/kurtosis quantified, candidate distributions ranked by AIC), a **selection-bias check** comparing the dropped rows to the kept ones, and an **ICC variance partition** that quantifies how much of the outcome variance lives at each clustering level of the NHANES sample design.

```{r eda-distributional-fit}
#| label: eda-distributional-fit

cont_vars <- list(
  bmi   = dat_full$bmi,   waist = dat_full$waist,
  trig  = dat_full$trig,  hdl   = dat_full$hdl,
  sbp   = dat_full$sbp,   dbp   = dat_full$dbp,
  fpg   = dat_full$fpg,   tc    = dat_full$tc
)

skew_kurt <- vapply(cont_vars, function(x) {
  x <- x[!is.na(x)]
  c(skew = e1071::skewness(x), kurtosis = e1071::kurtosis(x))
}, numeric(2))

knitr::kable(
  data.frame(
    Variable = colnames(skew_kurt),
    n        = vapply(cont_vars, \(x) sum(!is.na(x)), integer(1)),
    Skew     = round(skew_kurt["skew", ], 2),
    `Excess kurtosis` = round(skew_kurt["kurtosis", ], 2),
    Shape    = ifelse(abs(skew_kurt["skew", ]) > 1, "Right-skewed",
                ifelse(abs(skew_kurt["skew", ]) > 0.5, "Mildly skewed", "Roughly symmetric")),
    check.names = FALSE
  ),
  caption = "Skewness and excess kurtosis on the eight continuous candidate variables. |skew| > 1 indicates a heavy tail that materially affects the choice of regression family or transformation. The right-skewed variables get a formal distribution fit in the next table."
)

skewed <- names(cont_vars)[abs(skew_kurt["skew", ]) > 1]
fit_results <- list()
for (v in skewed) {
  x <- cont_vars[[v]]
  x <- x[!is.na(x) & x > 0]
  fits <- list(
    normal = try(fitdistrplus::fitdist(x, "norm"),  silent = TRUE),
    lnorm  = try(fitdistrplus::fitdist(x, "lnorm"), silent = TRUE),
    gamma  = try(fitdistrplus::fitdist(x, "gamma"), silent = TRUE)
  )
  aics <- vapply(fits, function(f) {
    if (inherits(f, "try-error") || is.null(f)) NA_real_ else f$aic
  }, numeric(1))
  fit_results[[v]] <- aics
}

if (length(fit_results) > 0) {
  fit_mat <- do.call(rbind, fit_results)
  best    <- colnames(fit_mat)[apply(fit_mat, 1, function(r) which.min(r))]
  knitr::kable(
    data.frame(
      Variable = rownames(fit_mat),
      `AIC normal` = round(fit_mat[, "normal"], 0),
      `AIC log-normal` = round(fit_mat[, "lnorm"], 0),
      `AIC gamma` = round(fit_mat[, "gamma"], 0),
      `Best fit`  = best,
      check.names = FALSE
    ),
    caption = "AIC comparison of three candidate distributions on the right-skewed variables (positive values only). Lowest AIC = best fit. The 'best fit' column is the recommended starting family for any regression that takes one of these as outcome — for triglycerides and fasting glucose the answer is almost always log-normal or gamma rather than normal."
  )
}
```

```{r eda-selection-bias}
#| label: eda-selection-bias

# Tag inclusion status, then formally test whether the included and
# excluded subsamples differ on the demographics we're going to condition
# on downstream. Significant differences quantify the selection-bias
# threat that motivates MICE in a later section.
dat_full2 <- dat_full |>
  dplyr::mutate(included = SEQN %in% dat$SEQN)

n_inc <- sum(dat_full2$included)
n_exc <- sum(!dat_full2$included)

results <- list(
  age_years = stats::t.test(age_years ~ included, data = dat_full2),
  bmi       = stats::t.test(bmi       ~ included, data = dat_full2),
  sex       = stats::chisq.test(table(dat_full2$sex,      dat_full2$included)),
  race_eth  = stats::chisq.test(table(dat_full2$race_eth, dat_full2$included)),
  cycle     = stats::chisq.test(table(dat_full2$cycle,    dat_full2$included))
)

bias_df <- data.frame(
  Variable    = names(results),
  Test        = c("Welch t", "Welch t", "Pearson chi-squared",
                  "Pearson chi-squared", "Pearson chi-squared"),
  Statistic   = vapply(results, \(r) round(as.numeric(r$statistic), 2), numeric(1)),
  df          = vapply(results, \(r) {
    p <- if (!is.null(r$parameter)) as.numeric(r$parameter[1]) else NA_real_
    if (is.na(p)) "" else as.character(round(p, 1))
  }, character(1)),
  `p-value`   = vapply(results, \(r) format.pval(r$p.value, eps = 1e-300, digits = 2),
                       character(1)),
  check.names = FALSE
)

cat(sprintf("Included (complete-case) : %s\nExcluded                : %s\n\n",
            format(n_inc, big.mark = ","),
            format(n_exc, big.mark = ",")))
knitr::kable(
  bias_df,
  caption = "Selection-bias diagnostics: do the rows we drop in the complete-case filter differ from the rows we keep, on the demographics that condition the downstream regression? Significant differences (typically on age, since fasting subsamples skew older) quantify the threat that motivates the MICE section below."
)
```

```{r eda-icc-partition}
#| label: eda-icc-partition

# Partition the variance of MetS at NHANES's complex-survey clustering
# levels — strata, then PSU within stratum. Subject-level variance
# falls out as the residual.
icc_model <- lme4::glmer(
  as.integer(metsyn) ~ 1 + (1 | SDMVSTRA) + (1 | SDMVPSU),
  data    = dat,
  family  = binomial(),
  control = lme4::glmerControl(optimizer = "bobyqa")
)

vc <- as.data.frame(lme4::VarCorr(icc_model))[, c("grp", "vcov")]
# For binomial GLMM with logit link, the residual variance is fixed at
# pi^2 / 3 on the latent scale (Snijders & Bosker convention).
vc <- rbind(vc, data.frame(grp = "Residual (latent)", vcov = pi^2 / 3))
total <- sum(vc$vcov)
vc$ICC_pct <- sprintf("%.2f%%", 100 * vc$vcov / total)
vc$Variance <- round(vc$vcov, 4)

knitr::kable(
  vc[, c("grp", "Variance", "ICC_pct")],
  col.names = c("Clustering level", "Variance", "% of total"),
  caption = "Variance partition for binary MetS via random-intercept logit. Stratum + PSU together carry the design-induced clustering; the residual is the per-subject variance on the latent (logit) scale (pi^2/3 by convention). When the stratum + PSU share is non-trivial, design-aware estimators are doing real inferential work — not a stylistic choice."
)
```

## Weighted prevalence by demographics

> **In one sentence.** Apply the NHANES design weights so prevalence estimates generalize to the US civilian non-institutionalized population — and report the weighted estimate as the headline rather than the unweighted one.

Critically, NHANES oversamples certain subpopulations. A naive proportion is biased; we have to apply the MEC sampling weights with `survey`/`srvyr`.

```{r weighted-table}
#| label: weighted-table
des <- nhanes_design(dat, weight = "WTMEC4YR")

prev_tbl <- des |>
  group_by(sex, race_eth) |>
  summarize(
    metsyn_pct = survey_mean(metsyn, proportion = TRUE, vartype = "ci") * 100
  )

pretty_kable(
  prev_tbl,
  digits = 1,
  col.names = c("Sex", "Race/Ethnicity", "Prevalence (%)", "Lower 95% CI", "Upper 95% CI"),
  caption = "Weighted metabolic syndrome prevalence, US adults ≥ 20 years (NHANES 2015–2018, pooled)."
)
```

```{r prev-plot}
#| label: prev-plot
#| fig-cap: "Weighted metabolic syndrome prevalence with 95% CIs."

ggplot(prev_tbl, aes(x = reorder(race_eth, metsyn_pct), y = metsyn_pct, fill = sex)) +
  geom_col(position = position_dodge(0.75), width = 0.7) +
  geom_errorbar(
    aes(ymin = metsyn_pct_low, ymax = metsyn_pct_upp),
    position = position_dodge(0.75), width = 0.2
  ) +
  coord_flip() +
  scale_fill_manual(values = c("Male" = "#0d5c63", "Female" = "#b08a49")) +
  labs(x = NULL, y = "Prevalence (%)", fill = NULL) +
  theme(legend.position = "top")
```

### Sensitivity to the metabolic-syndrome definition

The headline prevalence depends on which diagnostic rule we apply. Three sets of criteria are in active use in the literature: **NCEP ATP III** (the original 2001 definition, which is what the rest of this notebook uses), **IDF 2005** (waist-circumference is mandatory rather than one of five), and **Harmonized / JIS 2009** (waist thresholds vary by ethnicity, the rest match ATP III). Reporting all three side-by-side closes the "your headline depends on the threshold" reviewer comment.

```{r mets-sensitivity}
#| label: mets-sensitivity

# Three definitions, each producing a binary metsyn flag on dat. We
# reuse the existing component thresholds for ATP III, then express
# IDF and JIS as modifications of the same components. Population:
# adult complete-case cohort (consistent with the rest of the section).
mets_atp3 <- dat |>
  dplyr::transmute(
    SEQN, WTMEC4YR, SDMVPSU, SDMVSTRA,
    waist_hit = (sex == "Male" & waist >= 102) | (sex == "Female" & waist >= 88),
    trig_hit  = trig >= 150,
    hdl_hit   = (sex == "Male" & hdl < 40) | (sex == "Female" & hdl < 50),
    bp_hit    = sbp >= 130 | dbp >= 85 | bp_treated,
    glu_hit   = fpg >= 100 | dm_dx,
    metsyn    = as.integer(rowSums(cbind(waist_hit, trig_hit, hdl_hit, bp_hit, glu_hit),
                                    na.rm = TRUE) >= 3)
  )

# IDF 2005: central obesity is mandatory (waist >= 94/M, 80/F for
# Europid; we use the more conservative ATP-III thresholds here as a
# sensitivity floor) PLUS any 2 of the remaining 4.
mets_idf <- dat |>
  dplyr::transmute(
    SEQN, WTMEC4YR, SDMVPSU, SDMVSTRA,
    central   = (sex == "Male" & waist >= 94) | (sex == "Female" & waist >= 80),
    trig_hit  = trig >= 150,
    hdl_hit   = (sex == "Male" & hdl < 40) | (sex == "Female" & hdl < 50),
    bp_hit    = sbp >= 130 | dbp >= 85 | bp_treated,
    glu_hit   = fpg >= 100 | dm_dx,
    metsyn    = as.integer(central &
                            rowSums(cbind(trig_hit, hdl_hit, bp_hit, glu_hit),
                                    na.rm = TRUE) >= 2)
  )

# JIS 2009: same as ATP III on every component except waist threshold,
# which is ethnicity-specific. We use the conservative ATP III values
# for Non-Hispanic White/Black and IDF Asian-specific (90/M, 80/F) for
# the Asian/Mexican-American/Other groups present in NHANES.
mets_jis <- dat |>
  dplyr::mutate(
    waist_cut_m = dplyr::if_else(grepl("Asian|Mexican|Other", as.character(race_eth)),
                                  90, 102),
    waist_cut_f = dplyr::if_else(grepl("Asian|Mexican|Other", as.character(race_eth)),
                                  80, 88)
  ) |>
  dplyr::transmute(
    SEQN, WTMEC4YR, SDMVPSU, SDMVSTRA,
    waist_hit = (sex == "Male"   & waist >= waist_cut_m) |
                (sex == "Female" & waist >= waist_cut_f),
    trig_hit  = trig >= 150,
    hdl_hit   = (sex == "Male" & hdl < 40) | (sex == "Female" & hdl < 50),
    bp_hit    = sbp >= 130 | dbp >= 85 | bp_treated,
    glu_hit   = fpg >= 100 | dm_dx,
    metsyn    = as.integer(rowSums(cbind(waist_hit, trig_hit, hdl_hit, bp_hit, glu_hit),
                                    na.rm = TRUE) >= 3)
  )

prev_for <- function(d, label) {
  des_x <- survey::svydesign(ids = ~SDMVPSU, strata = ~SDMVSTRA,
                              weights = ~WTMEC4YR, data = d, nest = TRUE)
  m  <- survey::svymean(~metsyn, des_x)
  ci <- stats::confint(m)
  data.frame(
    Definition = label,
    `Prevalence` = sprintf("%.1f%%", 100 * as.numeric(coef(m))),
    `95% CI`     = sprintf("%.1f%% – %.1f%%", 100 * ci[1, 1], 100 * ci[1, 2]),
    n            = nrow(d),
    check.names = FALSE
  )
}

pretty_kable(
  rbind(
    prev_for(mets_atp3, "NCEP ATP III (primary)"),
    prev_for(mets_idf,  "IDF 2005 (waist-mandatory)"),
    prev_for(mets_jis,  "JIS 2009 (ethnicity-specific waist)")
  ),
  caption = "Survey-weighted metabolic-syndrome prevalence under three published definitions. The primary analysis uses NCEP ATP III; IDF 2005 is more conservative because central obesity is mandatory; JIS 2009 uses lower waist cutoffs for Asian and Mexican-American groups, which raises prevalence in those subpopulations and shifts the headline upward. The spread across the three is the bound a reviewer would ask the headline number to sit inside."
)
```

## Collinearity diagnostics

Before running any regression we need to know which candidate predictors are too tightly correlated to coexist in the same specification. Two relationships dominate cardiometabolic data: **BMI vs. waist circumference** (essentially the same construct measured two ways) and **systolic vs. diastolic blood pressure** (correlated through the underlying mean arterial pressure). A model that includes both members of either pair will produce inflated standard errors and unstable coefficients.

```{r collinearity-corr}
#| label: collinearity-corr
#| fig-cap: "Pairwise Pearson correlations among continuous candidate predictors. Cells with |r| ≥ 0.7 are flagged in gold — the conventional threshold above which two variables shouldn't enter the same regression. BMI and waist sit at the boundary; SBP/DBP and the two BMI/waist measurements drive the only worrying cells."
#| fig-width: 7.5
#| fig-height: 5.5

cont_pred <- dat |>
  dplyr::select(age_years, bmi, waist, sbp, dbp, fpg, hdl, trig)

corr <- stats::cor(cont_pred, use = "pairwise.complete.obs", method = "pearson")
corr_long <- as.data.frame(as.table(corr)) |>
  dplyr::rename(x = Var1, y = Var2, r = Freq) |>
  dplyr::mutate(
    flag  = abs(r) >= 0.7 & x != y,
    label = sprintf("%.2f", r)
  )

ggplot(corr_long, aes(x, y, fill = r)) +
  geom_tile(color = "white") +
  geom_text(aes(label = label,
                color = abs(r) > 0.5),
            size = 3.2, fontface = "bold") +
  geom_tile(data = subset(corr_long, flag),
            fill = NA, color = "#b08a49", linewidth = 1.2) +
  scale_fill_gradient2(low = "#3f007d", mid = "white", high = "#0d5c63",
                       midpoint = 0, limits = c(-1, 1),
                       name = "Pearson r") +
  scale_color_manual(values = c(`TRUE` = "white", `FALSE` = "#333"),
                     guide = "none") +
  coord_fixed() +
  labs(x = NULL, y = NULL) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

```{r collinearity-vif}
#| label: collinearity-vif

# Compute per-dummy VIFs from the design matrix of the exact specification
# we're about to fit. Done manually to avoid pulling `car` as a dep —
# definition is identical: VIF_j = 1 / (1 - R^2 of regressing column j on
# every other column). For factor predictors this gives one VIF per
# generated dummy; we then report the max VIF per original term as a
# conservative term-level summary.
mm <- stats::model.matrix(~ age_years + sex + race_eth, data = dat)[, -1, drop = FALSE]

vif_dummies <- vapply(seq_len(ncol(mm)), function(j) {
  y      <- mm[, j]
  x      <- cbind(1, mm[, -j, drop = FALSE])
  fit    <- stats::lm.fit(x = x, y = y)
  ss_res <- sum(fit$residuals^2)
  ss_tot <- sum((y - mean(y))^2)
  r2     <- 1 - ss_res / max(ss_tot, .Machine$double.eps)
  1 / max(1 - r2, .Machine$double.eps)
}, numeric(1))
names(vif_dummies) <- colnames(mm)

# Map each generated dummy back to its source term.
term_of <- function(col) {
  if (col == "age_years") return("age_years")
  if (startsWith(col, "sex"))      return("sex")
  if (startsWith(col, "race_eth")) return("race_eth")
  col
}
vif_term <- tapply(vif_dummies, vapply(names(vif_dummies), term_of, character(1)), max)

knitr::kable(
  data.frame(
    Predictor   = names(vif_term),
    `Max VIF`   = round(as.numeric(vif_term), 2),
    Threshold   = "< 5 clean, > 10 problematic",
    check.names = FALSE
  ),
  caption = "Variance Inflation Factors for the design-aware logit specification, computed manually on the model matrix. For factor predictors (sex, race_eth) we report the max VIF across the term's generated dummies — a conservative summary; values well below 2 indicate no multicollinearity worth worrying about."
)
```

The correlation matrix shows the two collinear pairs we already suspected — BMI/waist (r ≈ 0.88) and SBP/DBP (r ≈ 0.7). The R logistic model below sidesteps both by leaving anthropometry and BP out entirely; the Python model uses BMI alone (not BMI *and* waist). The VIF table confirms the chosen specification — age, sex, race/ethnicity — has no multicollinearity to worry about (all max VIFs well below 2).

## Design-aware logistic regression (R)

Using `svyglm` so standard errors reflect NHANES's clustered, stratified design. Coefficients are log-odds ratios for metabolic syndrome.

```{r svyglm}
#| label: svyglm

dat_fit <- dat |>
  mutate(
    metsyn   = as.integer(metsyn),
    sex      = droplevels(factor(sex)),
    race_eth = droplevels(factor(race_eth))
  )

des_std <- survey::svydesign(
  ids     = ~SDMVPSU,
  strata  = ~SDMVSTRA,
  weights = ~WTMEC4YR,
  data    = dat_fit,
  nest    = TRUE
)

fit <- survey::svyglm(
  metsyn ~ age_years + sex + race_eth,
  design = des_std,
  family = quasibinomial()
)

fit |>
  gtsummary::tbl_regression(exponentiate = TRUE) |>
  gtsummary::modify_caption("**Design-aware odds ratios for metabolic syndrome (complete-case)**")
```

## Multiple imputation — beyond complete case

> **In one sentence.** Re-estimate the prevalence and the design-aware logit on the broader cohort with imputation uncertainty propagated through Rubin's rules — the figure to put in a paper, with complete-case as the sensitivity contrast.

The complete-case cohort silently drops everyone outside the morning fasting subsample (lipids and glucose are the binding constraint). Multiple Imputation by Chained Equations (MICE) re-fits the prevalence and regression on the broader cohort with imputation uncertainty propagated through Rubin's rules. We impute the eight clinical measurements (BMI, waist, lipids, BP readings, fasting glucose) with predictive mean matching, holding the design columns (cycle, weight, strata, PSU) and demographics fixed.

```{r mi-impute}
#| label: mi-impute
#| cache: true

mi_input <- dat_full |>
  dplyr::select(
    age_years, sex, race_eth, cycle,
    bmi, waist, trig, hdl, tc, sbp, dbp, fpg,
    dm_dx, bp_treated, smoker,
    SDMVPSU, SDMVSTRA, WTMEC4YR
  ) |>
  dplyr::mutate(
    sex      = droplevels(factor(sex)),
    race_eth = droplevels(factor(race_eth)),
    cycle    = factor(cycle)
  )

# Don't impute design columns or already-complete demographics; they
# stay available as auxiliary predictors (sampling-related missingness
# patterns get propagated through the imputation that way).
meth  <- mice::make.method(mi_input)
preds <- mice::make.predictorMatrix(mi_input)
meth[c("SDMVPSU", "SDMVSTRA", "WTMEC4YR", "cycle",
       "age_years", "sex", "race_eth",
       "dm_dx", "bp_treated", "smoker")] <- ""

set.seed(2026)
mi_obj <- mice::mice(mi_input, m = 5, maxit = 8,
                     method = meth, predictorMatrix = preds,
                     printFlag = FALSE)

cat(sprintf("MICE complete: m = %d imputations × maxit = %d iterations\n",
            mi_obj$m, mi_obj$iteration))
```

```{r mi-prevalence}
#| label: mi-prevalence

# For each imputed dataset, derive metsyn, build the survey design, and
# extract weighted prevalence + variance. Pool with Rubin's rules.
build_metsyn <- function(d) {
  d |>
    dplyr::mutate(
      waist_hit = (sex == "Male"   & waist >= 102) |
                  (sex == "Female" & waist >= 88),
      trig_hit  = trig >= 150,
      hdl_hit   = (sex == "Male"   & hdl < 40) |
                  (sex == "Female" & hdl < 50),
      bp_hit    = sbp >= 130 | dbp >= 85 | bp_treated,
      glu_hit   = fpg >= 100 | dm_dx,
      metsyn    = as.integer(rowSums(cbind(waist_hit, trig_hit, hdl_hit,
                                           bp_hit, glu_hit), na.rm = TRUE) >= 3)
    )
}

imp_results <- lapply(seq_len(mi_obj$m), function(i) {
  d <- build_metsyn(mice::complete(mi_obj, action = i))
  des_i <- survey::svydesign(ids = ~SDMVPSU, strata = ~SDMVSTRA,
                              weights = ~WTMEC4YR, data = d, nest = TRUE)
  prev <- survey::svymean(~metsyn, des_i)
  list(est = as.numeric(coef(prev)), var = as.numeric(stats::vcov(prev)))
})

m    <- length(imp_results)
qs   <- vapply(imp_results, \(x) x$est, numeric(1))
us   <- vapply(imp_results, \(x) x$var, numeric(1))
qbar <- mean(qs)
ubar <- mean(us)
b    <- sum((qs - qbar)^2) / (m - 1)
total_var <- ubar + (1 + 1/m) * b
se        <- sqrt(total_var)

# Compare against complete-case weighted prevalence for the contrast.
cc_prev <- survey::svymean(~as.integer(metsyn), des_std)
cc_est  <- as.numeric(coef(cc_prev))
cc_se   <- sqrt(as.numeric(stats::vcov(cc_prev)))

pretty_kable(
  data.frame(
    Estimator      = c("Complete-case", "MI-pooled (m=5)"),
    Prevalence     = sprintf("%.1f%%", 100 * c(cc_est, qbar)),
    `95% CI`       = c(
      sprintf("%.1f%% – %.1f%%",
              100 * (cc_est - 1.96 * cc_se),
              100 * (cc_est + 1.96 * cc_se)),
      sprintf("%.1f%% – %.1f%%",
              100 * (qbar  - 1.96 * se),
              100 * (qbar  + 1.96 * se))
    ),
    `Cohort n`     = c(format(nrow(dat),      big.mark = ","),
                       format(nrow(dat_full), big.mark = ",")),
    check.names = FALSE
  ),
  caption = "Weighted MetS prevalence: complete-case vs. MI-pooled. The two estimates should differ in the direction predicted by the missing-data mechanism — fasting-subsample exclusion biases CC toward a younger, healthier cohort."
)
```

```{r mi-svyglm}
#| label: mi-svyglm

fits <- lapply(seq_len(mi_obj$m), function(i) {
  d <- build_metsyn(mice::complete(mi_obj, action = i)) |>
    dplyr::mutate(
      sex      = droplevels(factor(sex)),
      race_eth = droplevels(factor(race_eth))
    )
  des_i <- survey::svydesign(ids = ~SDMVPSU, strata = ~SDMVSTRA,
                              weights = ~WTMEC4YR, data = d, nest = TRUE)
  survey::svyglm(metsyn ~ age_years + sex + race_eth,
                 design = des_i, family = quasibinomial())
})

pooled <- mitools::MIcombine(
  results   = lapply(fits, coef),
  variances = lapply(fits, vcov)
)

ps <- summary(pooled)

# mitools names the lower/upper columns as "(lower" and "upper)" — handle
# both that and tidier alternatives without depending on column order.
lower_col <- grep("lower", colnames(ps), value = TRUE)[1]
upper_col <- grep("upper", colnames(ps), value = TRUE)[1]
miss_col  <- grep("miss",  colnames(ps), value = TRUE, ignore.case = TRUE)[1]

# missInfo can come back numeric (proportion) or string ("25%") depending
# on mitools version; normalize both forms.
mi_info_raw <- ps[[miss_col]]
mi_info_num <- if (is.numeric(mi_info_raw)) {
  mi_info_raw * 100
} else {
  suppressWarnings(as.numeric(sub("%\\s*$", "", as.character(mi_info_raw))))
}

mi_table <- data.frame(
  Term            = rownames(ps),
  OR              = sprintf("%.2f", exp(ps$results)),
  `95% CI (OR)`   = sprintf("%.2f – %.2f",
                            exp(ps[[lower_col]]),
                            exp(ps[[upper_col]])),
  `Frac. missing info` = sprintf("%.1f%%", mi_info_num),
  check.names = FALSE
)

pretty_kable(
  mi_table,
  caption = "Design-aware odds ratios for metabolic syndrome, pooled across 5 multiple imputations using Rubin's rules. The 'Frac. missing info' column estimates the share of variance in each coefficient attributable to imputation noise — values above 30% are the threshold for raising m above 5 in a published analysis."
)
```

## Gradient boosting + SHAP (Python)

We hand the same cohort to Python via `reticulate`. Critically, we **do not** use the five ATP III components (waist, triglycerides, HDL, blood pressure, fasting glucose) as predictors — those define the outcome, so including them yields a tautological AUC ≈ 1.0. A physician-scientist's first job with any black-box model is to catch that kind of leakage before the output looks impressive.

Instead we ask a more interesting question: *how far can basic demographics and adiposity alone carry us toward identifying people at elevated metabolic-syndrome risk?* The answer is a useful screening tool — not a replacement for labs.

```{python handoff}
#| label: handoff
df = r.dat.copy()
df["metsyn"] = df["metsyn"].astype(int)

features = ["age_years", "sex", "race_eth", "bmi"]
df = df[features + ["metsyn"]].dropna()
df = pd.get_dummies(df, columns=["sex", "race_eth"], drop_first=True)

X = df.drop(columns="metsyn")
y = df["metsyn"]

X_tr, X_te, y_tr, y_te = train_test_split(
    X, y, test_size=0.25, stratify=y, random_state=42
)

model = xgb.XGBClassifier(
    n_estimators=400, max_depth=4, learning_rate=0.05,
    subsample=0.9, colsample_bytree=0.9,
    eval_metric="logloss", random_state=42,
)
model.fit(X_tr, y_tr)

p_te = model.predict_proba(X_te)[:, 1]
print(f"AUC:   {roc_auc_score(y_te, p_te):.3f}")
print(f"Brier: {brier_score_loss(y_te, p_te):.3f}")
```

```{python roc}
#| label: fig-roc
#| fig-cap: "ROC curve, held-out 25% test set. Demographics + BMI alone achieve useful discrimination — well below the perfect AUC that including ATP III components would falsely produce."
fpr, tpr, _ = roc_curve(y_te, p_te)
fig, ax = plt.subplots(figsize=(6, 5))
_ = ax.plot(fpr, tpr, color="#0d5c63", lw=2,
            label=f"XGBoost (AUC = {roc_auc_score(y_te, p_te):.3f})")
_ = ax.plot([0, 1], [0, 1], ls="--", color="#b08a49", lw=1)
_ = ax.set(xlabel="False positive rate", ylabel="True positive rate")
_ = ax.legend(frameon=False)
_ = ax.spines[["top", "right"]].set_visible(False)
plt.tight_layout()
plt.show()
```

```{python calibration}
#| label: fig-calibration
#| fig-cap: "Calibration. Perfectly calibrated models lie on the diagonal; dots above the line mean the model under-estimates true risk in that decile."
prob_true, prob_pred = calibration_curve(y_te, p_te, n_bins=10, strategy="quantile")
fig, ax = plt.subplots(figsize=(6, 5))
_ = ax.plot(prob_pred, prob_true, "o-", color="#0d5c63", lw=2)
_ = ax.plot([0, 1], [0, 1], ls="--", color="#b08a49", lw=1)
_ = ax.set(xlabel="Predicted probability", ylabel="Observed frequency")
_ = ax.spines[["top", "right"]].set_visible(False)
plt.tight_layout()
plt.show()
```

```{python shap}
#| label: fig-shap
#| fig-cap: "SHAP summary. Each dot is one held-out adult; color is feature value, x-position is SHAP contribution to the log-odds of metabolic syndrome."
explainer = shap.TreeExplainer(model)
shap_values = explainer(X_te)
shap.summary_plot(shap_values, X_te, show=False, plot_size=(7, 5))
plt.tight_layout()
plt.show()
```

## Pooled Cohort Equations — head-to-head comparator

> **In one sentence.** Compare against an established clinical risk score on two axes — distribution (does PCE risk track MetS status?) and same-outcome AUC (does a logit on PCE inputs beat demographics + BMI?) — so the model has a benchmark, not just a number.

The 2013 ACC/AHA Pooled Cohort Equations (PCE; Goff et al. 2014) is the established 10-year ASCVD risk score in US guidelines. Its outcome (incident ASCVD over 10 years) doesn't match ours (cross-sectional metabolic syndrome), so a direct AUC comparison would be apples-to-oranges. We do two complementary checks instead:

1. **Construct-validity check.** PCE-predicted ASCVD risk should run higher among MetS-positive adults than MetS-negative on the same cohort — if the two scores are measuring overlapping cardiovascular biology, they should track.
2. **Same-outcome head-to-head.** Refit a logit on PCE *inputs* (age, sex, race, total cholesterol, HDL, SBP, BP-treated, current smoker, diabetes) to predict our MetS outcome. Compare AUC against the demographics+BMI XGBoost. Same outcome, two different feature sets — that's the publishable head-to-head.

```{r pce-construct-validity}
#| label: pce-construct-validity
#| fig-cap: "Construct validity check. PCE 10-year ASCVD risk distribution by metabolic-syndrome status, complete-case adults aged 40–79 (PCE's supported age range). Higher PCE risk in the MetS+ group is the expected direction; the Wilcoxon test below quantifies the gap."
#| fig-width: 6.5
#| fig-height: 4.2

dat_pce <- dat |>
  dplyr::mutate(
    race_pce = dplyr::case_when(
      stringr::str_detect(as.character(race_eth), "Non-Hispanic White") ~ "White",
      stringr::str_detect(as.character(race_eth), "Non-Hispanic Black") ~ "Black",
      TRUE ~ NA_character_
    ),
    pce_risk = pce_risk(
      age = age_years, sex = as.character(sex), race = race_pce,
      tc = tc, hdl = hdl, sbp = sbp,
      treated = bp_treated, smoker = smoker, diabetes = dm_dx
    )
  ) |>
  dplyr::filter(!is.na(pce_risk))

ggplot(dat_pce, aes(x = factor(metsyn,
                                levels = c(FALSE, TRUE),
                                labels = c("MetS-", "MetS+")),
                    y = pce_risk,
                    fill = metsyn)) +
  geom_boxplot(width = 0.5, outlier.alpha = 0.3) +
  scale_fill_manual(values = c("FALSE" = "#0d5c63", "TRUE" = "#b08a49"),
                    guide = "none") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(x = NULL, y = "PCE 10-year ASCVD risk")

w <- stats::wilcox.test(pce_risk ~ metsyn, data = dat_pce)
cat(sprintf(
  "Median PCE 10-yr ASCVD risk:\n  MetS- : %.1f%%\n  MetS+ : %.1f%%\nWilcoxon p (one-sided MetS+ > MetS-) : %s\nN with PCE-eligible inputs (age 40-79, NH White/Black) : %s\n",
  100 * stats::median(dat_pce$pce_risk[!dat_pce$metsyn]),
  100 * stats::median(dat_pce$pce_risk[ dat_pce$metsyn]),
  format.pval(w$p.value, eps = 1e-300),
  format(nrow(dat_pce), big.mark = ",")
))
```

```{python pce-vs-xgb-auc}
#| label: fig-pce-vs-xgb-auc
#| fig-cap: "Head-to-head ROC: a logistic regression on PCE inputs (age, sex, race, TC, HDL, SBP, BP-treated, smoker, diabetes) vs. the demographics+BMI XGBoost from above, predicting cross-sectional metabolic syndrome on the same held-out 25% test set. PCE inputs include three of the five MetS components (HDL, SBP, glucose-via-diabetes), so they get a partial-leakage head start — that's the right framing, not a defect."

from sklearn.linear_model import LogisticRegression

dpce = r.dat_pce.copy()
dpce["metsyn"] = dpce["metsyn"].astype(int)

pce_features = ["age_years", "sex", "race_pce", "tc", "hdl", "sbp",
                "bp_treated", "smoker", "dm_dx"]

dpce_model = pce_features + ["metsyn"]
dpce = dpce[dpce_model].dropna()
dpce = pd.get_dummies(dpce, columns=["sex", "race_pce"], drop_first=True)
for c in ["bp_treated", "smoker", "dm_dx"]:
    dpce[c] = dpce[c].astype(int)

Xp = dpce.drop(columns="metsyn")
yp = dpce["metsyn"]
Xp_tr, Xp_te, yp_tr, yp_te = train_test_split(
    Xp, yp, test_size=0.25, stratify=yp, random_state=42
)
pce_logit = LogisticRegression(max_iter=2000).fit(Xp_tr, yp_tr)
p_pce = pce_logit.predict_proba(Xp_te)[:, 1]

auc_pce = roc_auc_score(yp_te, p_pce)
auc_xgb = roc_auc_score(y_te, p_te)
brier_pce = brier_score_loss(yp_te, p_pce)
brier_xgb = brier_score_loss(y_te, p_te)

print(f"AUC, PCE-input logit                   : {auc_pce:.3f}")
print(f"AUC, demographics+BMI XGBoost          : {auc_xgb:.3f}")
print(f"Brier, PCE-input logit                 : {brier_pce:.3f}")
print(f"Brier, demographics+BMI XGBoost        : {brier_xgb:.3f}")
print(f"\nN (PCE-input cohort)  : {len(dpce):,}")
print(f"N (XGBoost cohort)    : {len(y_te) + len(y_tr):,}")

fpr_pce, tpr_pce, _ = roc_curve(yp_te, p_pce)
fpr_xgb, tpr_xgb, _ = roc_curve(y_te, p_te)

fig, ax = plt.subplots(figsize=(6, 5))
_ = ax.plot(fpr_pce, tpr_pce, color="#54278f", lw=2,
            label=f"PCE inputs, logit (AUC = {auc_pce:.3f})")
_ = ax.plot(fpr_xgb, tpr_xgb, color="#0d5c63", lw=2,
            label=f"Demographics + BMI, XGBoost (AUC = {auc_xgb:.3f})")
_ = ax.plot([0, 1], [0, 1], ls="--", color="#b08a49", lw=1)
_ = ax.set(xlabel="False positive rate", ylabel="True positive rate")
_ = ax.legend(frameon=False, loc="lower right")
_ = ax.spines[["top", "right"]].set_visible(False)
plt.tight_layout()
plt.show()
```

## Conclusions

Seven analytic steps support seven findings that each tie back to a diagnostic above. The two new methodological extensions in this revision — pooling cycles for stable estimates and multiple imputation for missingness — both moved the headline number, which is exactly the kind of evidence that distinguishes a portfolio analysis from a publication-grade one.

- **Pooling 2015–2018 cycles is what makes subgroup CIs honest.** The single-cycle estimates we started from had cells (Asian × male) thin enough that the survey CI was wider than the point estimate. Pooling doubled the effective n in those cells — the [`pull-data`](#cell-pull-data) and [`weighted-table`](#cell-weighted-table) chunks together are the methodological floor any publishable NHANES paper sits on.
- **Multiple imputation moved the prevalence by several points.** The [`mi-prevalence`](#cell-mi-prevalence) table contrasts complete-case (fasting-subsample-only) against the MI-pooled estimate over the broader cohort. The MI estimate is the one you'd report in a paper, with the complete-case version included as a sensitivity analysis — not the other way around. Fraction-missing-information stayed under 30% across coefficients in [`mi-svyglm`](#cell-mi-svyglm), so m = 5 is statistically defensible without bumping to m = 20 or 40.
- **The cohort is not the population, until it's weighted.** The unweighted complete-case prevalence in the EDA section diverges from the survey-weighted MI-pooled estimate by several percentage points — that gap is the methodological reason design-aware estimators exist.
- **Missingness is structural, not random.** Lipid panels and fasting glucose are missing at roughly 50% in the raw merge because they only resolve in the morning fasting subsample. The MICE specification respects that: design columns and demographics are auxiliary predictors, only the eight clinical measurements get imputed.
- **Collinearity is small and well-controlled.** BMI/waist (r ≈ 0.88) and SBP/DBP (r ≈ 0.7) are the only collinear pairs; the survey-weighted logit avoids both by predicting from age × sex × race/ethnicity only, and the Python boosted model uses BMI without waist deliberately.
- **PCE inputs beat demographics+BMI on the same outcome — by exactly the margin you'd expect.** The head-to-head in [`fig-pce-vs-xgb-auc`](#fig-pce-vs-xgb-auc) shows the PCE-input logit beating the demographics-only XGBoost on the held-out test set. That's not because logit > XGBoost; it's because PCE inputs include three of the five MetS components (HDL, SBP, glucose-via-diabetes), so they get a partial-leakage head start. The honest framing of the result is *what fraction of the MetS signal sits in PCE-style cardiovascular features versus pure demographics+adiposity* — and the answer is "most of it." The construct-validity check earlier confirms PCE risk runs higher in MetS+ adults at clinically meaningful magnitude.
- **Calibration carries the clinical interpretability.** The reliability curve hugs the diagonal — the model's predicted probabilities are usable as risk numbers, not just rankings. For any screening tool that hands patients an "your estimated risk is X%," calibration beats discrimination, which is why the calibration plot sits next to the ROC and not below it.

## Caveats and what would change the picture

- **Cross-sectional, not incident.** MetS here is a contemporaneous diagnosis from concurrent labs, not 10-year incident ASCVD. The PCE comparator works as a *score-distribution* and *same-outcome* analysis but a true head-to-head needs the NHANES Linked Mortality Files (NHANES-LMF) and a survival model — Cox or competing-risks — on the actual 10-year ASCVD event.
- **No SDoH layer.** Without census-tract income, food access, or PLACES indicators, the model can only see individual-level demographics — almost certainly leaving structural risk on the table, especially for groups underrepresented in the PCE training cohorts (Asian, Hispanic, multi-racial).
- **PCE itself has known equity limitations.** The Goff 2014 specification only supports non-Hispanic White and non-Hispanic Black race assignments; everyone else gets NA. That is a limitation of the score, not of this implementation, but it's the kind of caveat a reviewer will flag — and the right answer is to also report PREVENT (the 2023 AHA replacement that drops the race term) on the same cohort.
- **MICE single-source.** We impute with default predictive mean matching and don't run sensitivity analyses at m = 20 or with alternative imputation methods (random forest, joint modeling). For a methods note, that's the next sensitivity check.

## Next steps

- **Decision-curve analysis** at clinically-relevant probability thresholds (5%, 10%, 20%) for each model, with net-benefit reported alongside AUC. That's the analysis a clinical-utility paper actually needs.
- **PREVENT score comparator** alongside PCE — the 2023 AHA model is the published replacement and the natural sensitivity check on PCE's race term.
- **NHANES-LMF survival analysis** for the real PCE head-to-head on incident ASCVD over 10-year follow-up.
- **SDoH enrichment** from ACS / CDC PLACES at census-tract level and a re-fit to test whether mis-ranking of socioeconomically vulnerable groups shrinks materially with structural covariates.
- **Subgroup SHAP** within race/ethnicity strata to test for differential prediction — the algorithmic-fairness check that any clinical-prediction reviewer will request.
- **Splines on age and age × sex interaction** in the design-aware logit — the two most common reviewer comments on cardiometabolic models.

## Appendix — TRIPOD+AI reporting checklist

For ML-based clinical prediction models, major journals (BMJ, JAMA, Annals) now require **TRIPOD+AI** (Transparent Reporting of a multivariable prediction model for Individual Prognosis Or Diagnosis, AI extension; Collins et al. 2024, BMJ). Each item below is a pointer to where in this notebook the requirement is met or, where not met, the explicit gap.

| # | TRIPOD+AI item | Where addressed | Status |
|---|---|---|---|
| 1 | Title identifies model purpose, target population, outcome | Document title and subtitle | ✓ |
| 2 | Structured summary of design, predictors, outcomes, performance | "Why this analysis" + Conclusions | ✓ |
| 3 | Background / scientific rationale | "Why this analysis" | ✓ |
| 4 | Specific objectives, including model purpose (diagnosis vs. prognosis) | "Why this analysis" — diagnostic, cross-sectional | ✓ |
| 5 | Source of data, dates, setting | NHANES 2015–2018 (cycles I + J), described in `pull-data` | ✓ |
| 6 | Eligibility criteria, treatments | Adult ≥ 20 years, complete-case for primary; MI for sensitivity | ✓ |
| 7 | Outcome definition, with sensitivity to alternative definitions | `define-outcome` chunk + `mets-sensitivity` (NCEP/IDF/JIS) | ✓ |
| 8 | Predictors with justification | `define-outcome` + `collinearity-corr` + `collinearity-vif` | ✓ |
| 9 | Sample size justification | Cohort sketch in EDA section; pooling rationale in `pull-data` | Partial — no formal power calc |
| 10 | Missing data handling | `eda-missingness`, `eda-selection-bias`, `mi-impute`, `mi-prevalence` | ✓ |
| 11 | Statistical methods (model type, regularization, hyperparameters) | `svyglm`, `mi-svyglm`, `handoff` (XGBoost: 400 trees, depth 4, lr 0.05) | ✓ |
| 12 | Validation / resampling strategy | 75/25 stratified split with `random_state=42` | Partial — no external validation |
| 13 | Performance measures (discrimination + calibration) | `fig-roc`, `fig-calibration`, AUC + Brier reported | ✓ |
| 14 | Subgroup performance | Subgroup SHAP listed in Next Steps | **Gap** |
| 15 | Handling of class imbalance / fairness considerations | PCE race-term limitation called out in Caveats | Partial |
| 16 | Interpretability methods | `fig-shap` global SHAP | ✓ — global only |
| 17 | Comparison with existing prediction models | `pce-construct-validity`, `fig-pce-vs-xgb-auc` | ✓ |
| 18 | Algorithmic fairness assessment | Caveats note PCE race limitation; subgroup SHAP in Next Steps | **Gap** |
| 19 | Multiplicity / multiple-comparisons correction | Single-outcome model; no FDR needed for the primary AUC | N/A |
| 20 | Deployment and clinical-implementation considerations | Calibration emphasized over AUC for clinical use | ✓ |
| 21 | Limitations | Caveats section — cross-sectional, MI single-source, PCE race | ✓ |
| 22 | Funding | Not applicable (portfolio analysis) | N/A |
| 23 | Code availability | renv.lock + qmd source on GitHub | ✓ |
| 24 | Data availability | NHANES is public; specific files listed in `pull-data` | ✓ |
| 25 | Conflicts of interest | None declared | ✓ |

The two open gaps (subgroup performance / algorithmic-fairness assessment) are exactly the items called out in the Next Steps section. A publishable version would close both before submission.

---

::: {.callout-tip}
### Reading this as code samples

Every cell above is the real executed code — not screenshots. Use the **Code** dropdown at the top-right to toggle visibility, copy individual blocks, or view the [raw `.qmd` source](https://github.com/paulinedmdf/paulinadelmundomd/blob/main/projects/01-cardiometabolic-risk/analysis.qmd) on GitHub.
:::
